---
title: "pH Curves"
author: "Alberto Barron"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_width: 11
    fig_height: 7
editor_options: 
  chunk_output_type: console
---

#### Description

This script will consolidate the processed data from the initial GC analysis and merge it into a larger file. Then it finds triplicate combinations of strain and pH. Finally it calculates mean growth rate and sd for each combination and plots the values against the pH levels.

```{r setup, include=FALSE}
library(here)
knitr::opts_chunk$set(echo = T, cache = F)

# Output directories (gitignored; created on a fresh clone)
dir.create(here("output/pH/plots/pH_curves"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/rds"), recursive = TRUE, showWarnings = FALSE)

knitr::opts_knit$set(root.dir = here("output/pH"))
options(scipen=10000, digits = 12)
set.seed(seed = 711)
```

```{r general settings, message=FALSE, warning=FALSE, results='hold', echo=FALSE}

library(growthcurver)
library(tidyverse)
library(reshape2)
library(purrr)
library(tuple)
library(tidyr)
library(broom)

apatheme=theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family='Arial'))



```

#### Helper functions

General-purpose functions used throughout this analysis are collected here so they are defined before any downstream chunk calls them.

**`%||%`** — Null-coalescing operator: returns `a` unless it is `NULL`, in which case it returns `b`.

```{r helper-null-coalesce, message=FALSE, warning=FALSE}
# helper for a safe default (returns b if a is NULL)
`%||%` <- function(a, b) if (is.null(a)) b else a
```

**`calculate_auc`** — Fit loess curves of growth rate (r) and carrying capacity (K) against pH for a strain and return the trapezoidal area under each curve (AUC).

```{r helper-calculate-auc, message=FALSE, warning=FALSE}
# Function to calculate area under the curve for each strain
calculate_auc <- function(data) {
  if (nrow(data) == 0) {
    return(data.frame(strain = NA, area_under_curve = NA))
  }
  
  loess_model_r <- loess(r_mean ~ pH, data = data)
  loess_model_k <- loess(k_mean ~ pH, data = data)
  predicted_values_r <- predict(loess_model_r, data$pH)
  area_under_curve_r <- trapz(data$pH, predicted_values_r)
  predicted_values_k <- predict(loess_model_k, data$pH)
  area_under_curve_k <- trapz(data$pH, predicted_values_k)
  return(data.frame(strain = unique(data$strain),
                    area_under_curve_r = area_under_curve_r,
                    area_under_curve_k = area_under_curve_k))
}
```

**`get_max_rk_mean`** — Fit loess of r and K against pH for a strain and return the maximum predicted r and K with the pH optimum at which each occurs.

```{r helper-get-max-rk, message=FALSE, warning=FALSE}
# Function to find each strain's maximum predicted r and K and the pH at which each occurs
get_max_rk_mean <- function(data) {
  # Fit the loess model
  loess_model_r <- loess(r_mean ~ pH, data = data)
  loess_model_k <- loess(k_mean ~ pH, data = data)
  
  # Predict values for a range of x-values
  predicted_values_r <- predict(loess_model_r, data$pH)
  predicted_values_k <- predict(loess_model_k, data$pH)
  
  # Find the index of the maximum predicted r and K values
  max_index_r <- which.max(predicted_values_r)
  max_index_k <- which.max(predicted_values_k)
  
  # Get the pH at the maximum r and K
  max_pH_r <- data$pH[max_index_r]
  max_pH_k <- data$pH[max_index_k]
  
  # Get the maximum r and K values
  max_r_mean <- predicted_values_r[max_index_r]
  max_k_mean <- predicted_values_k[max_index_k]
  
  # Return a data frame with the results
  data.frame(strain = unique(data$strain),
             opt_pH_r = max_pH_r,
             max_r_mean = max_r_mean,
             opt_pH_k = max_pH_k,
             max_k_mean = max_k_mean)
}
```

**`dotplot_traits_by_subclade`** — Dot plot of a single curve-parameter (trait) across subclades, with per-strain points, group means, and bootstrap CIs.

```{r helper-dotplot-traits, message=FALSE, warning=FALSE}
dotplot_traits_by_subclade <- function(
  params_df,     # your d_params_ssfull
  traits,        # character vector of trait names, e.g. c("topt","rmax","ctmax")
  subclades = NULL,      # which subclades to plot; NULL = all
  facet_by_clade = T,    # if FALSE, facet only by term
  filename  = NULL,      # if non-NULL, path to save the PNG
  width_mm  = 180,
  height_mm = 100,
  dpi       = 300
){
  # 1. join metadata and pivot longer
  df_long <- params_df %>%
    filter(clade != "UND", !is.na(subclade)) %>%
    { if(!is.null(subclades)) filter(., subclade %in% subclades) else . } %>%
    filter(Parameter %in% traits) |>
    mutate(forcats::fct_rev(forcats::fct_inorder(subclade)))
  
  # 2. build the plot
  plt <- ggplot(df_long,
                aes(x = Value,
                    y = subclade,
                    colour = subclade)) +
    geom_jitter(height = 0.15, alpha = 0.35) +
    stat_summary(fun = mean, geom = "point",
                 size = 3, shape = 21, fill = "white") +
    stat_summary(fun.data = mean_cl_boot,
                 geom     = "errorbarh",
                 height   = 0.2) +
    scale_y_discrete(limits = rev) +
    scale_colour_manual(values = subclade_col, guide = "none") +
    labs(x = "Estimate", y = "Subclade") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y      = element_text(face = "bold"),
      strip.text       = element_text(size = 16),
      panel.grid.major = element_line(colour = "grey90"),
      panel.grid.minor = element_blank(),
      legend.position  = "none"
    )
  
  # 3. Add faceting
  if (facet_by_clade) {
      plt <- plt + facet_grid(clade ~ Parameter, scales = "free", space = "free_y") 
  } else {
    plt <- plt + facet_grid(~ Parameter, scales = "free", space = "free_y") 
  }
  
  
  
  # 4. print, and optionally save
  print(plt)
  if(!is.null(filename)){
    ggsave(filename   = filename,
           plot       = plt,
           width      = width_mm,
           height     = height_mm,
           units      = "mm",
           dpi        = dpi)
  }
  
  return(plt)
}
```

**`plot_traits_subclade`** — Plot curve-parameter distributions by subclade, running per-trait normality/variance checks plus omnibus and post-hoc tests to annotate significance.

```{r helper-plot-traits-subclade, message=FALSE, warning=FALSE}
plot_traits_subclade <- function(
  params_df,
  clade_info = NULL,          # <-- make optional
  traits,
  subclades = NULL,
  exclude_subs = NULL,
  facet_by_clade = TRUE,
  clade_levels = NULL,
  subclade_levels = NULL, # order of subclades
  flip = FALSE,
  term_labels = NULL,
  show_annotations = c("both","omnibus","letters","none"),
  omnibus_in_clade = NULL,
  subclade_col = NULL,
  filename = NULL, width_mm = 180, height_mm = 100, dpi = 300
){
  show_annotations <- match.arg(show_annotations)

  # --- 1) Prepare long data, with optional join ---
  df0 <- params_df
  if (!is.null(clade_info)) {
    df0 <- dplyr::left_join(df0, clade_info, by = c("Strain" = "strain"))
  }
  # Ensure we actually have clade/subclade
  stopifnot(all(c("clade","subclade") %in% names(df0)))

  df_long <- df0 %>%
    dplyr::filter(!is.na(subclade), clade != "UND") %>%
    { if (!is.null(subclades)) dplyr::filter(., subclade %in% subclades) else . } %>%
    { if (!is.null(exclude_subs)) dplyr::filter(., !subclade %in% exclude_subs) else . } %>%
    tidyr::pivot_longer(dplyr::all_of(traits), names_to = "term", values_to = "value") %>%
    dplyr::mutate(
      subclade = if (!is.null(subclade_levels)) forcats::fct_relevel(subclade, subclade_levels)
                 else if (!is.null(subclades)) forcats::fct_relevel(subclade, rev(subclades))
                 else forcats::fct_rev(forcats::fct_inorder(subclade)),
      clade = factor(clade)
    )
  
  df_long <- df_long %>% tidyr::drop_na(subclade)  # just to be safe
  y_levels <- levels(droplevels(df_long$subclade))
  
  if (!is.null(traits)) {
    term_levels <- c("opt_pH_r","max_pH_r","min_pH_r", "breadth_r", "max_r_mean",
                     "opt_pH_k","max_pH_k","min_pH_k", "breadth_k", "max_k_mean")

  df_long <- df_long |>
    mutate(term = factor(term, levels = term_levels))
  }
  
  facet_labeller <- if (!is.null(term_labels)) {
        labeller(term = as_labeller(term_labels))
      } else {
        label_value  # default
      }

  # ---------- 2) Base plot ----------
  if (!flip) {
    plt <- ggplot2::ggplot(df_long, ggplot2::aes(x = value, y = subclade, colour = subclade)) +
      ggplot2::geom_jitter(height = 0.15, alpha = 0.35) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 3, shape = 21, fill = "white") +
      ggplot2::stat_summary(fun.data = mean_cl_boot, geom = "errorbarh", height = 0.2) +
      ggplot2::labs(x = "pH", y = "Subclade")
    plt <- plt + if (facet_by_clade)
      ggplot2::facet_grid(clade ~ term, scales = "free", space = "free_y",
                          labeller = facet_labeller)
    else
      ggplot2::facet_grid(~ term, scales = "free", space = "free_y",
                          labeller = facet_labeller)
  } else {
    plt <- ggplot2::ggplot(df_long, ggplot2::aes(y = value, x = subclade, colour = subclade)) +
      ggplot2::geom_jitter(width = 0.15, alpha = 0.35) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 3, shape = 21, fill = "white") +
      ggplot2::stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
      ggplot2::labs(y = "pH", x = "Subclade") +
      scale_y_continuous(position = "right")
    plt <- plt + if (facet_by_clade)
      ggplot2::facet_grid(term ~ clade, scales = "free", space = "free_x", switch = "y",
                          labeller = facet_labeller)
    else
      ggplot2::facet_grid(term ~ ., scales = "free", space = "free_x", switch = "y",
                          labeller = facet_labeller)
  }

  if (!is.null(subclade_col)) {
    plt <- plt + ggplot2::scale_colour_manual(values = subclade_col, guide = "none")
  } else {
    plt <- plt + ggplot2::guides(colour = "none")
  }

  plt <- plt +
    ggplot2::theme_minimal(base_size = 8) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(face = "bold"),
      strip.text  = ggplot2::element_text(size = 8),
      panel.grid.minor = ggplot2::element_blank()
    )

  # ---------- 3) Stats (omnibus + posthocs) ----------
  # Helper: run per (clade, term) if faceting by clade; else per term (across clades)
  #stats_grouping <- if (facet_by_clade) c("clade","term") else c("term")
  stats_grouping <- "term"
  stat_input <- df_long %>% dplyr::select(clade, subclade, term, value)

  posthocs <- stat_input %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(stats_grouping))) %>%
    dplyr::group_modify(function(dat, key){
      sub_counts <- dplyr::count(dat, subclade)
      if (dplyr::n_distinct(dat$subclade) < 2 || any(sub_counts$n < 2)) {
        return(tibble::tibble())
      }

      # safer Shapiro/Levene calls
      sw_p  <- tryCatch(rstatix::shapiro_test(dat$value) %>% dplyr::pull(p), error = function(e) NA_real_)
      lev_p <- tryCatch(rstatix::levene_test(value ~ subclade, data = dat) %>% dplyr::pull(p), error = function(e) NA_real_)

      if (!is.na(sw_p) && sw_p > 0.05 && !is.na(lev_p) && lev_p > 0.05) {
        omni <- rstatix::anova_test(value ~ subclade, data = dat)
        ph   <- rstatix::tukey_hsd(value ~ subclade, data = dat) %>%
          dplyr::mutate(method = "Tukey HSD")
        test_type <- "ANOVA"
      } else {
        omni <- rstatix::kruskal_test(value ~ subclade, data = dat)
        ph   <- rstatix::dunn_test(value ~ subclade, data = dat, p.adjust.method = "BH") %>%
          dplyr::mutate(method = "Dunn (BH)")
        test_type <- "KW"
      }

      vmax <- max(dat$value, na.rm = TRUE)
      omni_row <- tibble::tibble(
        group1 = NA_character_, group2 = NA_character_,
        p.adj  = as.numeric(omni$p)[1],
        p.adj.signif = as.character(stats::symnum(as.numeric(omni$p)[1],
                                                  corr = FALSE,
                                                  cutpoints = c(0,.001,.01,.05,1),
                                                  symbols   = c("***","**","*","n.s."))),
        xpos = vmax * 1.02, step = "omnibus", method = test_type
      )

      ph_rows <- ph %>%
        dplyr::transmute(
          group1, group2, p.adj,
          p.adj.signif = as.character(stats::symnum(p.adj, corr = FALSE,
                                                    cutpoints = c(0,.001,.01,.05,1),
                                                    symbols   = c("***","**","*","n.s."))),
          xpos = vmax * 0.08, step = "posthoc", method
        )

      dplyr::bind_rows(omni_row, ph_rows)
    }) %>%
    dplyr::ungroup()


  
  # --- robust letters per term (no clade grouping) ---
  # subclade -> clade map (use df_long so it matches the plot)
sub2clade <- df_long %>% dplyr::distinct(subclade, clade)

letter_df <- posthocs %>%
  dplyr::filter(step == "posthoc") %>%
  dplyr::select(term, group1, group2, p.adj) %>%
  dplyr::group_by(term) %>%
  tidyr::nest() %>%
  dplyr::mutate(tbl = purrr::map2(data, term, function(ph, term_val) {

    # subclades that actually appear in the plot for this term
    subs_this_term <- df_long %>%
      dplyr::filter(term == term_val) %>%
      dplyr::distinct(subclade) %>%
      dplyr::pull(subclade) %>%
      as.character()

    # no pairwise rows? -> give every subclade "a"
    if (nrow(ph) == 0L) {
      return(tibble::tibble(
        subclade = factor(subs_this_term, levels = y_levels),
        letter   = "a"
      ))
    }

    # build named vector of p-values: "A-B" = p
    pv <- stats::setNames(as.numeric(ph$p.adj),
                          paste(ph$group1, ph$group2, sep = "-"))
    pv <- pv[!duplicated(names(pv))]
    pv <- pv[!is.na(pv)]

    if (length(pv) == 0L) {
      return(tibble::tibble(
        subclade = factor(subs_this_term, levels = y_levels),
        letter   = "a"
      ))
    }

    # multcomp letters with fail-safe
    lets <- tryCatch(
      multcompView::multcompLetters(pv, threshold = 0.05)$Letters,
      error = function(e) stats::setNames(rep("a", length(subs_this_term)), subs_this_term)
    )

    tibble::tibble(
      subclade = factor(names(lets), levels = y_levels),
      letter   = unname(lets)
    )
  })) %>%
  dplyr::select(term, tbl) %>%
  tidyr::unnest(tbl) %>%
  dplyr::left_join(sub2clade, by = "subclade")


  # One omnibus per term (place in chosen clade if facetted)
  if (facet_by_clade) {
    posthocs_omni_one <- posthocs %>%
      dplyr::filter(step == "omnibus") %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(p.adj.signif = dplyr::first(p.adj.signif),
                       xpos         = dplyr::first(xpos),
                       .groups = "drop") %>%
      dplyr::mutate(clade = factor(omnibus_in_clade, levels = clade_levels))
  } else {
    posthocs_omni_one <- posthocs %>%
      dplyr::filter(step == "omnibus") %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(p.adj.signif = dplyr::first(p.adj.signif),
                       xpos         = dplyr::first(xpos),
                       .groups = "drop")
  }

  
  # compute per-term y positions for annotations (works with facet rows = term)
  rng <- df_long %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      span = max - min,
      # put letters a bit above the top of the data for that term
      lab = max + 0.06 * span,
      # position for the omnibus star
      star = max + 0.12 * span,
      .groups = "drop"
    )
  
  # join positions to your annotation data
  letters_plt <- letter_df %>% dplyr::left_join(rng, by = "term")
  omni_plt    <- posthocs_omni_one %>%
    #dplyr::distinct(term, p.adj.signif) %>%
    dplyr::left_join(rng, by = "term")

  # suppress comparison letters for terms where the omnibus test is not significant
  sig_terms   <- omni_plt %>% dplyr::filter(p.adj.signif != "n.s.") %>% dplyr::pull(term)
  letters_plt <- letters_plt %>% dplyr::filter(term %in% sig_terms)

  # ---------- 4) Add annotations ----------
  if (show_annotations %in% c("both","omnibus")) {
    if (!flip) {
      plt <- plt + ggplot2::geom_text(
        data = omni_plt,
        ggplot2::aes(x = star*0.85, y = -Inf, label = p.adj.signif),
        inherit.aes = FALSE, vjust = 1.2, size = 3
      )
    } else {
      plt <- plt + ggplot2::geom_text(
        data = omni_plt,
        ggplot2::aes(y = star*0.85, x = -Inf, label = p.adj.signif),
        inherit.aes = FALSE, hjust = -0.1, size = 3
      )
    }
  }

  if (show_annotations %in% c("both","letters")) {
    if (!flip) {
      plt <- plt + ggplot2::geom_text(
        data = letters_plt,
        ggplot2::aes(x = lab*0.85, y = subclade, label = letter),
        inherit.aes = FALSE, hjust = 0, size = 2.5
      )
    } else {
      plt <- plt + ggplot2::geom_text(
        data = letters_plt,
        ggplot2::aes(y = lab*0.85, x = subclade, label = letter),
        inherit.aes = FALSE, vjust = 0, hjust = 2.5, size = 2.5
      )
    }
  }

  # ---------- 5) Save & return ----------
  print(plt)
  if (!is.null(filename)) {
    ggplot2::ggsave(filename, plot = plt,
                    width = width_mm, height = height_mm,
                    units = "mm", dpi = dpi)
  }

  invisible(list(
    plot = plt,
    data_long = df_long,
    posthocs = posthocs,
    letters  = letter_df,
    omnibus_one = posthocs_omni_one
  ))
}
```


```{r set working directory and load and merge, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
# Working directory is handled by here(); GC_values inputs live in data/pH
files_df <- tibble(
  path = list.files(path = here("data/pH"), pattern = "^GC_values_pHassay\\..*\\.txt$", full.names = TRUE)
) %>%
  mutate(
    file = basename(path),
    # Extract YYYY-MM-DD that appears at the end segment before .txt
    date_str  = str_extract(file, "(?:19|20)\\d{2}-\\d{2}-\\d{2}"),
    file_date = ymd(date_str),
    year      = year(file_date),
    is_2025   = (year == 2025),
    # Extract rep number if present (rep1/rep2/rep3)
    rep_str   = str_extract(file, "rep\\d+"),
    plate     = if_else(is.na(rep_str), NA_character_,
                        paste0("P", str_extract(rep_str, "\\d+")))
  )


# Read all files; keep filename-derived columns alongside the data
all.data <- files_df %>%
  mutate(dat = map(path, ~ readr::read_tsv(.x, show_col_types = FALSE))) %>%
  select(file, file_date, year, is_2025, plate, dat) %>%
  unnest(dat) |>
  mutate(date = case_when(file_date == "2024-05-13" ~ file_date,
                          file_date == "2024-04-30" ~ file_date,
                          TRUE ~ date),
          date = ymd(date)) |>
  mutate(strain = case_when(strain == "Desert-08" ~ "Desert-8",
                            TRUE ~ strain))



# Quick sanity checks:
# 1) Which files were tagged 2025 and their plate labels?
all.data %>% distinct(file, is_2025, plate) %>% arrange(file) %>% print(n=50)

# 2) If a 'date' column exists in your data files, verify it matches file_date
if ("date" %in% names(all.data)) {
  all.data <- all.data %>% mutate(date = ymd(date))
  mism <- all.data %>% filter(!is.na(date), !is.na(file_date), date != file_date)
  if (nrow(mism) > 0) {
    message("Warning: some rows have date mismatch between file content and filename.")
    print(mism %>% select(file, date, file_date) %>% distinct())
  }
}



```


### Number of assays and ANOVAS for each pH level and strain across date of the assay

```{r number of assays summary and ANOVAS, message=FALSE, warning=FALSE, echo=FALSE, fig.height=18, fig.width=12}

## Let's get the metadata info for each strain. For that we need to read in the master metadata table.
master_md <- read.table(here("data/shared/master_md_all_sequenced_Isolates_with_extendedMD.txt"), header = T, sep = "\t")

master_md <- master_md %>% select(genomeID, Name, clade, subclade)


clade_info <- merge(all.data, master_md, by.x = "strain", by.y = "Name", all.x = T) %>% select(strain, clade, subclade) %>% distinct()

all.data <-merge(all.data, clade_info, by = "strain", all.x = T)


all.data$strain.date<-c(paste(all.data$strain, all.data$date, sep = "."))
all.data$strain.rep<-c(paste(all.data$strain, all.data$replicate, sep = "."))
all.data$strain.date.rep<-c(paste(all.data$strain, all.data$date, all.data$replicate, sep = "."))



all.data$date <- as.Date(all.data$date)

#now let's match the clade colors to the metadata
clade_colors <- read.table(here("data/shared/clade_colors.txt"), header = T, sep = "\t", comment.char = "")
all.data$clade_color <- clade_colors$color[match(all.data$clade, clade_colors$variable)]
all.data$subclade_color <- clade_colors$color[match(all.data$subclade, clade_colors$variable)]

subclade_colors_vec <- clade_colors |>
  tibble::deframe()


all.data_25 <- all.data %>%
  filter(!is.na(replicate)) %>%
  filter(year(date) == 2025)



# Assumes all.data has: strain, date, plate, (optionally replicate/well-level)
# 1) Count replication structure per strain
rep_summary <- all.data %>%
  mutate(date = as.Date(date)) %>%
  group_by(strain) %>%
  summarise(
    n_dates  = n_distinct(date, na.rm = TRUE),
    # count unique plate runs; if older data lack `plate`, treat each date as one plate
    n_plates = n_distinct(ifelse(is.na(plate),
                                 paste0("date:", date),        # fallback: 1 plate per date
                                 paste0(date, "::", plate) ),  # plate identities within date
                           na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    replication_type = case_when(
      n_dates >= 2 ~ "multi-date",
      n_dates == 1 & n_plates >= 2 ~ "single-date_multi-plate",
      TRUE ~ "single-date_single-plate"
    ),
    keep = (n_dates >= 2) | (n_dates == 1 & n_plates >= 2)
  )

# 2) Join clade info and keep strains per rule
rep_keep <- rep_summary %>%
  left_join(clade_info, by = "strain") %>%
  filter(keep)

rep_keep |>
  group_by(clade) |>
  count() |>
  print(n = 21)

rep_keep |>
  group_by(clade, subclade) |>
  count() |>
  print(n = 21)





# rep_keep now has the strains you’ll include
# If you still want a dataframe at the row level filtered to these strains:
all_kept <- all.data %>%
  semi_join(rep_keep %>% select(strain), by = "strain")


all.data <- all.data %>% filter(strain.date != "35G2A07.2023-07-19") |> filter(r < 2)



all_kept <- all.data %>%
  semi_join(rep_keep %>% select(strain), by = "strain")


data_cleaned <- all.data %>%
  filter(!is.na(replicate)) %>%            # your original filter
  select(strain, pH, date, plate, r, k) %>%
  drop_na(strain, pH, r) %>%
  mutate(
    date  = as.Date(date),                 # keep as Date
    plate = as.factor(plate)
  )



# run the appropriate ANOVA per (strain, pH)
anova_results <- data_cleaned %>%
  group_by(strain, pH) %>%
  group_modify(function(.x, .key) {
    nd <- n_distinct(.x$date)
    np <- n_distinct(.x$plate)

    if (nd >= 2) {
      fit <- aov(r ~ date, data = .x)
      broom::tidy(fit) %>%
        filter(term == "date") %>%
        mutate(test_factor = "date",
               n_dates = nd, n_plates = np)
    } else if (nd == 1 && np >= 2) {
      fit <- aov(r ~ plate, data = .x)
      broom::tidy(fit) %>%
        filter(term == "plate") %>%
        mutate(test_factor = "plate",
               n_dates = nd, n_plates = np)
    } else {
      tibble(
        term = "none", df = NA_real_, sumsq = NA_real_, meansq = NA_real_,
        statistic = NA_real_, p.value = NA_real_,
        test_factor = "insufficient",
        n_dates = nd, n_plates = np
      )
    }
  }) %>%
  ungroup()

# BH correction within strain (optional but recommended)
anova_results <- anova_results %>%
  group_by(strain) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  ungroup()


# join clade metadata
final.table <- anova_results %>%
  left_join(clade_info, by = "strain")

# ---- robust color vectors (avoid duplicates from row-level all.data) ----
## create strings to color by:
## clade:
clade_col <- as.character(all.data$clade_color)
names(clade_col) <- as.character(all.data$clade)
# strain by clade color
strain_clade_col <- as.character(all.data$clade_color)
names(strain_clade_col) <- as.character(all.data$strain)

## subclade:
subclade_col <- as.character(all.data$subclade_color)
names(subclade_col) <- as.character(all.data$subclade)
# strain by clade color
strain_subclade_col <- as.character(all.data$subclade_color)
names(strain_subclade_col) <- as.character(all.data$strain)

clade_color_vector <- setNames(clade_colors$color, clade_colors$variable)


# ---- plot: show which factor was tested ----

ggplot(
  final.table %>% filter(test_factor != "insufficient"),
  aes(x = pH, y = p.value, color = clade, shape = test_factor)
) +
  geom_point(size = 2.8, alpha = 0.9) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  scale_color_manual(values = clade_col) +
  scale_shape_manual(values = c(date = 16, plate = 17)) +  # circle vs triangle
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.3)) +
  labs(
    title = "ANOVA p-values testing within-group variation of r",
    subtitle = "If ≥2 dates → test by date; else if 1 date & ≥2 plates → test by plate",
    y = "p value"
  ) +
  facet_wrap(strain ~ .) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))




data_cleaned <- all.data %>% filter(!is.na(replicate)) %>% 
  select(strain, pH, date, r, k) %>%
  drop_na() %>%
  mutate(date = as.factor(date))

data_filtered <- data_cleaned %>%
  group_by(strain, pH) %>%
  filter(n_distinct(date) > 1) %>%
  ungroup()

anova_results <- data_filtered %>%
  group_by(strain, pH) %>%
  nest() %>%
  mutate(aov_result = map(data, ~aov(r ~ date, data = .)),
         anova_summary = map(aov_result, tidy)) %>%
  ungroup()


final.table <- anova_results %>%
  select(strain, pH, anova_summary) %>%
  mutate(anova_summary = map(anova_summary, ~filter(.x, term == "date"))) %>%
  unnest(anova_summary)

final.table <- merge(final.table, clade_info, by = "strain", all.x = T)

write.csv(final.table, file = here("anova_pH_results.csv"), row.names = FALSE)

## create strings to color by:
## clade:
clade_col <- as.character(all.data$clade_color)
names(clade_col) <- as.character(all.data$clade)
# strain by clade color
strain_clade_col <- as.character(all.data$clade_color)
names(strain_clade_col) <- as.character(all.data$strain)

## subclade:
subclade_col <- as.character(all.data$subclade_color)
names(subclade_col) <- as.character(all.data$subclade)
# strain by clade color
strain_subclade_col <- as.character(all.data$subclade_color)
names(strain_subclade_col) <- as.character(all.data$strain)



ggplot(final.table,
       aes(x = pH,
                    color = clade,
                    group = clade,
                    y = p.value)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.05, color = "red", linetype = "dashed") +
  scale_color_manual(values = clade_col) +
  #ggbreak::scale_y_cut(c(0.01), which = c(2), scales = c(1, 3)) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.3)) +
  #ylim(0, 0.5) +
  labs(title = "p Values for ANOVA results testing the effect of date (of the assay) on the growth rate for each pH and strain") +
  ylab("p value") +
  facet_wrap(strain~.) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  

```

### pH preference curves for r, K and AUC

Mean growth rate  (r) carrying capacity (K) and Area Under the Curve (AUC), and  S.D. for each parameter, for each combination of strain and pH level.

```{r summarized pH curve, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}
table.sum<-all.data %>%
  filter(!is.na(replicate)) %>%
  dplyr::group_by(strain, pH) %>%
  dplyr::summarize(r_mean = mean(r),
                   r_sd = sd(r),
                   k_mean = mean(k),
                   k_sd = sd(k),
                   auc_e_mean = mean(auc_e),
                   auc_e_sd = sd(auc_e)) %>%
  merge(., clade_info, by = "strain", all.x = T)


table.sum.date<-all.data %>%
  filter(!is.na(replicate)) %>%
  dplyr::group_by(strain, pH, date) %>%
  dplyr::summarize(r_mean = mean(r),
                   r_sd = sd(r),
                   k_mean = mean(k),
                   k_sd = sd(k),
                   auc_e_mean = mean(auc_e),
                   auc_e_sd = sd(auc_e)) %>%
  merge(., clade_info, by = "strain", all.x = T)


# make a stable id you can join on
all.data <- all.data %>%
  mutate(id_sdp = paste(strain, date, plate, sep = "|"))


random_ids <- all.data %>%
  filter(!is.na(replicate)) %>%
  group_by(strain, date, plate) %>%
  dplyr::sample_n(1) %>%       # dplyr >=1.1: use slice_sample not sample_n
  ungroup() %>%
  pull(id_sdp)


random_data <- all.data %>%
  filter(id_sdp %in% random_ids) %>%
  group_by(strain, pH) %>%
  summarise(
    r_mean     = mean(r, na.rm = TRUE),
    r_sd       = sd(r, na.rm = TRUE),
    k_mean     = mean(k, na.rm = TRUE),
    k_sd       = sd(k, na.rm = TRUE),
    auc_e_mean = mean(auc_e, na.rm = TRUE),
    auc_e_sd   = sd(auc_e, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  merge(clade_info, by = "strain", all.x = TRUE) %>%
  filter(!strain %in% c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03"))


## pH preference curve values averaged by date of the assay for each pH vale and strain

pH_curve_all_mean_by_date<-ggplot(table.sum.date |> mutate(date = as.factor(date))
                 ,
                 aes(x = pH,
                    color = subclade,
                    group = subclade,
                    y = r_mean,
                    shape = date)) +
  scale_color_manual(values = subclade_col) +
  geom_smooth(aes(fill = subclade)) +
  scale_fill_manual(values = subclade_col) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = r_mean-r_sd, ymax = r_mean+r_sd), width = .2) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_wrap(clade~strain) +
  ylim(0, 0.6) +
  labs(title = "pH preference curve for growth rate (r)",
       subtitle = "avg. r by pH level and date") +
  ylab("Growth rate (r)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        plot.title = element_text(color = "black", size = 12),
        plot.subtitle = element_text(color = "black", size = 10),
        legend.position = "bottom")

print(pH_curve_all_mean_by_date)


pH_curve_all<-ggplot(all.data %>% filter(!is.na(replicate)) |> mutate(date = as.factor(date))
                 ,
                 aes(x = pH,
                    color = subclade,
                    group = subclade,
                    y = r
                    , shape = date
                    )) +
  scale_color_manual(values = subclade_col) +
  geom_smooth(aes(fill = subclade)) +
  scale_fill_manual(values = subclade_col) +
  geom_point(size = 2) +
  scale_shape_manual(values = c(0:30)) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_wrap(clade~strain) +
  ylim(0, 0.6) +
  labs(title = "pH preference curve for growth rate (r)",
       subtitle = "all values") +
  ylab("Growth rate (r)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        plot.title = element_text(color = "black", size = 12),
        plot.subtitle = element_text(color = "black", size = 10),
        legend.position = "bottom")

print(pH_curve_all)


strains_to_filter <- c("25G4A05", "18G3A05", "23G3B04", "04G1A01", "24G2B02", "35D04", "35G4A01", "18G4B09", "24G4A05", "25G4B06", "28G1A08", "41G3A07", "18G1A07", "23G1B10", "24G2B09", "28G4A05", "28G4B01")

bad_strains <- table.sum %>%
  group_by(strain) %>%
  summarise(
    n_total = n(),
    n_zero  = sum(r_mean == 0, na.rm = TRUE),
    frac_zero = n_zero / n_total
  ) %>%
  filter(frac_zero >= 0.9)

strains_to_filter <- unique(c(strains_to_filter, bad_strains$strain))


for (cld in c("I", "II","III", "IV")) {
  
   # get all strains in this subclade
  all_strains <- table.sum %>%
    filter(!strain %in% strains_to_filter &
             clade != "UND" &
             !is.na(subclade)) %>%
    filter(clade == cld) %>%
    arrange(subclade) |> 
    pull(strain) %>%
    unique()
  
   # split into chunks of up to 16
  strain_chunks <- split(all_strains, 
                         ceiling(seq_along(all_strains) / 16))
  
  # loop over each chunk (page)
  for (i in seq_along(strain_chunks)) {
    this_strains <- strain_chunks[[i]]

    pH_curve <- table.sum |>
      filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
      filter(clade == cld, strain %in% this_strains) |> 
      mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean)) |> 
      ggplot(aes(x = pH,
                          color = subclade,
                          group = subclade,
                          y = r_mean)) +
        scale_color_manual(values = subclade_col, guide = "none") +
        geom_smooth(aes(fill = subclade), linewidth = 0.5) +
        scale_fill_manual(values = subclade_col, guide = "none") +
        geom_point(size = 1) +
        geom_errorbar(aes(ymin = r_mean-r_sd, ymax = r_mean+r_sd), width = .2) +
        scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
        ggh4x::facet_nested_wrap(subclade~strain, ncol = 4) +
        ylim(0, 0.6) +
        ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
        apatheme +
        theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
              text = element_text(size = 8),
              strip.background = element_blank(),
              #strip.text.x = element_blank(),
              plot.title = element_text(color = "black", size = 8),
              plot.subtitle = element_text(color = "black", size = 6),
              legend.position = "bottom")
      
      print(pH_curve)
      
      # now save it
    fn <- here(sprintf("plots/pH_curves/pH_curves_%s_page%02d.png", cld, i))
    ggsave(
      filename = fn,
      plot     = pH_curve,
      width    = 160,
      height   = 120,
      units    = "mm",
      dpi      = 300
    )
    }
  }
  
  
  
  
   pH_curve_rndm_clades<-ggplot(random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
  mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean))
                 ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = r_mean)) +
  scale_color_manual(values = strain_clade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_clade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_grid(~clade, space = "free") +
  ylim(0, 0.6) +
  ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 20),
        strip.background.x = element_blank(),
        text = element_text(size = 35),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none")

print(pH_curve_rndm_clades)

ggsave(here("plots/pH_curve_1rp_by_date_clades.png"),
       width = 30,
       height = 10,
       units = "cm",
       dpi = 600)



clade_cols <- c(
  "I"   = "#7171FF",
  "II"  = "#EEC900",
  "III" = "#FF0000",
  "IV"  = "#DC89FF"
)



pH_curve_rndm_subclades<-ggplot(random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
  mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean))
                 ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = r_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  # here’s the magic:
  ggh4x::facet_nested( ~clade + subclade,
    nest_line = element_line(linetype = 2),
    drop       = TRUE, # <— ensure unused combos get dropped
    strip     = ggh4x::strip_nested(
      # background for the top-level (clade) and second-level (subclade) strips
      background_x = ggh4x::elem_list_rect(
        fill = c(clade_cols, rep("white", 14)),  # one rect per clade level
        by_layer_x = F)
    )
  ) +
  ylim(0, 0.6) +
  ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 20),
        strip.background.x = element_blank(),
        text = element_text(size = 35),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none")

print(pH_curve_rndm_subclades)

ggsave(here("plots/pH_curve_1rp_by_date_subclades.png"),
       width = 60,
       height = 10,
       units = "cm",
       dpi = 600)


pH_curve_rndm_TopSubclades <- random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade) &
                                 subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) %>%
  mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean)) |> 
  ggplot(
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = r_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  # here’s the magic:
  ggh4x::facet_nested( ~subclade) +
  ylim(0, 0.6) +
  ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 16),
        strip.background.x = element_blank(),
        text = element_text(size = 14),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none") 

print(pH_curve_rndm_TopSubclades)

ggsave(here("plots/pH_curve_1rp_by_date_TopSubclades.png"),
       width = 30,
       height = 7.5,
       units = "cm",
       dpi = 600)

random_avg <- random_data |>
  group_by(subclade, pH) |>
  summarise(avg_r = mean(r_mean)) |>
  mutate(avg_r = if_else(pH %in% c(9.8, 10.1) & avg_r > 0.00001, 0, avg_r))



pH_curve_avg_subclades <- random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade) &
                                 subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) %>%
   mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean)) |>
   ggplot(aes(x = pH,
                    color = strain,
                    group = strain,
                    y = r_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.1, alpha = 0.25) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = c(3.2, 6.6, 10.1)) +
      # 2) tell ggplot we want a *new* colour scale
      ggnewscale::new_scale_color() +
      geom_smooth(data = random_avg |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")),
                 aes(pH, avg_r, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = T) +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_colors_vec) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_colors_vec) +
      ylim(0, 0.6) +
      facet_wrap(.~ subclade, nrow = 1) +
      ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
      xlab("pH") +
      apatheme +
      theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 4.5),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            #strip.text.x = element_blank(),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(pH_curve_avg_subclades)

```

### pH curves for K

```{r pH curves for K, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}


pH_curve_k<-ggplot(table.sum,
                 aes(x = pH,
                    color = subclade,
                    group = subclade,
                    y = k_mean)) +
  scale_color_manual(values = subclade_col) +
  geom_smooth(aes(fill = subclade)) +
  scale_fill_manual(values = subclade_col) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = k_mean-k_sd, ymax = k_mean+k_sd), width = .2) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_wrap(strain~.) +
  ylim(0, 0.6) +
  labs(title = "pH preference curve for carrying capacity (K)",
       subtitle = "avg. K by pH level") +
  ylab("Carrying capacity (K)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        plot.title = element_text(color = "black", size = 12),
        plot.subtitle = element_text(color = "black", size = 10),
        legend.position = "bottom")

print(pH_curve_k)


for (cld in c("I", "II","III", "IV")) {
  
   # get all strains in this subclade
  all_strains <- table.sum %>%
    filter(!strain %in% strains_to_filter &
             clade != "UND" &
             !is.na(subclade)) %>%
    filter(clade == cld) %>%
    arrange(subclade) |> 
    pull(strain) %>%
    unique()
  
   # split into chunks of up to 16
  strain_chunks <- split(all_strains, 
                         ceiling(seq_along(all_strains) / 16))
  
  # loop over each chunk (page)
  for (i in seq_along(strain_chunks)) {
    this_strains <- strain_chunks[[i]]

    pH_curve <- table.sum |>
      filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
      filter(clade == cld, strain %in% this_strains) |> 
      mutate(k_mean = if_else(pH %in% c(9.8, 10.1) & k_mean > 0.00001, 0, k_mean)) |> 
      ggplot(aes(x = pH,
                          color = subclade,
                          group = subclade,
                          y = k_mean)) +
        scale_color_manual(values = subclade_col, guide = "none") +
        geom_smooth(aes(fill = subclade), linewidth = 0.5) +
        scale_fill_manual(values = subclade_col, guide = "none") +
        geom_point(size = 1) +
        geom_errorbar(aes(ymin = k_mean-k_sd, ymax = k_mean+k_sd), width = .2) +
        scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
        ggh4x::facet_nested_wrap(subclade~strain, ncol = 4) +
        ylim(0, 0.6) +
        labs(y = expression("Yield (OD" [600] * ")")) +
        apatheme +
        theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
              text = element_text(size = 8),
              strip.background = element_blank(),
              #strip.text.x = element_blank(),
              plot.title = element_text(color = "black", size = 8),
              plot.subtitle = element_text(color = "black", size = 6),
              legend.position = "bottom")
      
      print(pH_curve)
      
      # now save it
    fn <- here(sprintf("plots/pH_curves/pH_curves_K_%s_page%02d.png", cld, i))
    ggsave(
      filename = fn,
      plot     = pH_curve,
      width    = 160,
      height   = 120,
      units    = "mm",
      dpi      = 300
    )
    }
  }
  
  
  
  
   pH_curve_rndm_clades_K<-ggplot(random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
  mutate(k_mean = if_else(pH %in% c(9.8, 10.1) & k_mean > 0.00001, 0, k_mean))
                 ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = k_mean)) +
  scale_color_manual(values = strain_clade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_clade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_grid(~clade, space = "free") +
  ylim(0, 0.6) +
  labs(y = expression("Yield (OD" [600] * ")")) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 20),
        strip.background.x = element_blank(),
        text = element_text(size = 35),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none")

print(pH_curve_rndm_clades_K)

ggsave(here("plots/pH_curve_1rp_by_date_clades_K.png"),
       width = 30,
       height = 10,
       units = "cm",
       dpi = 600)



clade_cols <- c(
  "I"   = "#7171FF",
  "II"  = "#EEC900",
  "III" = "#FF0000",
  "IV"  = "#DC89FF"
)



pH_curve_rndm_subclades_K<-ggplot(random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
  mutate(k_mean = if_else(pH %in% c(9.8, 10.1) & k_mean > 0.00001, 0, k_mean))
                 ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = k_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  # here’s the magic:
  ggh4x::facet_nested( ~clade + subclade,
    nest_line = element_line(linetype = 2),
    drop       = TRUE, # <— ensure unused combos get dropped
    strip     = ggh4x::strip_nested(
      # background for the top-level (clade) and second-level (subclade) strips
      background_x = ggh4x::elem_list_rect(
        fill = c(clade_cols, rep("white", 14)),  # one rect per clade level
        by_layer_x = F)
    )
  ) +
  ylim(0, 0.6) +
  labs(y = expression("Yield (OD" [600] * ")")) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 20),
        strip.background.x = element_blank(),
        text = element_text(size = 35),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none")

print(pH_curve_rndm_subclades_K)

ggsave(here("plots/pH_curve_1rp_by_date_subclades_K.png"),
       width = 60,
       height = 10,
       units = "cm",
       dpi = 600)


pH_curve_rndm_TopSubclades_K <- random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade) &
                                 subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) %>%
  mutate(k_mean = if_else(pH %in% c(9.8, 10.1) & k_mean > 0.00001, 0, k_mean)) |> 
  ggplot(
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = k_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.85, alpha = 0.65) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  # here’s the magic:
  ggh4x::facet_nested( ~subclade) +
  ylim(0, 0.6) +
  labs(y = expression("Yield (OD" [600] * ")")) +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        axis.text.y = element_text(size = 16),
        strip.background.x = element_blank(),
        text = element_text(size = 14),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 18),
        plot.subtitle = element_text(color = "black", size = 16),
        legend.position = "none") 

print(pH_curve_rndm_TopSubclades_K)

ggsave(here("plots/pH_curve_1rp_by_date_TopSubclades_K.png"),
       width = 30,
       height = 7.5,
       units = "cm",
       dpi = 600)

random_avg_k <- random_data |>
  group_by(subclade, pH) |>
  summarise(avg_k = mean(k_mean)) |>
  mutate(avg_k = if_else(pH %in% c(9.8, 10.1) & avg_k > 0.00001, 0, avg_k))


```

##### Figures included in the publication
###### Single pH curves per subclades and avg. curve for top subclades

The per-subclade average-yield (K) curves below (`pH_curve_avg_subclades_K`, `pH_curve_avgOnly_subclades_K`) are included in the publication.

- Main figure from this: `pH_curve_avgOnly_subclades_K`

```{r figs-not-in-pub-K-curves, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}
pH_curve_avg_subclades_K <- random_data %>%
                        filter(!strain %in% strains_to_filter &
                                 clade != "UND" & !is.na(subclade) &
                                 subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) %>%
   mutate(k_mean = if_else(pH %in% c(9.8, 10.1) & k_mean > 0.00001, 0, k_mean)) |>
   ggplot(aes(x = pH,
                    color = strain,
                    group = strain,
                    y = k_mean)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.1, alpha = 0.25) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = #seq(3.2, 10.1, 0.6)
                       c(3.2, 6.6, 10.1)) +
      # 2) tell ggplot we want a *new* colour scale
      ggnewscale::new_scale_color() +
      geom_smooth(data = random_avg_k |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")),
                 aes(pH, avg_k, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = T) +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_colors_vec) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_colors_vec) +
      ylim(0, 0.4) +
      facet_wrap(.~ subclade, nrow = 1) +
      labs(y = expression("Yield (OD" [600] * ")")) +
      xlab("pH") +
      apatheme +
      theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 4.5),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(pH_curve_avg_subclades_K)


pH_curve_avgOnly_subclades_K <- random_avg_k |>
  filter(subclade  %in% c("IC", "IIG", "IIIA", "IVB")) |>
   ggplot() +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
      geom_smooth(
                 aes(pH, avg_k, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = F,
                key_glyph = "rect") +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_colors_vec) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_colors_vec) +
          # change legend keys from line to square
      guides(
        colour = guide_legend(
          override.aes = list(
            shape = 22,   # 22 = filled square (uses both color & fill)
            size  = 1,
            linetype = 2,  # remove line segment in legend
            fill = clade_color_vector[names(clade_color_vector)  %in% c("IC", "IIG", "IIIA", "IVB")],
            alpha = 0.95
          )
        ),
        fill = "none"     # hide separate fill legend, keep just one
      ) +
      ylim(0, 0.5) +
      labs(y = expression("Yield (OD" [600] * ")")) +
      xlab("pH") +
      apatheme +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 6),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            legend.position = "inside",
            legend.position.inside = c(0.8, 0.85),
            legend.background = element_rect(fill = alpha("white", 0), color = NA),
            legend.key = element_rect(fill = NA),
            legend.key.size = unit(0.3, "lines"),
            legend.title = element_text(size = 11,
                                        margin = margin(b = 0, unit = "pt")),
            legend.text = element_text(size = 10),
            legend.key.height = unit(0.4, "lines"),
            legend.key.width = unit(0.7, "lines"),
            legend.spacing.y = unit(0, "cm"))
print(pH_curve_avgOnly_subclades_K)
```

```{r pH-curves-for-K-cont, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}



pH_curve_auc<-ggplot(table.sum,
                 aes(x = pH,
                    color = subclade,
                    group = subclade,
                    y = auc_e_mean)) +
  scale_color_manual(values = subclade_col) +
  geom_smooth(aes(fill = subclade)) +
  scale_fill_manual(values = subclade_col) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = auc_e_mean-auc_e_sd, ymax = auc_e_mean+auc_e_sd), width = .2) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
  facet_wrap(strain~.) +
  #ylim(0, 0.6) +
  labs(title = "pH preference curve for Area Under the Curve (AUC)",
       subtitle = "avg.AUC by pH level") +
  ylab("Area under the curve (AUC)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        plot.title = element_text(color = "black", size = 12),
        plot.subtitle = element_text(color = "black", size = 10),
        legend.position = "bottom")

print(pH_curve_auc)

```

##### Summary of strains assayed

```{r}

setdiff(strains_to_filter, c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03"))
intersect(strains_to_filter, c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03"))
union(strains_to_filter, c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03"))


length(unique(c(strains_to_filter, c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03"))))

length(unique(all.data$strain))
length(unique(all.data$strain)[!unique(all.data$strain)  %in% strains_to_filter])

length(strains_to_filter)


all_strains_to_filter <- unique(c(strains_to_filter, c("23G1B10", "23G4B11", "24G2B09", "28G4A05", "28G4B03")))

length(all_strains_to_filter)


## ok strains
all.data |> distinct(strain, .keep_all = T) |> 
  filter(!strain %in% all_strains_to_filter) |> 
  group_by(clade) |> count()

all.data |> distinct(strain, .keep_all = T) |> 
  filter(!strain %in% all_strains_to_filter) |> 
  group_by(subclade) |> count()

## not ok strains

all.data |> distinct(strain, .keep_all = T) |> 
  filter(strain %in% all_strains_to_filter) |> 
  group_by(clade) |> count()

all.data |> distinct(strain, .keep_all = T) |> 
  filter(strain %in% all_strains_to_filter) |> 
  group_by(subclade) |> count()


for (cld in c("I", "II","III", "IV")) {
  
   # get all strains in this subclade
  all_strains <- table.sum %>%
    filter(strain %in% all_strains_to_filter &
             clade != "UND" &
             !is.na(subclade)) %>%
    filter(clade == cld) %>%
    arrange(subclade) |> 
    pull(strain) %>%
    unique()
  
   # split into chunks of up to 16
  strain_chunks <- split(all_strains, 
                         ceiling(seq_along(all_strains) / 16))
  
  # loop over each chunk (page)
  for (i in seq_along(strain_chunks)) {
    this_strains <- strain_chunks[[i]]

    pH_curve <- table.sum |>
      filter(strain %in% all_strains_to_filter &
                                 clade != "UND" & !is.na(subclade)) %>%
      filter(clade == cld, strain %in% this_strains) |> 
      mutate(r_mean = if_else(pH %in% c(9.8, 10.1) & r_mean > 0.00001, 0, r_mean)) |> 
      ggplot(aes(x = pH,
                          color = subclade,
                          group = subclade,
                          y = r_mean)) +
        scale_color_manual(values = subclade_col, guide = "none") +
        geom_smooth(aes(fill = subclade), linewidth = 0.5) +
        scale_fill_manual(values = subclade_col, guide = "none") +
        geom_point(size = 1) +
        geom_errorbar(aes(ymin = r_mean-r_sd, ymax = r_mean+r_sd), width = .2) +
        scale_x_continuous(breaks = seq(3.2, 10.1, 0.6)) +
        ggh4x::facet_nested_wrap(subclade~strain, ncol = 4) +
        ylim(0, 0.6) +
        ylab("Growth rate (r)") +
        apatheme +
        theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
              text = element_text(size = 8),
              strip.background = element_blank(),
              #strip.text.x = element_blank(),
              plot.title = element_text(color = "black", size = 8),
              plot.subtitle = element_text(color = "black", size = 6),
              legend.position = "bottom")
      
      print(pH_curve)
      
    }
  }



#### now let's get a table to map onto the tree which strains have been tested and

all_strains_to_filter <- c(all_strains_to_filter, "04D05", "04G2A08")


ok_strains <- all.data |> distinct(strain, .keep_all = T) |> 
  filter(!strain %in% all_strains_to_filter) |> 
  mutate(pH_assay_ok = "YES") |>
  select(strain, pH_assay_ok)

not_ok_strains <- all.data |> distinct(strain, .keep_all = T) |> 
  filter(strain %in% all_strains_to_filter) |> 
  mutate(pH_assay_ok = "NO") |>
  select(strain, pH_assay_ok)


pH_ok_strains <- bind_rows(ok_strains, not_ok_strains)


write.table(pH_ok_strains,
            file = here("output/pH/pH_assay_ok_strain_table.txt"),
            quote = F,
            row.names = F,
            col.names = T,
            sep = "\t")



## let's pull some of the positive controls for the Magnesium test

control_strains <- 
  all.data |>
  filter(strain %in%  c("25G2A02", "26G4B12", "23G4B03", "18G4B09")) |>
  select(strain, date) |>
  distinct(.keep_all = T)

all.data |>
  filter(strain %in%  "23G4B03") |>
  select(strain, date) |>
  distinct(.keep_all = T)

```

#### pH curves parameters calculation

Now we can calculate the pH curves parameters for each strain:

  1. Area under the curve
  2. Range
  3. Optimal pH


```{r calculate curve paramters, message=TRUE, warning=TRUE}
library(pracma)

head(random_data)


random_data$r_mean[random_data$pH %in% c(9.8, 10.1) & random_data$r_mean > 0.0001] <- 0
random_data$k_mean[random_data$pH %in% c(9.8, 10.1) & random_data$k_mean > 0.0001] <- 0



# Group by strain and calculate area under the curve for each strain
result_auc <- random_data %>%
  group_by(strain) %>%
  group_split() %>%
  lapply(calculate_auc) %>%
  bind_rows()

# Print the result data frame
print(result_auc)



# Apply the function for each strain in the result_df
max_rk_mean_results <- random_data %>%
  group_by(strain) %>%
  do(get_max_rk_mean(.))

# Print the results
print(max_rk_mean_results)



result_range_r <- random_data %>%
  filter(r_mean != 0) %>%
  group_by(strain) %>%
  summarise(min_pH_r = min(pH),
            max_pH_r = max(pH),
            breadth_r = max(pH) - min(pH))

print(result_range_r)

result_range_k <- random_data %>%
  filter(k_mean != 0) %>%
  group_by(strain) %>%
  summarise(min_pH_k = min(pH),
            max_pH_k = max(pH),
            breadth_k = max(pH) - min(pH))

print(result_range_k)



curve_params <- merge(result_auc, max_rk_mean_results, by = "strain") %>%
  merge(., result_range_r, by = "strain") %>%
  merge(., result_range_k, by = "strain") %>%
  filter(!strain %in% strains_to_filter)

#write.table(as.data.frame(curve_params$strain), file = "~/SyncFolder/CA_curto/isolates/fastANI_out_WI/working_isolates.tsv", row.names = FALSE, quote = F, sep = "\t")

write.table(as.data.frame(curve_params |>
                            select(strain) |>
                            rename(StrainID = strain) |>
                            mutate(pH = "yes")),
            file = here("output/pH/pH_strain_table.txt"), row.names = FALSE, quote = F, sep = "\t")


### PCA of curve parameters
curve_params <- curve_params %>% tibble::column_to_rownames("strain")


r_traits <- c(colnames(curve_params |> select(ends_with("_r") & !starts_with("area"))))
k_traits <- c(colnames(curve_params |> select(ends_with("_k") & !starts_with("area"))))

pca <-prcomp(curve_params |>
               tibble::rownames_to_column("strain") |> 
               left_join(clade_info, by = "strain") |>
               filter(!is.na(clade),
                      clade != "UND",
                      !is.na(subclade)) |>
               filter(subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVC")) |>
               tibble::column_to_rownames("strain") |>
               select(all_of(r_traits))
             , scale. = T, center = T)


pca.df <- data.frame(pca$x) %>%
  select(PC1, PC2) %>%
  merge(., master_md, by.x = "row.names", by.y = "Name", all.x = T)

pca.df$clade_color <- clade_colors$color[match(pca.df$clade, clade_colors$variable)]
pca.df$subclade_color <- clade_colors$color[match(pca.df$subclade, clade_colors$variable)]
PCAloads <- data.frame(Variables = rownames(pca$rotation), pca$rotation) |>
  mutate(Variables = recode(Variables,
                            "min_pH_r" = "Min. pH",
                            "opt_pH_r" = "Opt. pH",
                            "max_pH_r" = "Max. pH",
                            "max_r_mean" = "Growth rate",
                            "breadth_r" = "Breadth"))

pca_plot<-ggplot(pca.df
                 , aes(PC1, PC2)) +
  geom_point(data = pca.df, aes(color = subclade),  size = 2, alpha = 0.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               arrow = arrow(length = unit(0.3, "cm")),
               colour="grey", inherit.aes = FALSE) +
 geom_label(data=PCAloads,
           aes(x=PC1*2.65, y=PC2*2.65, label=Variables),
           size=3,
           colour = "darkblue") +
  xlab(paste0("PC1 (",round(summary(pca)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 10),
        legend.position = "inside",
        legend.position.inside = c(0.85, 0.8),
        legend.background = element_rect(fill = alpha("white", 0.3), color = NA),
        legend.key.size = unit(0.5, "lines"))
pca_plot

ggsave(
      filename = here("plots/pca_curve_params_r_topSubclades.pdf"),
      plot     = pca_plot,
      width    = 90,
      height   = 90,
      units    = "mm",
      dpi      = 300,
      device = cairo_pdf
    )
  

summary(pca)


pca_k <-prcomp(curve_params |>
               tibble::rownames_to_column("strain") |> 
               left_join(clade_info, by = "strain") |>
               filter(!is.na(clade),
                      clade != "UND",
                      !is.na(subclade)) |>
               filter(subclade %in% c("IC", "IIG", "IIIA",
                                      #"IIIE",
                                      "IVB"#, "IVC"
                                      )) |>
               tibble::column_to_rownames("strain") |>
               select(all_of(k_traits))
             , scale. = T, center = T)


pca.df_k <- data.frame(pca_k$x) %>%
  select(PC1, PC2) %>%
  merge(., master_md, by.x = "row.names", by.y = "Name", all.x = T)

pca.df_k$clade_color <- clade_colors$color[match(pca.df_k$clade, clade_colors$variable)]
pca.df_k$subclade_color <- clade_colors$color[match(pca.df_k$subclade, clade_colors$variable)]
PCAloads_k <- data.frame(Variables = rownames(pca_k$rotation), pca_k$rotation) |>
  mutate(Variables = recode(Variables,
                            "min_pH_k" = "Min. pH",
                            "opt_pH_k" = "Opt. pH",
                            "max_pH_k" = "Max. pH",
                            "max_k_mean" = "Yeild",
                            "breadth_k" = "Breadth"))

pca_plot_k<-ggplot(pca.df_k
                 , aes(PC1, PC2)) +
  geom_point(data = pca.df_k, aes(color = subclade),  size = 1.5, alpha = 0.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads_k,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               colour="grey", inherit.aes = FALSE) +
   geom_text(data=PCAloads_k,
           aes(x=PC1*4.5, y=PC2*4.5, label=Variables),
           size=2,
           colour = "black",
           fontface = "bold") +
  lims(x = c(-4, 4),
       y = c(-6, 6)) +
  xlab(paste0("PC1 (",round(summary(pca_k)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca_k)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 8),
        legend.position = "none",
        legend.position.inside = c(0.9, 0.2),
        legend.background = element_rect(fill = alpha("white", 0), color = NA),
        legend.key.size = unit(0.5, "lines"),
        legend.title = element_text(size = 6,
                                    margin = margin(b = 0, unit = "pt")),
        legend.text = element_text(size = 5),
        legend.key.height = unit(0.4, "lines"),
        legend.key.width = unit(0.7, "lines"),
        legend.spacing.y = unit(0, "cm"))

pca_plot_k

ggsave(
      filename = here("plots/pca_curve_params_K_topSubclades.pdf"),
      plot     = pca_plot_k,
      width    = 90,
      height   = 90,
      units    = "mm",
      dpi      = 300,
      device = cairo_pdf
    )


summary(pca_k)


curve_params |>
               tibble::rownames_to_column("strain") |> 
               left_join(clade_info, by = "strain") |>
               filter(!is.na(clade),
                      clade != "UND",
                      !is.na(subclade)) |>
               filter(subclade %in% c("IC", "IIG", "IIIA",
                                      #"IIIE",
                                      "IVB"#, "IVC"
                                      )) |>
  summarise(avg.pHopt = mean(opt_pH_k, na.rm. = T),
            avg.minpH = mean(min_pH_k, na.rm. = T),
            avg.maxpH = mean(max_pH_k, na.rm. = T),
            avg.breadth = mean(breadth_k, na.rm. = T))

curve_params |>
               tibble::rownames_to_column("strain") |> 
               left_join(clade_info, by = "strain") |>
               filter(!is.na(clade),
                      clade != "UND",
                      !is.na(subclade)) |>
               filter(subclade %in% c("IC", "IIG", "IIIA",
                                      #"IIIE",
                                      "IVB"#, "IVC"
                                      )) |>
  group_by(subclade) |>
  summarise(avg.pHopt = mean(opt_pH_k, na.rm. = T),
            avg.minpH = mean(min_pH_k, na.rm. = T),
            avg.maxpH = mean(max_pH_k, na.rm. = T),
            avg.breadth = mean(breadth_k, na.rm. = T))

D <- dist(pca_k$x, "euclidean")

perm <- vegan::adonis2(D ~ subclade, data = pca.df_k, permutations = 999)

perm



```


#### heat maps of curve parameters

```{r heat maps, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}
head(curve_params)
library(ggh4x)
# Add a new column for rownames (Strain IDs)
curve_params$Strain <- rownames(curve_params)

curve_params <- merge(curve_params, clade_info, by.x = "Strain", by.y = "strain", all.x = T)

# Arrange the data by the clade variable
curve_params <- curve_params %>%
  arrange(clade, subclade, Strain) #%>% filter(Strain != "33G1A03")

write.table(as.data.frame(curve_params |>
                            select(Strain) |>
                            rename(StrainID = Strain) |>
                            mutate(pH = "yes")),
            file = here("output/pH/pH_strain_table.txt"), row.names = FALSE, quote = F, sep = "\t")



# Assuming your data is stored in a dataframe called curve_params
# Convert the data to long format
curve_params_long <- pivot_longer(curve_params,
                                  cols = where(is.numeric),
                                  names_to = "Parameter",
                                  values_to = "Value")

# Scale the data to z-scores by grouping by 'Parameter'
curve_params_long <- curve_params_long %>%
  group_by(Parameter) %>%
  mutate(Value_z = scale(Value)) %>%
  ungroup()


clade_color_vector <- setNames(clade_colors$color, clade_colors$variable)


# Create the heatmap
ggplot(curve_params_long %>% filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r") & clade != "UND")
       ,
       aes(x = Parameter,
                              y = factor(Strain, levels = unique(Strain)),
                              fill = Value_z)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +  # Adjust colors as necessary
  labs(x = "Parameter", y = "Strain", fill = "Z-score") +
  ggh4x::facet_grid2(clade ~ ., scales = "free_y", space = "fixed", switch = "y",
              strip = ggh4x::strip_themed(
                background_y = ggh4x::elem_list_rect(
                  fill = clade_color_vector[unique(curve_params_long$clade)]
                )
              )) +
  scale_x_discrete(position = "top", labels = c("Max. pH", "rmax", "Opt. pH")) +
  apatheme +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9, face = "bold"),  # Align text horizontally and center
    axis.text.y = element_blank(),  # Remove y-axis text
    axis.ticks.y = element_blank(),  # Remove y-axis ticks
    strip.placement = "inside",  # Place facet strips outside the plot area
    strip.text = element_text(face = "bold"),  # Bold text for facet labels
    panel.spacing = unit(0, "lines"),  # Adjust spacing between panels
    strip.background = element_blank(),  # Remove the default strip background
    strip.text.y.left = element_text(angle = 0, vjust = 1, hjust = 1),  # Align facet labels on the left
    legend.title=element_text(size=11), 
    legend.text=element_text(size=10),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.spacing = unit(0, "lines"),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    text = element_text(size = 22),
    panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank()# Ensure the y-axis title is removed
  ) +
   scale_y_discrete(position = "right") 
  # Align facet labels on the left

ggsave(here("plots/heatmap_curve_params.png"),
       width = 6,
       height = 12,
       units = "cm",
       dpi = 600)

# Create the horizontal heatmap
ggplot(curve_params_long %>% filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r") & clade != "UND"),
       aes(x = factor(Strain, levels = unique(Strain)),
           y = Parameter,
           fill = Value_z)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +  # Adjust colors as necessary
  labs(x = "Strain", y = "Parameter", fill = "Z-score") +
  ggh4x::facet_grid2(. ~ clade, scales = "free_x", space = "fixed",
                     strip = ggh4x::strip_themed(
                       background_x = elem_list_rect(
                         fill = clade_color_vector[unique(curve_params_long$clade)]
                       )
                     )) +
  scale_y_discrete(labels = c("Max. pH", "rmax", "Opt. pH")) +  # Adjust the y-axis labels
  theme_minimal() +
  theme(
    text = element_text(size = 22),
    axis.text.y = element_text(angle = 0, hjust = 0.5, size = 14, face = "bold"),  # Align text horizontally and center
    axis.text.x = element_blank(),  # Remove x-axis text (Strain)
    axis.ticks.x = element_blank(),  # Remove x-axis ticks (Strain)
    strip.placement = "inside",  # Place facet strips inside the plot area
    strip.text = element_text(face = "bold"),  # Bold text for facet labels
    panel.spacing = unit(0, "lines"),  # Adjust spacing between panels
    strip.background = element_blank(),  # Remove the default strip background
    strip.text.x.top = element_text(angle = 0, vjust = 1, hjust = 0.5),  # Align facet labels on the top
    legend.title = element_text(size = 11, margin = margin(b = 0)), 
    legend.text = element_text(size = 10),
    legend.position = "right",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.spacing = unit(0, "lines"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),  # Ensure the y-axis title is removed
    #plot.margin = margin(10, 10, 10, 10)  # Adjust plot margins if necessary
  )

ggsave(here("plots/heatmap_curve_params.png"),
       width = 27,
       height = 4,
       units = "cm",
       dpi = 600)


# Create the horizontal heatmap subclade level
ggplot(curve_params_long %>% filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r") & clade != "UND"),
       aes(x = factor(Strain, levels = unique(Strain)),
           y = Parameter,
           fill = Value_z)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +  # Adjust colors as necessary
  labs(x = "Strain", y = "Parameter", fill = "Z-score") +
  ggh4x::facet_grid2(. ~ subclade, scales = "free_x", space = "fixed",
                     strip = ggh4x::strip_themed(
                       background_x = elem_list_rect(
                         fill = clade_color_vector[unique(curve_params_long$subclade)]
                       )
                     )) +
  scale_y_discrete(labels = c("Max. pH", "rmax", "Opt. pH")) +  # Adjust the y-axis labels
  theme_minimal() +
  theme(
    text = element_text(size = 22),
    axis.text.y = element_text(angle = 0, hjust = 0.5, size = 14, face = "bold"),  # Align text horizontally and center
    axis.text.x = element_blank(),  # Remove x-axis text (Strain)
    axis.ticks.x = element_blank(),  # Remove x-axis ticks (Strain)
    strip.placement = "inside",  # Place facet strips inside the plot area
    strip.text = element_text(face = "bold"),  # Bold text for facet labels
    panel.spacing = unit(0, "lines"),  # Adjust spacing between panels
    strip.background = element_blank(),  # Remove the default strip background
    strip.text.x.top = element_text(angle = 0, vjust = 1, hjust = 0.5),  # Align facet labels on the top
    legend.title = element_text(size = 11, margin = margin(b = 0)), 
    legend.text = element_text(size = 10),
    legend.position = "right",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.spacing = unit(0, "lines"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),  # Ensure the y-axis title is removed
    #plot.margin = margin(10, 10, 10, 10)  # Adjust plot margins if necessary
  )

ggsave(here("plots/heatmap_curve_params_subclade.png"),
       width = 27,
       height = 4,
       units = "cm",
       dpi = 600)



write.table(curve_params |> select(-c(clade, subclade)),
            file = here("output/pH/curve_params.tsv"), row.names = FALSE, quote = F, sep = "\t")



#### dot plot



# Plot just two subclades, and save:
curve_params_dot_plot <- dotplot_traits_by_subclade(
  params_df  = curve_params_long,
  traits     = c("opt_pH_r", "max_r_mean", "max_pH_r"),
  subclades  = NULL,
  facet_by_clade = T,
  filename   = here("plots/curve_params_dotplot_Subclades.png")
)

# Plot just two subclades, and save:
curve_params_dot_plot_k <- dotplot_traits_by_subclade(
  params_df  = curve_params_long,
  traits     = c("breadth_k", "opt_pH_k", "max_k_mean", "max_pH_k"),
  subclades  = NULL,
  facet_by_clade = T,
  filename   = here("plots/curve_params_dotplot_Subclades_K.png")
)


# Plot just two subclades, and save:
curve_params_dot_plot_top <- dotplot_traits_by_subclade(
  params_df  = curve_params_long,
  traits     = c("opt_pH_r", "max_r_mean", "max_pH_r"),
  subclades  = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),
  facet_by_clade = F,
  filename   = here("plots/curve_params_dotplot_TopSubclades.png")
)

# Plot just two subclades, and save:
curve_params_dot_plot_top_k <- dotplot_traits_by_subclade(
  params_df  = curve_params_long,
  traits     = c("breadth_k","opt_pH_k", "max_k_mean", "max_pH_k"),
  subclades  = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),
  facet_by_clade = F,
  filename   = here("plots/curve_params_dotplot_TopSubclades_K.png")
)




posthocs <- curve_params_long %>%
  # only keep the subclades you care about
  filter(subclade %in% c("IC","IIG","IIIA","IIIE","IVB","IVE")) %>%
  group_by(Parameter) %>%
  do({
    dat <- .

    # 1) must have ≥2 subclades and each with ≥2 obs
    sub_counts <- dat %>% count(subclade) 
    if (n_distinct(dat$subclade) < 2 ||
        any(sub_counts$n < 2)) {
      return(tibble()) 
    }

    # 2) normality?
    if (nrow(dat) >= 3) {
      sw_p <- dat |> rstatix::shapiro_test(Value) |> pull(p)
    } else {
      sw_p <- NA_real_
    }

    # 3) homogeneity?
    lev_p <- rstatix::levene_test(dat, Value ~ subclade)$p

    # 4) choose test
    if (!is.na(sw_p) && sw_p > 0.05 && lev_p > 0.05) {
      # ANOVA + Tukey
      omni <- rstatix::anova_test(dat, Value ~ subclade)
      ph   <- rstatix::tukey_hsd(dat, Value ~ subclade) %>%
                mutate(method = "Tukey HSD")
      test_type <- "ANOVA"
    } else {
      # Kruskal–Wallis + Dunn
      omni <- rstatix::kruskal_test(dat, Value ~ subclade)
      ph   <- rstatix::dunn_test(dat, Value ~ subclade, p.adjust.method = "BH") %>%
                mutate(method = "Dunn (BH)")
      test_type <- "KW"
    }

    # 5) build omnibus row
    x_max <- max(dat$Value, na.rm = TRUE)
    omnibus_df <- tibble(
      Parameter        = unique(dat$Parameter),
      group1       = NA_character_,
      group2       = NA_character_,
      p.adj        = omni$p,
      p.adj.signif = as.character(
                                  symnum(omni$p,
                                         corr = F,
                                         cutpoints = c(0, 0.001, 0.01, 0.05, 1),
                                         symbols   = c("***","**","*","ns"))
                                  ),
      x.position   = x_max * 1.08,
      method       = test_type
      )
    
    # 6) position pairwise rows just below
    ph_df <- ph %>%
      mutate(
        Parameter         = unique(dat$Parameter),
        p.adj.signif = as.character(
                          symnum(p.adj,
                                 corr = F,
                                 cutpoints = c(0, 0.001, 0.01, 0.05, 1),
                                 symbols   = c("***","**","*","ns"))
                        ),
        x.position = x_max * 1.05)

    bind_rows(omnibus_df, ph_df)
  }) %>%
  ungroup()

# Inspect your posthocs table
posthocs

# 2. Compute max per panel so we can place omnibus labels
panel_ranges <- curve_params_long %>%
  filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r")) |>
  group_by(Parameter) %>%
  summarise(xmax = max(Value, na.rm=TRUE), .groups="drop") %>%
  mutate(xpos = xmax * 1.02)     # 2% beyond the data

# 4. Join in the xpos
posthocs <- posthocs %>%
  left_join(panel_ranges, by = "Parameter")


# 1) Pull out just your omnibus rows
posthocs_omni <- posthocs %>%
  filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r"),) |>
  filter(is.na(group1))


# 2) For each term, build a named vector of p’s and feed to multcompLetters
letter_df <- posthocs %>%
  filter(Parameter %in% c("opt_pH_r", "max_r_mean", "max_pH_r"),) |>
  filter(!is.na(group1)) %>% 
  select(Parameter, group1, group2, p.adj) |> 
  group_by(Parameter) %>%
  summarise(
    # build a named vector c("A-B"=p1, "A-C"=p2, ...)
    pv = list( setNames(p.adj, paste(group1, group2, sep = "-")) ),
    .groups = "drop"
  ) %>%
  mutate(
    # run multcompLetters on each vector
    letters = map(pv, ~ multcompView::multcompLetters(.x, threshold = 0.05)$Letters)
  ) %>%
  select(-pv) %>%
  unnest_longer(letters, indices_to = "subclade", values_to = "letter") |>
  left_join(panel_ranges, by = "Parameter")


# 2) Add the text directly
curve_params_dot_plot_top +
  geom_text(
    data        = posthocs_omni,
    aes(x        = x.position * 0.85,
        y        = 0.65,
        label    = p.adj.signif),
    inherit.aes = FALSE,
    size = 6,
    hjust = 0      # left‐justify so it sits just outside
  ) +
  geom_text(
    data        = letter_df,
    aes(x     = xpos * 0.1,
        y     = subclade,
        label = letter),
    inherit.aes = FALSE,
    size        = 5,
    hjust       = 0
  )


ggsave(
      filename = here("plots/curve_params_dotplot_TopSubclades.png"),
      width    = 180,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )






out_r <- plot_traits_subclade(
  params_df = curve_params |> select(-c(clade, subclade)),
  clade_info = clade_info,
  traits = c("breadth_r","opt_pH_r", "max_pH_r"),
  subclades = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    opt_pH_r  = "opt pH",
    max_pH_r  = "max pH",
    breadth_r = "breadth (pH)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("plots/curve_params_dotplot_Subclades_flip.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)

out_r_all_subclades <- plot_traits_subclade(
  params_df = curve_params |> select(-c(clade, subclade)),
  clade_info = clade_info,
  traits = c("breadth_r","opt_pH_r", "max_pH_r"),
  #subclades = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  exclude_subs = c("IIIB","IVA"),
  clade_levels = c("I", "II", "III", "IV"),
  #subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = T,
  flip = T,
  term_labels = c(
    opt_pH_r  = "Opt. pH",
    max_pH_r  = "Max. pH",
    breadth_r = "Breadth (pH)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("plots/curve_params_dotplot_Subclades_flip_r_all_subclades.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)





names(curve_params)
dplyr::glimpse(curve_params)
table(is.na(curve_params$subclade))



ggpubr::ggarrange(pH_curve_avg_subclades, out_r$plot,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = "AUTO",
                  hjust = 0.05,
                  font.label = list(size = 10))

ggsave(
      filename = here("plots/combined_avg_params_r.png"),
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )


#### for K

out_k <- plot_traits_subclade(
  params_df = curve_params |> select(-c(clade, subclade)),
  clade_info = clade_info,
  traits = c("breadth_k","opt_pH_k", "max_pH_k"),
  subclades = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    opt_pH_k  = "Opt. pH",
    max_pH_k  = "Max. pH",
    breadth_k = "Breadth (pH)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("plots/curve_params_dotplot_Subclades_flip_K.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)


```

##### Figures included in the publication pH
###### pH curves with K and dotplots and combined figures

The K-only trait dot plots (`out_k_all`, `out_k_all_subclades`) and the combined multi-panel figures below are included in the publication

```{r figs-not-in-pub-K-all, fig.width=18, fig.height = 14, message=FALSE, warning=FALSE, echo=FALSE}
#### for K_all

out_k_all <- plot_traits_subclade(
  params_df = curve_params |> select(-c(clade, subclade)),
  clade_info = clade_info,
  traits = c("opt_pH_k","max_pH_k","min_pH_k", "breadth_k", "max_k_mean"),
  subclades = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    opt_pH_k  = "opt pH",
    max_pH_k  = "max pH",
    breadth_k = "breadth (pH)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("plots/curve_params_dotplot_Subclades_flip_K_all.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)


out_k_all_subclades <- plot_traits_subclade(
  params_df = curve_params |> select(-c(clade, subclade)),
  clade_info = clade_info,
  traits = c("opt_pH_k","max_pH_k","min_pH_k", "breadth_k", "max_k_mean"),
  #subclades = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  exclude_subs = c("IIIB","IVA"),
  clade_levels = c("I", "II", "III", "IV"),
  #subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = T,
  flip = T,
  term_labels = c(
    opt_pH_k  = "opt pH",
    max_pH_k  = "max pH",
    breadth_k = "breadth (pH)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("plots/curve_params_dotplot_Subclades_flip_K_all_subclades.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)

names(curve_params)
dplyr::glimpse(curve_params)
table(is.na(curve_params$subclade))


# per-subclade optimal pH (where avg_k is maximal)
opt_by_sub <- curve_params |>
  dplyr::filter(subclade %in% c("IC","IIG","IIIA","IIIE","IVB","IVE")) |>
  dplyr::group_by(subclade) |>
  summarise(n = sum(!is.na(opt_pH_k)),
            mean_pHopt = mean(opt_pH_k),
            med_pHopt = median(opt_pH_k),
            pHopt_se = sd(opt_pH_k)/sqrt(n),
            pHopt_sd = sd(opt_pH_k),
            mean_maxPH = mean(max_pH_k)
            ) |>
  dplyr::ungroup()


pH_curve_avg_subclades_K <- pH_curve_avg_subclades_K +
  geom_vline(data = opt_by_sub,
             aes(xintercept = mean_pHopt),
             linetype = "dashed", linewidth = 0.4, colour = "black",
             inherit.aes = FALSE)
print(pH_curve_avg_subclades_K)

ggsave(here("plots/pH_curve_TopSubclades_K.png"),
       width = 120,
       height = 40,
       units = "mm",
       dpi = 300)


saveRDS(pH_curve_avg_subclades_K, here("output/rds/pH_curve_avg_subclades_K.rds"))
saveRDS(out_k$plot, here("output/rds/out_k_plot.rds"))




ggpubr::ggarrange(pH_curve_avg_subclades_K, out_k$plot,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = c("B", "D"),
                  hjust = 0.05,
                  font.label = list(size = 10))

ggsave(
      filename = here("plots/combined_avg_params_k.png"),
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300,
      device = "png",
      bg = "white"
    )


ggpubr::ggarrange(pH_curve_avg_subclades_K, pca_plot_k,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = "AUTO",
                  hjust = 0.05,
                  font.label = list(size = 10))

ggsave(
      filename = here("plots/combined_avg_params_pca_k.png"),
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )


pH_curve_avgOnly_subclades_K_mean <- pH_curve_avgOnly_subclades_K +
  # vertical SE band
  geom_rect(
    data = opt_by_sub %>% filter(subclade %in% c("IC","IIG","IIIA","IVB")),
    aes(xmin = mean_pHopt - pHopt_se, xmax = mean_pHopt + pHopt_se, ymin = -Inf, ymax = Inf, fill = subclade),
    inherit.aes = FALSE,
    alpha = 0.12,
    show.legend = FALSE
  ) +
  geom_vline(data = opt_by_sub |> dplyr::filter(subclade %in% c("IC","IIG","IIIA","IVB")),
             aes(xintercept = mean_pHopt, colour = subclade),
             linetype = "dashed", linewidth = 0.75)
print(pH_curve_avgOnly_subclades_K_mean)

ggsave(here("plots/pH_curve_TopSubclades_K_avg_only.png"),
       width = 120,
       height = 40,
       units = "mm",
       dpi = 300)

ggpubr::ggarrange(pH_curve_avgOnly_subclades_K_mean, pca_plot_k,
                  ncol = 1,align = "hv",
                  heights = c(1, 1), labels = c("B","D"),
                  hjust = 0.05,
                  font.label = list(size = 10))

ggsave(
      filename = here("plots/combined_avgCurveOnly_pca_k.png"),
      width    = 60,
      height   = 100,
      units    = "mm",
      dpi      = 300,
      device = "png"
    )


```
