---
title: "Thermal Performance Curves"
author: "Alberto Barron"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_width: 11
    fig_height: 7
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

#### Description

This script will consolidate the processed data from the initial GC
analysis and merge it into a larger file. Then it calculates mean values
across replicates of the same strain and temperature. Finally it
calculates mean growth rate and sd for each combination and plots the
values against temperature.

```{r setup, include=FALSE}
library(here)
knitr::opts_chunk$set(echo = F, cache = F)

# Output directories (gitignored; created on a fresh clone)
dir.create(here("output/temp/plots/temp_curves"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/temp/tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/rds"), recursive = TRUE, showWarnings = FALSE)

knitr::opts_knit$set(root.dir = here("output/temp"))
options(scipen=10000, digits = 12)
set.seed(seed = 711)
```

```{r general settings, message=FALSE, warning=FALSE, results='hold', echo=FALSE}

library(growthcurver)
library(lubridate)
library(ggplot2)
library(dplyr)
library(reshape2)
library(purrr)
library(tuple)
library(tidyr)
library(broom)

apatheme=theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family='Arial'))


#"#c5add8" "#c90076" "#8da8f3" "#fff2cc" "#ff0000" "#00ff00" "#c90077"
```

```{r set working directory and load and merge, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
# Working directory handled by here(); GC_values inputs live in data/temp
file_list <- list.files(path = here("data/temp"), pattern = "^GC_values_Temp_assay\\..*\\.txt$", full.names = TRUE)


for (file in file_list){
      
  # if the merged dataset doesn't exist, create it
  if (!exists("all.data")){
    all.data <- read.table(file, header=TRUE, sep="\t")
  }
  
  # if the merged dataset does exist, append to it
  if (exists("all.data")){
    temp_dataset <-read.table(file, header=TRUE, sep="\t")
    all.data<-unique(plyr::rbind.fill(all.data, temp_dataset))
    rm(temp_dataset)
  }
}
```

### Number of assays and ANOVAS for each Temperature and strain across date of the assay

```{r number of assays summary and ANOVAS, message=FALSE, warning=FALSE, echo=FALSE, fig.height=18, fig.width=12}

## Let's get the metadata info for each strain. For that we need to read in the master metadata table.
master_md <- read.table(here("data/shared/master_md_all_sequenced_Isolates_with_extendedMD.txt"), header = T, sep = "\t")


clade_info <- merge(all.data, master_md, by.x = "StrainID", by.y = "Name", all.x = T) %>% select(StrainID, clade, subclade) %>% distinct()

all.data <-merge(all.data, clade_info, by = "StrainID", all.x = T)

all.data$strain.date<-c(paste(all.data$StrainID, all.data$date, sep = "."))
all.data$strain.rep<-c(paste(all.data$StrainID, all.data$replicate, sep = "."))
all.data$strain.date.rep<-c(paste(all.data$StrainID, all.data$date, all.data$replicate, sep = "."))

#now let's match the clade colors to the metadata
clade_colors <- read.table(here("data/shared/clade_colors.txt"), header = T, sep = "\t", comment.char = "")
all.data$clade_color <- clade_colors$color[match(all.data$clade, clade_colors$variable)]
all.data$subclade_color <- clade_colors$color[match(all.data$subclade, clade_colors$variable)]

na_strains <- all.data %>% filter(is.na(clade)) %>% distinct(StrainID)


## create strings to color by:
## clade:
clade_col <- as.character(all.data$clade_color)
names(clade_col) <- as.character(all.data$clade)
# strain by clade color
strain_clade_col <- as.character(all.data$clade_color)
names(strain_clade_col) <- as.character(all.data$StrainID)

## subclade:
subclade_col <- as.character(all.data$subclade_color)
names(subclade_col) <- as.character(all.data$subclade)
# strain by clade color
strain_subclade_col <- as.character(all.data$subclade_color)
names(strain_subclade_col) <- as.character(all.data$StrainID)


## now let's find all the Strains that have been tested on different assays (dates)

strain_counts <- all.data %>%
  group_by(StrainID) %>%
  filter(n_distinct(date) >= 2) %>%
  summarise(dates = list(unique(date)))

```

We will first analyse those strain that have been tested in different
assays (dates) to asses whether the results are consistent.

```{r assesing reproducibility, message=FALSE, warning=FALSE, echo=FALSE, fig.height=8, fig.width=12}
tested_2orMore <- all.data %>%
  filter(StrainID %in% strain_counts$StrainID) %>%
  select(StrainID, replicate, r, date, Temp)

# Step 2: Count the number of unique dates for each strain
strain_date_counts <- tested_2orMore %>%
  group_by(StrainID) %>%
  summarise(unique_dates = n_distinct(date))

# Step 3: Filter strains with 3 or more unique dates
strains_with_3_or_more_dates <- strain_date_counts %>%
  filter(unique_dates >= 3)


# let's leave out from the analysis this one for now and move on with the others
tested_2orMore <- tested_2orMore %>%
  filter(!StrainID %in% c(na_strains,"33G4B08"))


# Function to calculate correlations for a given subset of data
calculate_correlations <- function(df, dates_to_compare) {
  # Pivot data to wide format with r values for each date as separate columns
  pivoted_data <- df %>%
    filter(date %in% dates_to_compare) %>%
    pivot_wider(names_from = date, values_from = r, names_prefix = "Date_", values_fn = list)
  
  # Ensure all necessary columns are present and fill missing values with NA
  necessary_columns <- c("StrainID", "replicate", "Temp", paste0("Date_", dates_to_compare))
  for (col in necessary_columns) {
    if (!col %in% colnames(pivoted_data)) {
      pivoted_data[[col]] <- NA
    }
  }
  
  # Unnest list columns to individual rows
  pivoted_data <- pivoted_data %>%
    unnest(cols = c(starts_with("Date_")))
  
  # Convert columns to numeric to ensure compatibility with cor()
  pivoted_data <- pivoted_data %>%
    mutate(across(starts_with("Date_"), ~ as.numeric(unlist(.))))
  
  # Calculate correlation for each subset
  calc_correlation <- function(df) {
    numeric_cols <- df[, -c(1:3)]
    if (ncol(numeric_cols) < 2) {
      return(NA)
    }
    # Check for at least two non-NA values in each column
    if (any(colSums(!is.na(numeric_cols)) < 2)) {
      return(NA)
    }
    # Calculate the correlation
    cor_result <- cor(numeric_cols[[1]], numeric_cols[[2]], use = "pairwise.complete.obs")
    return(cor_result)
  }
  
  # Apply the function to calculate correlations for the subset
  correlation_value <- calc_correlation(pivoted_data)
  
  return(tibble(StrainID = df$StrainID[1],
                correlation = correlation_value))
}

# Filter data to include only the dates we want to compare
dates_to_compare_1 <- c("2024-01-14", "2024-02-20")
dates_to_compare_2 <- c("2024-03-14", "2024-04-15")

# Process data by StrainID and replicate
correlation_results_1 <- tested_2orMore %>%
  group_by(StrainID, #replicate
           ) %>%
  arrange(Temp) %>%
  do(calculate_correlations(., dates_to_compare_1)) %>%
  ungroup()

# Process data by StrainID and replicate
correlation_results_2 <- tested_2orMore %>%
  group_by(StrainID, #replicate
           ) %>%
  arrange(Temp) %>%
  do(calculate_correlations(., dates_to_compare_2)) %>%
  ungroup()

# Combine the two sets of correlation results
combined_cor_results <- bind_rows(
  correlation_results_1 %>% mutate(date_set = "2024-01-14 to 2024-02-20"),
  correlation_results_2 %>% mutate(date_set = "2024-03-14 to 2024-04-15")
)


# now let's do the same for carrying capacity (K)

tested_2orMore <- all.data %>%
  filter(StrainID %in% strain_counts$StrainID) %>%
  select(StrainID, replicate, k, date, Temp)

# let's leave out from the analysis this one for now and move on with the others
tested_2orMore <- tested_2orMore %>%
  filter(!StrainID %in% c(na_strains,"33G4B08"))

# Function to calculate correlations for a given subset of data
calculate_correlations_k <- function(df, dates_to_compare) {
  # Pivot data to wide format with r values for each date as separate columns
  pivoted_data <- df %>%
    filter(date %in% dates_to_compare) %>%
    pivot_wider(names_from = date, values_from = k, names_prefix = "Date_", values_fn = list)
  
  # Ensure all necessary columns are present and fill missing values with NA
  necessary_columns <- c("StrainID", "replicate", "Temp", paste0("Date_", dates_to_compare))
  for (col in necessary_columns) {
    if (!col %in% colnames(pivoted_data)) {
      pivoted_data[[col]] <- NA
    }
  }
  
  # Unnest list columns to individual rows
  pivoted_data <- pivoted_data %>%
    unnest(cols = c(starts_with("Date_")))
  
  # Convert columns to numeric to ensure compatibility with cor()
  pivoted_data <- pivoted_data %>%
    mutate(across(starts_with("Date_"), ~ as.numeric(unlist(.))))
  
  # Calculate correlation for each subset
  calc_correlation <- function(df) {
    numeric_cols <- df[, -c(1:3)]
    if (ncol(numeric_cols) < 2) {
      return(NA)
    }
    # Check for at least two non-NA values in each column
    if (any(colSums(!is.na(numeric_cols)) < 2)) {
      return(NA)
    }
    # Calculate the correlation
    cor_result <- cor(numeric_cols[[1]], numeric_cols[[2]], use = "pairwise.complete.obs")
    return(cor_result)
  }
  
  # Apply the function to calculate correlations for the subset
  correlation_value <- calc_correlation(pivoted_data)
  
  return(tibble(StrainID = df$StrainID[1],
                correlation = correlation_value))
}

# Process data by StrainID and replicate
correlation_results_1_k <- tested_2orMore %>%
  group_by(StrainID, #replicate
           ) %>%
  arrange(Temp) %>%
  do(calculate_correlations_k(., dates_to_compare_1)) %>%
  ungroup()

# Process data by StrainID and replicate
correlation_results_2_k <- tested_2orMore %>%
  group_by(StrainID, #replicate
           ) %>%
  arrange(Temp) %>%
  do(calculate_correlations_k(., dates_to_compare_2)) %>%
  ungroup()

# Combine the two sets of correlation results
combined_cor_results_k <- bind_rows(
  correlation_results_1_k %>% mutate(date_set = "2024-01-14 to 2024-02-20"),
  correlation_results_2_k %>% mutate(date_set = "2024-03-14 to 2024-04-15")
)


combined_cor_results_all <- bind_rows(
  combined_cor_results %>% mutate(parameter = "r"),
  combined_cor_results_k %>% mutate(parameter = "k")
)



# Create the density plot
cor_plot <- ggplot(combined_cor_results_all, aes(x = correlation, fill = date_set)) +
  geom_density(alpha = 0.5) +
  facet_grid(~parameter, scales = "free", space = "free_y") +
  labs(title = "Density Plot of Correlation Values for carrying capacity (K) and growth rate (r)",
       subtitle = "Strains tested across different dates",
       x = "Correlation",
       y = "Density",
       fill = "Date Set") +
  apatheme +
  theme(legend.position = "bottom")
print(cor_plot)

ggsave("plots/correlation_densityplots.png",
       width = 25,
       height = 20,
       units = "cm",
       dpi = 600)


```

### Fitting thermal performance curves

Now let's plot the curves using a fitted model. for this we will use the
package "rTPC"

```{r fitting thermnal performance curves with rTPC, message=FALSE, warning=FALSE, echo=FALSE, fig.height=18, fig.width=12}
# load packages
library(rTPC)
library(nls.multstart)
library(MuMIn)

## first we need to do a clean up the data, remove unecessary columns and probably calculating averages for growthg rate (r)

d <- all.data %>%
  filter(!StrainID %in% c(na_strains,"33G4B08")) |>
  select(StrainID, r, k, Temp, date, replicate, strain.date) %>%
  group_by(StrainID, date, Temp) %>% 
  summarise(r_mean = mean(r),
            r_sd   = sd(r),
            k_mean = mean(k),
            k_sd   = sd(k), .groups = "drop") %>% 
  rename(temp = Temp, rate = r_mean, strain = StrainID)

### Set aside a few other strain for Alejandra, we will use both growth rate (r) and carrying capacity (K)
### NOTE: "Grass-7", "MMLR14-010" and "Grass-32", not tested because we only tested clade 1 Alex's strains.


d4A <-  as.data.frame(all.data %>%
  filter(
        StrainID %in% c("Grass-54",
                        "Grass-40",
                        "Grass-44",
                        "Grass-45",
                        "Grass-57",
                        "Grass-31",
                        "Grass-42",
                        "Grass-43",
                        "Mmlr14-002",
                        "Grass-36") |
          subclade %in% c("IA", "IB")) %>%
  select(StrainID, r, k, Temp, date, replicate, subclade) %>%
  group_by(StrainID, date, Temp) %>% 
  summarise(r_mean = mean(r),
            k_mean = mean(k),
            r_sd = sd(r),
            k_sd = sd(k), .groups = "drop") %>% 
    left_join(clade_info |> select(StrainID, subclade),
              by = "StrainID") |> 
  rename(temp = Temp, strain = StrainID, rate = r_mean))

write.table(d4A,
            file = "tables/observed_values_selected_strains.txt",
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")


```

```{r fit fit curves function, message=FALSE, warning=FALSE, echo=FALSE}
# when scaling up our code to fit hundreds of models, its nice to have a progress bar
# edit nls_multstart to allow for a progress bar
# Function to fit models with progress bar
nls_multstart_progress <- function(formula,
                                   data = parent.frame(),
                                   iter,
                                   start_lower, 
                                   start_upper,
                                   supp_errors = c("Y", "N"),
                                   convergence_count = 100, 
                                   control,
                                   modelweights, ...){
  if(!is.null(pb)){
    pb$tick()
  }
  tryCatch({
    nls_multstart(formula = formula,
                  data = data,
                  iter = iter,
                  start_lower = start_lower, 
                  start_upper = start_upper,
                  supp_errors = supp_errors,
                  convergence_count = convergence_count, 
                  control = control,
                  modelweights = modelweights, ...)
  }, error = function(e) {
    return(NULL)
  })
}
```

We will fit several different models and evaluate the model performance
through AICc scores **and** select the overall best fitting model

-   Sharpe–Schoolfield (full) (sharpeschoolfull_1981)
-   Sharpe–Schoolfield (high) (sharpeschoolhigh_1981)
-   Johnson–Lewin (1946) (johnsonlewin_1946)
-   Pawar (2018) (pawar_2018)
-   briereextended_2021() briere2_1999()
-   weibull_1995()

######### Fitting curves for Alejandra

Since we already calculate curves with growth rate (r) now we will just
calculate TPC with carrying capacity (k) or max yield.

```{r r and k curves 4 Alejandra, message=FALSE, warning=FALSE, echo=FALSE}


# Filter out strains with too many zero values
d_filtered <- d4A %>%
  group_by(strain, date) %>%
  filter(sum(k_mean == 0) / n() < 0.5) %>% # Remove strains with 25% or more zero values
  ungroup() |>
  mutate(strain.date = paste(strain, date, sep = ".")) |>
  select(-c(strain, date, r_sd, rate, k_sd)) |>
  relocate(strain.date, .before = 1) |>
  rename(rate = k_mean)

# Filter out strains with too many zero values and keep them for later analysis
d_filtered_out <- d4A %>%
  group_by(strain, date) %>%
  filter(sum(k_mean == 0) / n() > 0.5) %>% # Remove strains with 25% or more zero values
  ungroup()

length(unique(d_filtered$strain.date))



# Set up progress bar
number_of_models <- 5
number_of_curves <- length(unique(d_filtered$strain.date))
pb <- progress::progress_bar$new(total = number_of_curves * number_of_models,
                                 clear = T,
                                 format = "[:bar] :percent :elapsedfull")

set.seed(711)
# Fit models

d_fits <- d_filtered %>%
  nest(data = c(temp, rate)) %>%
  mutate(
    sharpeschoolhigh = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      
      nls_multstart_progress(
        rate ~ sharpeschoolhigh_1981(temp = temp, r_tref, e, eh, th, tref = 20),
        data = .x,
        iter = c(3, 3, 3, 3),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    sharpeschoolfull = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      nls_multstart_progress(
        rate ~ sharpeschoolfull_1981(temp = temp, r_tref, e, el, tl, eh, th, tref = 20),
        data = .x,
        iter = rep(3, 6),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    pawar = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, model_name = "pawar_2018")
      ll <- get_lower_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      ul <- get_upper_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      nls_multstart_progress(
        rate ~ pawar_2018(temp = temp, r_tref, e, eh, th, tref = 20),
        data        = .x,
        iter        = rep(3, length(sv)),
        start_lower = sv - 5,
        start_upper = sv + 5,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
    johnsonlewin   = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "johnsonlewin_1946")
      ll <- get_lower_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      ul <- get_upper_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      nls_multstart_progress(
        rate ~ johnsonlewin_1946(temp = temp, r0, e ,eh, topt),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 2,
        start_upper = sv + 2,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
      weibull        = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "weibull_1995")
      ll <- get_lower_lims(.x$temp, .x$rate, "weibull_1995")
      ul <- get_upper_lims(.x$temp, .x$rate, "weibull_1995")
      nls_multstart_progress(
        rate ~ weibull_1995(temp = temp, a, topt, b, c),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 1,
        start_upper = sv + 1,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    })
  )



error_rows <- d_fits %>%
  filter(
    map_lgl(sharpeschoolhigh, is.null) |
    map_lgl(sharpeschoolfull,   is.null) |
    map_lgl(pawar,          is.null) |
    map_lgl(johnsonlewin,          is.null) |
    map_lgl(weibull,          is.null) #|
    #map_lgl(briere2,          is.null)
    )

successful_fits <- d_fits %>%
  filter(
    !map_lgl(sharpeschoolhigh, is.null) |
    !map_lgl(sharpeschoolfull,   is.null) |
    !map_lgl(pawar,          is.null) |
    !map_lgl(johnsonlewin,          is.null) |
    !map_lgl(weibull,          is.null)
    )

# Print summary of errors
print("Rows with fitting errors:")
print(error_rows)



# 1. Pivot to long: one row per strain × model
d_long <- successful_fits %>%
  pivot_longer(
    cols      = sharpeschoolhigh:weibull,     # or your five model columns
    names_to  = "model",
    values_to = "fit"
  ) %>%
  filter(!map_lgl(fit, is.null))             # drop failed fits

# 2. Compute AIC, AICc, BIC, sigma, df.resid
d_stats <- d_long %>%
  mutate(
    # glance() gives AIC, BIC, sigma, df.resid, etc.
    glance = map(fit, broom::glance),
    AICc   = map_dbl(fit, MuMIn::AICc)
  ) %>%
  unnest(glance) %>%
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  select(strain, strain.date, model, AIC, AICc, BIC, sigma, df.residual)

# 3. Within each strain, compute ΔAICc and Akaike weight
d_compare <- d_stats %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()

# 4a. “Best” model per strain
best_by_strain <- d_compare %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, AICc, weight)

overall <- d_compare %>%
  group_by(model) %>%
  summarise(
    prop_best   = mean(delta == 0),
    mean_weight = mean(weight)
  ) %>%
  arrange(desc(prop_best), desc(mean_weight))

print(overall)

# 4b. (Optional) Full model‐averaged prediction or parameters
#    You can join `weight` back onto your predictions or 
#    onto rTPC::calc_params() output and take weighted averages.

# === Examples of exploring the results ===

# How many strains picked each model?
best_by_strain %>% count(best_model)


# 1. See exactly which rows are infinite
d_compare %>% 
  filter(!is.finite(AICc)) %>% 
  count(model) %>% 
  arrange(desc(n))

# Peek at means
d_compare %>% 
  filter(is.finite(AICc)) |>
  group_by(model) |>
  summarise(mean = mean(AICc, na.rm = T),
            sd   = sd(AICc, na.rm = T),
            n    = n())

hist(d_compare$delta[d_compare$model == "sharpeschoolfull"], breaks = 100)

# Visualize ΔAICc distributions
ggplot(d_compare, aes(model, delta)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = expression(Delta~AIC[c]), x = NULL)

hist(d_compare$AICc[d_compare$model == "sharpeschoolfull"])

# Visualize AICc distributions
ggplot(d_compare, aes(model, AICc)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "AICc", x = NULL)

# -----------------------------------------------------------------------------
# 2. Compute fit statistics & Akaike weights
# -----------------------------------------------------------------------------
d_stats <- d_stats %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()


best_by_strain <- d_stats %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, best_AICc = AICc)

# -----------------------------------------------------------------------------
# 3. Extract TPC trait parameters and do model-averaging vs best-model
# -----------------------------------------------------------------------------

params_long <- d_long %>%
  mutate(params = map(fit, calc_params, .progress = T)) %>%
  select(strain.date, model, params) %>%
  unnest(params) |>
  pivot_longer(
    cols      = c(rmax, topt, ctmin, ctmax, e, eh, q10,
                  thermal_safety_margin, thermal_tolerance,
                  breadth, skewness),
    names_to  = "term",
    values_to = "estimate"
  )

params_long_k <- params_long |>
  filter(model == "sharpeschoolfull") |>
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # remove the original strain.date
  ) |>
  filter(date != "2024-01-14") |>
  rename(estimate_K = estimate)
  
  


# 4a) build per-strain prediction grids
pred_grids <- d_long %>%
  select(strain.date, data) %>%
  distinct() %>%
  mutate(
    newdata = map(data, ~ tibble(
      temp = seq(min(.x$temp), max(.x$temp), length.out = 50)
    ))
  ) %>%
  select(-data)


# 1. Get in‐sample predictions for each (strain × model)
pred_obs <- d_long %>% 
  # d_long has columns: strain, model, data (the original tibble), fit
  mutate(
    obs = map2(fit, data, ~ augment(.x, newdata = .y), .progress = T)
  ) %>%
  select(strain.date, model, obs) %>%
  unnest(obs) %>%
  # columns now include: strain, model, temp, rate (observed), .fitted
  left_join(
    d_stats %>% select(strain.date, model, weight),
    by = c("strain.date","model")
  )

# 4b) augment each fit
d_preds <- d_long %>%
  left_join(pred_grids, by = "strain.date") %>%
  mutate(preds = map2(fit, newdata, ~ augment(.x, newdata = .y), .progress = T)) %>%
  select(strain.date, model, preds) %>%
  unnest(preds)

# 4e) extract best‐model prediction
best_preds <- d_preds %>%
  filter(model == "sharpeschoolfull") %>%
  select(strain.date, temp, best_fitted = .fitted) |>
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  # (optionally) turn your date string into a real Date
  mutate(
    date = as.Date(date)
  )

## join clade info and original observations

preds <- merge(best_preds, clade_info,
               by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(!is.na(subclade))

d_filtered_wmd <- merge(d_filtered |> separate(
                                col    = strain.date,
                                into   = c("strain","date"),
                                sep    = "\\.",      # split on dot
                                extra  = "merge",    # if there are more dots, merge them into the last column
                                remove = FALSE),     # keep the original strain.date if you still need it
                        clade_info |> select(-subclade),
                        by.x = "strain", by.y = "StrainID", all.x = T
                        ) |>
  left_join(d |> select(strain,
                        date,
                        temp,
                        k_sd),
            by = c("strain", "date", "temp")) |>
  filter(!is.na(subclade))


## plot curves
temp_curve_k <- preds %>%
  ggplot(aes(x = temp,
                    color = strain,
                    group = strain.date,
                    y = best_fitted)) +
  scale_color_manual(values = strain_subclade_col, guide = "none") +
  geom_line(alpha = 1, linewidth = 2) +
  geom_point(data = d_filtered_wmd,
             aes(temp, rate, color = strain, shape = date), alpha = 0.75, size = 5) +
  scale_shape_discrete(name = "Date") +
  #geom_smooth(aes(fill = strain)) +
  #scale_fill_manual(values = strain_clade_col) +
  geom_errorbar(data = d_filtered_wmd,
                  aes(x = temp, ymin = rate-k_sd, ymax = rate+k_sd, color = strain),
                  width = .4, inherit.aes = F, alpha = 0.65) +
  scale_x_continuous(breaks = sort(unique(all.data$Temp))) +
  ggh4x::facet_nested_wrap(.~strain, ncol = 4) +
  ylim(0, 1) +
  #labs(title = paste("Subclade", subcld, "(page", i, "of", length(strain_chunks), ")")
  #     #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))
  #     ) +
  ylab(substitute(paste("Max. yield (", italic("K"),") (OD)"))) +
  xlab("Temperature (ºC)") +
  apatheme +
  theme(
        panel.grid.major.x = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.minor   = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        legend.position    = "bottom",
        legend.box         = "horizontal",
        axis.text.y = element_text(size = 20),
        strip.background = element_blank(),
        text = element_text(size = 35),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 24 ),
        plot.subtitle = element_text(color = "black", size = 16))
    
  print(temp_curve_k)
  
  ggsave(
      filename = "plots/temp_curves/temp_curves_K_strains4A.png",
      plot     = temp_curve_k,
      width    = 600,
      height   = 350,
      units    = "mm",
      dpi      = 300
    )
  

```

```{r fit temperature curves, message=FALSE, warning=FALSE, echo=FALSE}
# Filter out strains with too many zero values
d_filtered <- d %>%
  group_by(strain, date) %>%
  filter(sum(rate == 0) / n() < 0.5) %>% # Remove strains with 25% or more zero values
  ungroup() |>
  mutate(strain.date = paste(strain, date, sep = ".")) |>
  select(-c(#strain, date,
            r_sd, k_mean, k_sd)) |>
  relocate(strain.date, .before = 1)

# Filter out strains with too many zero values and keep them for later analysis
d_filtered_out <- d %>%
  group_by(strain, date) %>%
  filter(sum(rate == 0) / n() > 0.5) %>% # Remove strains with 25% or more zero values
  ungroup()

length(unique(d_filtered$strain.date))

# Vector of all strains
all_strains      <- unique(d$strain)
# Vector of those that passed
kept_strains     <- unique(d_filtered$strain)
# Those that were dropped
removed_strains2 <- setdiff(all_strains, kept_strains)


# Set up progress bar
number_of_models <- 5
number_of_curves <- length(unique(d_filtered$strain.date))
pb <- progress::progress_bar$new(total = number_of_curves * number_of_models,
                                 clear = T,
                                 format = "[:bar] :percent :elapsedfull")

set.seed(711)
# Fit models

d_fits <- d_filtered %>%
  nest(data = c(temp, rate)) %>%
  mutate(
    sharpeschoolhigh = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      
      nls_multstart_progress(
        rate ~ sharpeschoolhigh_1981(temp = temp, r_tref, e, eh, th, tref = 20),
        data = .x,
        iter = c(3, 3, 3, 3),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    sharpeschoolfull = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      nls_multstart_progress(
        rate ~ sharpeschoolfull_1981(temp = temp, r_tref, e, el, tl, eh, th, tref = 20),
        data = .x,
        iter = rep(3, 6),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    pawar = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, model_name = "pawar_2018")
      ll <- get_lower_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      ul <- get_upper_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      nls_multstart_progress(
        rate ~ pawar_2018(temp = temp, r_tref, e, eh, th, tref = 20),
        data        = .x,
        iter        = rep(3, length(sv)),
        start_lower = sv - 5,
        start_upper = sv + 5,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
    johnsonlewin   = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "johnsonlewin_1946")
      ll <- get_lower_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      ul <- get_upper_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      nls_multstart_progress(
        rate ~ johnsonlewin_1946(temp = temp, r0, e ,eh, topt),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 2,
        start_upper = sv + 2,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
      weibull        = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "weibull_1995")
      ll <- get_lower_lims(.x$temp, .x$rate, "weibull_1995")
      ul <- get_upper_lims(.x$temp, .x$rate, "weibull_1995")
      nls_multstart_progress(
        rate ~ weibull_1995(temp = temp, a, topt, b, c),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 1,
        start_upper = sv + 1,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    })
  )



error_rows <- d_fits %>%
  filter(
    map_lgl(sharpeschoolhigh, is.null) |
    map_lgl(sharpeschoolfull,   is.null) |
    map_lgl(pawar,          is.null) |
    map_lgl(johnsonlewin,          is.null) |
    map_lgl(weibull,          is.null) #|
    #map_lgl(briere2,          is.null)
    )

successful_fits <- d_fits %>%
  filter(
    !map_lgl(sharpeschoolhigh, is.null) |
    !map_lgl(sharpeschoolfull,   is.null) |
    !map_lgl(pawar,          is.null) |
    !map_lgl(johnsonlewin,          is.null) |
    !map_lgl(weibull,          is.null)
    )

# Print summary of errors
print("Rows with fitting errors:")
print(error_rows)
```

```{r fitted temperature curves analysis, message=FALSE, warning=FALSE, echo=FALSE}
# 1. Pivot to long: one row per strain × model
d_long <- successful_fits %>%
  pivot_longer(
    cols      = sharpeschoolhigh:weibull,     # or your five model columns
    names_to  = "model",
    values_to = "fit"
  ) %>%
  filter(!map_lgl(fit, is.null))             # drop failed fits

# 2. Compute AIC, AICc, BIC, sigma, df.resid
d_stats <- d_long %>%
  mutate(
    # glance() gives AIC, BIC, sigma, df.resid, etc.
    glance = map(fit, broom::glance),
    AICc   = map_dbl(fit, MuMIn::AICc)
  ) %>%
  unnest(glance) %>%
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  select(strain, strain.date, model, AIC, AICc, BIC, sigma, df.residual)

# 3. Within each strain, compute ΔAICc and Akaike weight
d_compare <- d_stats %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()

# 4a. “Best” model per strain
best_by_strain <- d_compare %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, AICc, weight)

overall <- d_compare %>%
  group_by(model) %>%
  summarise(
    prop_best   = mean(delta == 0),
    mean_weight = mean(weight)
  ) %>%
  arrange(desc(prop_best), desc(mean_weight))

print(overall)

# 4b. (Optional) Full model‐averaged prediction or parameters
#    You can join `weight` back onto your predictions or 
#    onto rTPC::calc_params() output and take weighted averages.

# === Examples of exploring the results ===

# How many strains picked each model?
best_by_strain %>%
  count(best_model) |>
  mutate(percent = (n/sum(n))*100)


# 1. See exactly which rows are infinite
d_compare %>% 
  filter(!is.finite(AICc)) %>% 
  count(model) %>% 
  arrange(desc(n))

# Peek at means
d_compare %>% 
  filter(is.finite(AICc)) |>
  group_by(model) |>
  summarise(mean = mean(AICc, na.rm = T),
            sd   = sd(AICc, na.rm = T),
            n    = n())

hist(d_compare$delta[d_compare$model == "sharpeschoolfull"], breaks = 100)

# Visualize ΔAICc distributions
ggplot(d_compare, aes(model, delta)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = expression(Delta~AIC[c]), x = NULL)

hist(d_compare$AICc[d_compare$model == "sharpeschoolfull"])

# Visualize AICc distributions
ggplot(d_compare, aes(model, AICc)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "AICc", x = NULL)
```

After testing the different models "Sharpescholfull" is the overall best
model so we will use that to fit all curves.

#### Calculate curve parameters

```{r model-avg-params-and-preds 0, message=FALSE, warning=FALSE, echo=FALSE}
# -----------------------------------------------------------------------------
# 2. Compute fit statistics & Akaike weights
# -----------------------------------------------------------------------------
d_stats <- d_stats %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()


best_by_strain <- d_stats %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, best_AICc = AICc)

# -----------------------------------------------------------------------------
# 3. Extract TPC trait parameters and do model-averaging vs best-model
# -----------------------------------------------------------------------------

params_long <- d_long %>%
  mutate(params = map(fit, calc_params, .progress = T)) %>%
  select(strain.date, model, params) %>%
  unnest(params) |>
  pivot_longer(
    cols      = c(rmax, topt, ctmin, ctmax, e, eh, q10,
                  thermal_safety_margin, thermal_tolerance,
                  breadth, skewness),
    names_to  = "term",
    values_to = "estimate"
  )
# → columns: strain, model, term, estimate

### subset parameters for Alejandra

params_long_4A <- params_long |>
  filter(model == "sharpeschoolfull") |>
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # remove the original strain.date
  ) |>
  filter(date != "2024-01-14" &
           strain %in% c("Grass-54",
                        "Grass-40",
                        "Grass-44",
                        "Grass-45",
                        "Grass-57",
                        "Grass-31",
                        "Grass-42",
                        "Grass-43",
                        "Mmlr14-002",
                        "Grass-36")) |>
  rename(estimate_r = estimate)


### merge r and K parameters into a single table for Alejandra

merged_params4A <- params_long_4A |>
  left_join(params_long_k, by = c("strain", "date", "model", "term"))

write.table(merged_params4A,
            file = "tables/curve_params_selected_strains.txt",
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")


avg_params <- params_long %>%
  left_join(
    d_stats %>% select(strain, strain.date, model, weight),
    by = c("strain.date", "model")
  ) %>%
  group_by(strain.date, term) %>%
  summarise(
    avg_estimate = sum(estimate * weight, na.rm = TRUE),
    .groups = "drop"
  )

best_params <- params_long %>%
  inner_join(
    best_by_strain %>% select(strain.date, strain, best_model),
    by = "strain.date"
  ) %>%
  filter(model == best_model) %>%
  select(strain.date, term, best_estimate = estimate)

# 4. Join for comparison
compare_params <- avg_params %>%
  inner_join(best_params, by = c("strain.date", "term")) %>%
  mutate(diff = avg_estimate - best_estimate)

# -----------------------------------------------------------------------------
# 4. Generate predictions on a fine grid and model-average vs best-model
# -----------------------------------------------------------------------------
# 4a) build per-strain prediction grids
pred_grids <- d_long %>%
  select(strain.date, data) %>%
  distinct() %>%
  mutate(
    newdata = map(data, ~ tibble(
      temp = seq(min(.x$temp), max(.x$temp), length.out = 50)
    ))
  ) %>%
  select(-data)
```

```{r calculate predictions and compare, warning=FALSE, message=FALSE, echo=FALSE}

# 1. Get in‐sample predictions for each (strain × model)
pred_obs <- d_long %>% 
  # d_long has columns: strain, model, data (the original tibble), fit
  mutate(
    obs = map2(fit, data, ~ augment(.x, newdata = .y), .progress = T)
  ) %>%
  select(strain.date, model, obs) %>%
  unnest(obs) %>%
  # columns now include: strain, model, temp, rate (observed), .fitted
  left_join(
    d_stats %>% select(strain.date, model, weight),
    by = c("strain.date","model")
  )

# 2. Compute the weighted‐average prediction at each (strain, temp)
avg_pred_obs <- pred_obs %>%
  group_by(strain.date, temp, rate) %>%
  summarise(
    avg_pred = sum(.fitted * weight),
    .groups = "drop"
  )

# 3. Extract the sharpeschoolhigh predictions
sh_pred_obs <- pred_obs %>%
  filter(model == "sharpeschoolfull") %>%
  select(strain.date, temp, rate, sh_pred = .fitted)

pw_pred_obs <- pred_obs %>%
  filter(model == "sharpeschoolhigh") %>%
  select(strain.date, temp, rate, pw_pred = .fitted)

# 4. Join them together and compute RMSE
rmse_cmp <- pw_pred_obs %>%
  inner_join(sh_pred_obs, by = c("strain.date","temp","rate")) %>%
  group_by(strain.date) %>%
  summarise(
    rmse_pw = sqrt(mean((pw_pred - rate)^2)),
    rmse_sh  = sqrt(mean((sh_pred  - rate)^2))
  ) %>%
  ungroup()

# 5. Inspect or plot
print(rmse_cmp)

# e.g. plot rmse_avg vs rmse_sh
ggplot(rmse_cmp, aes(rmse_sh, rmse_pw)) +
  geom_abline(lty = 2) +
  geom_point() +
  labs(
    x = "RMSE: Sharpe–Schoolfield Full",
    y = "RMSE: Sharpe–Schoolfield High",
    title = "Predictive RMSE by Strain"
  ) +
  theme_minimal()


# 4. Join them together and compute RMSE
rmse_cmp <- avg_pred_obs %>%
  inner_join(sh_pred_obs, by = c("strain.date","temp","rate")) %>%
  group_by(strain.date) %>%
  summarise(
    rmse_avg = sqrt(mean((avg_pred - rate)^2)),
    rmse_sh  = sqrt(mean((sh_pred  - rate)^2))
  ) %>%
  ungroup()

# 5. Inspect or plot
print(rmse_cmp)

# e.g. plot rmse_avg vs rmse_sh
ggplot(rmse_cmp, aes(rmse_sh, rmse_avg)) +
  geom_abline(lty = 2) +
  geom_point() +
  labs(
    x = "RMSE: Sharpe–Schoolfield",
    y = "RMSE: Model-averaged",
    title = "Predictive RMSE by Strain"
  ) +
  theme_minimal()
```

```{r model-avg-params-and-preds, message=FALSE, warning=FALSE, echo=FALSE}
# 4b) augment each fit
d_preds <- d_long %>%
  left_join(pred_grids, by = "strain.date") %>%
  mutate(preds = map2(fit, newdata, ~ augment(.x, newdata = .y), .progress = T)) %>%
  select(strain.date, model, preds) %>%
  unnest(preds)

# 4c) join weights and best‐model info
d_preds <- d_preds %>%
  left_join(d_stats %>% select(strain, strain.date, model, weight), by = c("strain.date","model")) %>%
  left_join(best_by_strain, by = "strain.date")

# 4e) extract best‐model prediction
best_preds <- d_preds %>%
  filter(model == "sharpeschoolfull") %>%
  select(strain.date, temp, best_fitted = .fitted) |>
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  # (optionally) turn your date string into a real Date
  mutate(
    date = as.Date(date)
  )

# -----------------------------------------------------------------------------
# Now you have:
# - compare_params: per‐strain, per‐trait avg vs best
# - avg_preds & best_preds: per‐strain fitted curves
# -----------------------------------------------------------------------------
```

### Plot predictions and observations

```{r preds, message=FALSE, warning=FALSE, echo=FALSE}

preds <- merge(best_preds, clade_info,
               by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(!is.na(subclade))

d_filtered_wmd <- merge(d_filtered |> separate(
                                col    = strain.date,
                                into   = c("strain","date"),
                                sep    = "\\.",      # split on dot
                                extra  = "merge",    # if there are more dots, merge them into the last column
                                remove = FALSE),     # keep the original strain.date if you still need it
                        clade_info,
                        by.x = "strain", by.y = "StrainID", all.x = T
                        ) |>
  left_join(d |> select(strain,
                        date,
                        temp,
                        r_sd),
            by = c("strain", "date", "temp")) |>
  filter(!is.na(subclade))

```

We can then plot the predictions of each curve using **ggplot2**.

```{r plot fitted thermal curves, fig.height = 12, fig.width=8, message=FALSE, warning=FALSE, echo=FALSE}


plot_temp_curves <- function(
  preds,
  observed = observed,
  all_data,
  strain_cols,
  subclades      = NULL,
  facet_by_clade = TRUE,
  nrow           = 1,
  ylim_max       = 1,
  width_mm       = 600,
  height_mm      = 100,
  units          = "mm",
  dpi            = 300,
  filename       = NULL
) {
  # filter by subclade if requested
  if(!is.null(subclades)){
    preds     <- preds     %>% filter(subclade %in% subclades)
    observed  <- observed  %>% filter(subclade %in% subclades)
  }

  temps <- sort(unique(all.data$Temp))
  
  temp_curve <- preds %>%
    filter(date != "2024-01-14") |> 
    ggplot(aes(x = temp,
                      color = strain,
                      group = strain.date,
                      y = best_fitted)) +
    scale_color_manual(values = strain_subclade_col) +
    geom_line(alpha = 0.65, linewidth = 0.75) +
    #geom_smooth(aes(fill = strain)) +
    #scale_fill_manual(values = strain_clade_col) +
    geom_point(data = observed |> filter(date != "2024-01-14"),
               aes(temp, rate, color = strain), alpha = 0.6) +
    #geom_errorbar(data = observed |> filter(date != "2024-01-14"),
    #              aes(x = temp, ymin = rate-r_sd, ymax = rate+r_sd, color = strain),
    #              width = .2, inherit.aes = F, alpha = 0.65) +
    scale_x_continuous(breaks = temps) +
    ylim(0, 1) +
    #labs(title = "Temperature performance",
         #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))) +
    ylab(substitute(paste("Growth rate (", italic("r"),")"))) +
    xlab("Temperature (ºC)") +
    apatheme +
    theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
          axis.text.y = element_text(size = 20),
          strip.background = element_blank(),
          text = element_text(size = 20),
          #strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 18),
          plot.subtitle = element_text(color = "black", size = 16),
          legend.position = "none")
  
  if(facet_by_clade) {
  temp_curve <- temp_curve +
    ggh4x::facet_nested_wrap(.~clade + subclade, nrow = nrow)
  } else {
    temp_curve <- temp_curve +
      facet_wrap(.~ subclade, nrow = nrow)
  }
  
  print(temp_curve )
  
  if(!is.null(filename)){
  ggsave(
         filename = filename,
         plot     = temp_curve,
         width    = width_mm,
         height   = height_mm,
         units    = units,
         dpi      = dpi
      )
  }
  invisible(temp_curve)
}
 ## plot top subclades
plot_temp_curves(
  preds          = preds,
  observed       = d_filtered_wmd,
  all_data       = all.data,
  strain_cols    = strain_subclade_col,
  subclades      = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),
  facet_by_clade = FALSE,
  filename       = "plots/temp_curves_topSubclades.png",
  nrow           = 1,
  width_mm       = 300,
  height_mm      = 75
)

 ## plot all subclades
plot_temp_curves(
  preds          = preds,
  observed       = d_filtered_wmd,
  all_data       = all.data,
  strain_cols    = strain_subclade_col,
  subclades      = NULL,
  facet_by_clade = T,
  filename       = "plots/temp_curves.png",
  nrow           = 1
)


  correct_order <- c(
  "IA", "IB", "IC",
  "IIA", "IIB", "IIC", "IID", "IIE", "IIF", "IIG", "IIH", "IIJ", "IIK",
  "IIIA", "IIIB", "IIIC", "IIID", "IIIE", "IIIF",
  "IVA", "IVB", "IVC", "IVD", "IVE",
  "VA"
)

preds <- preds |> mutate(subclade = factor(subclade, levels = correct_order))


for (subcld in c("IA", "IB", "IC",
                 "IIE", "IIG", "IIJ",
  "IIIA", "IIIB", "IIID", "IIIE",
  "IIIF",
  "IVA", "IVB", "IVC", "IVE", NA)) {
  
   # get all strains in this subclade
  all_strains <- preds %>%
    filter(subclade == subcld) %>%
    pull(strain) %>%
    unique() %>%
    sort()
  
   # split into chunks of up to 20
  strain_chunks <- split(all_strains, 
                         ceiling(seq_along(all_strains) / 20))
  
  # loop over each chunk (page)
  for (i in seq_along(strain_chunks)) {
    this_strains <- strain_chunks[[i]]
    
    temp_curve <- preds %>%
      filter(subclade == subcld, strain %in% this_strains) |> 
  ggplot(aes(x = temp,
                    color = strain,
                    group = strain.date,
                    y = best_fitted)) +
  scale_color_manual(values = strain_subclade_col, guide = "none") +
  geom_line(alpha = 1, linewidth = 2) +
  geom_point(data = d_filtered_wmd %>% filter(subclade == subcld,
                                              strain %in% this_strains),
             aes(temp, rate, color = strain, shape = date), alpha = 0.75, size = 5) +
  scale_shape_discrete(name = "Date") +
  #geom_smooth(aes(fill = strain)) +
  #scale_fill_manual(values = strain_clade_col) +
  geom_errorbar(data = d_filtered_wmd %>% filter(subclade == subcld,
                                              strain %in% this_strains),
                  aes(x = temp, ymin = rate-r_sd, ymax = rate+r_sd, color = strain),
                  width = .4, inherit.aes = F, alpha = 0.65) +
  scale_x_continuous(breaks = sort(unique(all.data$Temp))) +
  ggh4x::facet_nested_wrap(.~strain, ncol = 4) +
  ylim(0, 1) +
  labs(title = paste("Subclade", subcld, "(page", i, "of", length(strain_chunks), ")")
       #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))
       ) +
  ylab(substitute(paste("Growth rate (", italic("r"),")"))) +
  xlab("Temperature (ºC)") +
  apatheme +
  theme(
        panel.grid.major.x = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.minor   = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        legend.position    = "bottom",
        legend.box         = "horizontal",
        axis.text.y = element_text(size = 20),
        strip.background = element_blank(),
        text = element_text(size = 35),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 24 ),
        plot.subtitle = element_text(color = "black", size = 16))
    
  print(temp_curve)
  
  # now save it
    fn <- sprintf("plots/temp_curves/temp_curves_%s_page%02d.png", subcld, i)
    ggsave(
      filename = fn,
      plot     = temp_curve,
      width    = 600,
      height   = 350,
      units    = "mm",
      dpi      = 300
    )
  }
}


temp_curve_s4A <- preds %>%
      filter(strain %in% c("Grass-54",
                        "Grass-40",
                        "Grass-44",
                        "Grass-45",
                        "Grass-57",
                        "Grass-31",
                        "Grass-42",
                        "Grass-43",
                        "Mmlr14-002",
                        "Grass-36")) |> 
  ggplot(aes(x = temp,
                    color = strain,
                    group = strain.date,
                    y = best_fitted)) +
  scale_color_manual(values = strain_subclade_col, guide = "none") +
  geom_line(alpha = 1, linewidth = 2) +
  geom_point(data = d_filtered_wmd %>% filter(strain %in% c("Grass-54",
                        "Grass-40",
                        "Grass-44",
                        "Grass-45",
                        "Grass-57",
                        "Grass-31",
                        "Grass-42",
                        "Grass-43",
                        "Mmlr14-002",
                        "Grass-36")),
             aes(temp, rate, color = strain, shape = date), alpha = 0.75, size = 5) +
  scale_shape_discrete(name = "Date") +
  #geom_smooth(aes(fill = strain)) +
  #scale_fill_manual(values = strain_clade_col) +
  geom_errorbar(data = d_filtered_wmd %>% filter(strain %in% c("Grass-54",
                        "Grass-40",
                        "Grass-44",
                        "Grass-45",
                        "Grass-57",
                        "Grass-31",
                        "Grass-42",
                        "Grass-43",
                        "Mmlr14-002",
                        "Grass-36")),
                  aes(x = temp, ymin = rate-r_sd, ymax = rate+r_sd, color = strain),
                  width = .4, inherit.aes = F, alpha = 0.65) +
  scale_x_continuous(breaks = sort(unique(all.data$Temp))) +
  ggh4x::facet_nested_wrap(.~strain, ncol = 4) +
  ylim(0, 1) +
  #labs(title = paste("Subclade", subcld, "(page", i, "of", length(strain_chunks), ")")
       #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))
   #    ) +
  ylab(substitute(paste("Growth rate (", italic("r"),")"))) +
  xlab("Temperature (ºC)") +
  apatheme +
  theme(
        panel.grid.major.x = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.minor   = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        legend.position    = "bottom",
        legend.box         = "horizontal",
        axis.text.y = element_text(size = 20),
        strip.background = element_blank(),
        text = element_text(size = 35),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 24 ),
        plot.subtitle = element_text(color = "black", size = 16))
    
  print(temp_curve_s4A)
  
  # now save it

    ggsave(
      filename = "plots/temp_curves/temp_curves_r_strains4A.png",
      plot     = temp_curve_s4A,
      width    = 600,
      height   = 350,
      units    = "mm",
      dpi      = 300
    )


    
    
    
preds_avg <- preds |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND"))) |>
  group_by(subclade, temp) |>
  summarise(avg_r = mean(best_fitted))





### first recode subclade IB
temp_curve_avg <- preds %>%
    filter(!(
      (strain == "32G3B09" & date == "2024-03-14") |
      (strain == "39G1A09" & date == "2024-03-14") |
      (strain == "39G1A10" & date == "2024-03-14") |
      (strain == "30G3A10" & date == "2024-04-15") |
      (strain == "33G2A09" & date == "2024-04-15") |
      (strain == "35G4A05" & date == "2024-04-15") |
      (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
      (strain == "28G1A03" & date == "2024-03-14") |
      (strain == "28G1B05" & date == "2024-03-14") |
      (strain == "28G4A04" & date == "2024-03-14") |
      (strain == "35G1A07" & date == "2024-03-14") |
      (strain == "39G1A08" & date == "2024-03-14") |
      (strain == "30D10" & date == "2024-03-14") |
      (strain == "23G1B05" & date == "2024-03-14") |
      (subclade == "IIIE" & date == "2024-04-15") |
      (strain == "30G1B03" & date == "2024-03-14") |
      (strain == "23G1B05" & date == "2024-03-14") |
      (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
      (strain == "36G2A07" & date == "2024-03-14") |
      (strain == "36G3B02" & date == "2024-03-14") |
      (strain == "36G3B05" & date == "2024-03-14") |
      (strain == "37G4A03" & date == "2024-03-14") |
      (clade == "UND"))) |>
      filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) |>
      ggplot(aes(x = temp,
                        color = strain,
                        group = strain.date,
                        y = best_fitted)) +
      geom_line(alpha = 0.75, linewidth = 0.15) +
      scale_color_manual(values = strain_clade_col, guide = "none") +
      # 2) tell ggplot we want a *new* colour scale
      ggnewscale::new_scale_color() +
      geom_smooth(data = preds_avg |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")),
                 aes(temp, avg_r, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = T) + 
      scale_colour_manual(name   = "Subclade",
                          values = subclade_col) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_col) +
      #scale_x_continuous(breaks = seq(4,46, 6)) +
      scale_x_continuous(breaks = c(4, 25, 46)) +
      ylim(0, 1.1) +
      facet_wrap(.~ subclade, nrow = 1) +
      ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
      xlab("Temperature (ºC)") +
      apatheme +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 7),
            axis.text.y = element_text(size = 7),
            strip.background = element_blank(),
            text = element_text(size = 8),
            #strip.text.x = element_blank(),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
            #legend.position = c(0.95,0.85),
            #legend.justification = c(1, 1),         # anchor legend’s top-right corner
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(temp_curve_avg)


ggsave(
      filename = "plots/temp_curves/temp_curves_avg_topSubclades.png",
      width    = 140,
      height   = 80,
      units    = "mm",
      dpi      = 300
    )



temp_curve_avg_only <- preds_avg |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IVB")) |> 
      ggplot() +
      geom_smooth(
                 aes(temp, avg_r, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = F) +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_col) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_col) +
      #scale_x_continuous(breaks = seq(4,46, 6)) +
      scale_x_continuous(breaks = seq(4, 46, 6)) +
      ylim(0, 0.5) +
      ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
      xlab("Temperature (ºC)") +
      apatheme +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 6),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            #strip.text.x = element_blank(),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
            #legend.position = c(0.95,0.85),
            #legend.justification = c(1, 1),         # anchor legend’s top-right corner
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(temp_curve_avg_only)


ggsave(
      filename = "plots/temp_curves/temp_curves_avgOnly_topSubclades.png",
      width    = 80,
      height   = 80,
      units    = "mm",
      dpi      = 300
    )
  
```

###### plot for grant

Here we will plot all clade I curves in the back ground and highlight
only one curve per subclade, we also need to recode subclade IB into IBa
and IBb, so we have 4 subclades in clade I

**- Subclade IBa** - 28G4A10 - 35D02 - 35D04

**- Subclade IBb** - Pine-20 - 32G4B12

```{r quick figure for a grant, include=FALSE, echo=FALSE, warning=FALSE}
# we need to update (sub)clade info and colors
 clade_info_upd <- clade_info |>
  mutate(subclade = case_when(StrainID %in% c("28G4A10", "35D02", "35D04") ~ "IBa",
                              StrainID %in% c("Pine-20", "32G4B12") ~ "IBb",
                              TRUE ~ subclade))

selected_strains <- c("32G4B12", "Mcba15-009", "28G4A10", "35G1B04")

preds_select <- preds |>
  filter(strain %in% selected_strains,
         date != "2024-01-14") |>
  mutate(subclade = case_when(strain == "28G4A10" ~ "IBa",
                              strain == "32G4B12" ~ "IBb",
                              TRUE ~ subclade))



# update color info
subclade_colors <- clade_colors |>
  add_row(
    variable = c("IBa", "IBb"),
    color    = c("#009ACD", "#6EDEFD")
  ) |>
  tibble::deframe()


### first recode subclade IB
 preds %>%
    filter(date != "2024-01-14",
           clade == "I") |>
  select(strain) |>
  distinct() |>
  count()

length(unique(preds$strain[preds$clade == "I"]))



### first recode subclade IB
temp_curve <- preds %>%
    filter(date != "2024-01-14",
           clade == "I") |> 
    ggplot(aes(x = temp,
                      color = strain,
                      group = strain.date,
                      y = best_fitted)) +
    geom_line(alpha = 0.15, linewidth = 0.5) +
    scale_color_manual(values = strain_clade_col, guide = "none") +
    # 2) tell ggplot we want a *new* colour scale
    ggnewscale::new_scale_color() +
    geom_line(data = preds_select,
               aes(temp, best_fitted, color = subclade), alpha = 1, linewidth = 1.2) +
    scale_colour_manual(name   = "Subclade",
                        values = subclade_colors) +
    scale_x_continuous(breaks = seq(4,46, 3)) +
    ylim(0, 1.2) +
    #labs(title = "Temperature performance",
         #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))) +
    #ylab(substitute(paste("Growth rate (", italic("r"),")"))) +
    ylab(expression(paste("Growth rate (", h^{-1}, ")"))) +
    xlab("Temperature (ºC)") +
    apatheme +
    theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
          axis.text.y = element_text(size = 10),
          strip.background = element_blank(),
          text = element_text(size = 12),
          #strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 18),
          plot.subtitle = element_text(color = "black", size = 16),
          legend.position = "bottom")
print(temp_curve)


  ggsave(
      filename = "plots/temp_curves/temp_curves_selected_strains.pdf",
      plot     = temp_curve,
      width    = 120,
      height   = 80,
      units    = "mm",
      dpi      = 300,
      device = cairo_pdf
    )
  
  
## exporting data ##

clade_I_preds <- preds %>%
    filter(date != "2024-01-14",
           clade == "I") |>
  select(-c(clade, subclade)) |>
  left_join(clade_info_upd,
            by = c("strain" = "StrainID")) |>
  mutate(Alex_subclade_classification = case_when(subclade == "IBa" ~ "ID",
                              subclade == "IBb" ~ "IE",
                              subclade == "IC" ~ "IBC",
                              TRUE ~ subclade))
  

write.table(clade_I_preds,
            file = here("output/temp/clade_I_predictions_updated_classification.csv"),
            row.names = F,
            col.names = T,
            quote = F, sep = ",")

clade_I_obs <- d_filtered_wmd %>%
    filter(date != "2024-01-14",
           clade == "I") |>
  select(-c(clade, subclade)) |>
  left_join(clade_info_upd,
            by = c("strain" = "StrainID")) |>
  mutate(Alex_subclade_classification = case_when(subclade == "IBa" ~ "ID",
                              subclade == "IBb" ~ "IE",
                              subclade == "IC" ~ "IBC",
                              TRUE ~ subclade))

write.table(clade_I_obs,
            file = here("output/temp/clade_I_observations_updated_classification.csv"),
            row.names = F,
            col.names = T,
            quote = F, sep = ",")


```

```{r, include=FALSE}
# Function to create a plot for a subset of strains
#plot_subset <- function(data, strains_subset, d_filtered, plot_number, clade) {
#  p <- ggplot(data %>% filter(strain %in% strains_subset)) +
#    geom_line(aes(temp, .fitted, col = model_name)) +
#    geom_point(aes(temp, rate), d_filtered %>% filter(strain %in% strains_subset)) +
#    facet_wrap(~strain, scales = 'free_y', ncol = 5) +
#    theme_bw() +
#    theme(legend.position = 'none') +
#    scale_color_brewer(type = 'qual', palette = 2) +
#    labs(x = 'Temperature (ºC)',
#         y = 'Metabolic rate',
#         title = paste('Thermal performance curves - Plot', plot_number, '- Clade:', clade),
#         subtitle = 'gaussian in green; sharpeschoolhigh in orange')
#  return(p)
#}
#
## Define the number of strains per plot
#strains_per_plot <- 20
## Split data by clade
#d_preds_by_clade <- split(d_preds, d_preds$clade)
#
## Initialize a list to store the plots
#plots <- list()
#plot_number <- 1
#
## Loop over each clade and split further into subsets of strains_per_plot
#for (clade in names(d_preds_by_clade)) {
#  clade_data <- d_preds_by_clade[[clade]]
#  unique_strains <- unique(clade_data$strain)
#  
#  # Split the strains into subsets
#  strain_subsets <- split(unique_strains, ceiling(seq_along(unique_strains) / strains_per_plot))
#  
#  # Generate plots for each subset
#  clade_plots <- map2(strain_subsets, seq_along(strain_subsets), ~{
#    plot_subset(d_preds, .x, d_filtered, plot_number, clade)
#  })
#  
#  # Add the generated plots to the main list
#  plots <- c(plots, clade_plots)
#  plot_number <- plot_number + length(clade_plots)
#}
#
## Display the first plot as an example
##print(plots[[1]])
#
#
##for (plot in plots) {
##  print(plot)
##}

# Save the plots to files
#walk2(plots, seq_along(plots), ~ggsave(filename = paste0("plots/thermal_performance_curves_", .y, ".png"), plot = .x, width = 12, height = 8))
```

#### Thermal performance parameters

The traits of each thermal performance curve can also easily be
calculated.

```{r params , message=FALSE, warning=FALSE, echo=FALSE}
# stack models and calculate extra params
d_params <- pivot_longer(d_fits, names_to = 'model_name', values_to = 'fit',
                         c(sharpeschoolhigh,sharpeschoolfull,pawar)) %>%
  mutate(params = map(fit, calc_params, .progress = T)) %>%
  select(strain.date, model_name, params) %>%
  unnest(params) %>%
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = F       # keep the original strain.date if you still need it
  ) %>%
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

glimpse(d_params)

d_params <- d_params |>
   filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND" & !is.na(clade))))

d_params <- d_params %>% filter(!is.na(clade))


d_params_ssfull <- d_params %>%
  filter(model_name == "sharpeschoolfull") %>%
  select(strain, date, strain.date, rmax, ctmin, ctmax, topt, breadth) %>%
  filter(date != "2024-01-14") |>
  select(-c(date)) %>%
  tibble::column_to_rownames("strain.date")

d_params_ssfull[sapply(d_params_ssfull, is.infinite)] <- NA

pca <-prcomp(na.omit(d_params_ssfull) %>% filter(!strain %in% c("09D09", "23G4B04")) |> select(-strain)
             ,
             scale. = T, center = T)


pca.df <- data.frame(pca$x) %>%
  select(PC1, PC2) |>
  tibble::rownames_to_column("strain.date") |> 
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # keep the original strain.date if you still need it
  ) %>%
    merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T) |>
    filter(!is.na(subclade))

PCAloads<-data.frame(Variables = rownames(pca$rotation), pca$rotation)

pca_plot<-ggplot(pca.df #%>% filter(Row.names != "23G4B04")
                 , aes(PC1, PC2)) +
  geom_point(data = pca.df, aes(color = subclade),  size = 5, alpha = 0.5) +
  #geom_text(data = pca.df, aes(x=PC1, y=PC2-0.15), label=pca.df$Row.names, size=2.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               arrow = arrow(length = unit(0.3, "cm")),
               colour="grey", inherit.aes = FALSE) +
 geom_text(data=PCAloads,
           aes(x=PC1*2.5, y=PC2*2.5, label=Variables),
           size=6,
           colour = "darkblue") +
  xlab("PC1 (49.75%)") +
  ylab("PC2 (39.84%)") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 18))
pca_plot

summary(pca)

  

d_params_ssfull_topSubclades <- d_params %>%
  filter(model_name == "sharpeschoolfull") %>%
  filter(!strain %in% c("09D09", "23G4B04")) |>
  filter(subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  select(date, strain.date, ctmin, ctmax, topt, breadth) %>%
  filter(date != "2024-01-14") |>
  select(-c(date)) %>%
  tibble::column_to_rownames("strain.date")


d_params %>%
  filter(model_name == "sharpeschoolfull") %>%
  filter(!strain %in% c("09D09", "23G4B04")) |>
  filter(subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  filter(date != "2024-01-14") |>
  group_by(subclade) |> 
  summarise(avg.topt = mean(topt, na.rm = T),
            avg.ctmin = mean(ctmin, na.rm = T),
            avg.ctmax = mean(ctmax, na.rm = T),
            avg.breadth = mean(breadth, na.rm = T))





pca_ssfull_topSubclades <- prcomp(na.omit(d_params_ssfull_topSubclades ), scale. = T, center = T)

pca.ssfull_topSubclades.df <- data.frame(pca_ssfull_topSubclades$x) %>%
  select(PC1, PC2) %>% 
  tibble::rownames_to_column("strain.date") |> 
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # keep the original strain.date if you still need it
  ) %>%
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

PCAloads.ssfull_topSubclades <- data.frame(Variables = rownames(pca_ssfull_topSubclades$rotation), 
                                           pca_ssfull_topSubclades$rotation) |>
                                              mutate(Variables = recode(Variables,
                                                                        "ctmin" = "Min. Temp",
                                                                        "topt" = "Opt. Temp",
                                                                        "ctmax" = "Max. Temp",
                                                                        "breadth" = "Breadth"))

pca_plot.ssfull_topSubclades<-ggplot(pca.ssfull_topSubclades.df,
                                     aes(PC1, PC2)) +
  geom_point(data = pca.ssfull_topSubclades.df, aes(color = subclade),  size = 1.5, alpha = 0.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads.ssfull_topSubclades,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               colour="grey", inherit.aes = FALSE) +
  geom_text(data=PCAloads.ssfull_topSubclades,
           aes(x=PC1*4.5, y=PC2*4.5, label=Variables),
           size=2,
           colour = "black",
           alpha = 1,
           fontface = "bold") +
  lims(x = c(-6.5, 6.5),
       y = c(-4.5, 4.5)) +
  xlab(paste0("PC1 (",round(summary(pca_ssfull_topSubclades)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca_ssfull_topSubclades)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 8),
        legend.position = "none",
        legend.position.inside = c(0.9, 0.2),
        legend.background = element_rect(fill = alpha("white", 0), color = NA),
        legend.key.size = unit(0.5, "lines"),
        legend.title = element_text(size = 6,
                                    margin = margin(b = 0, unit = "pt")),
        legend.text = element_text(size = 5),
        legend.key.height = unit(0.4, "lines"),
        legend.key.width = unit(0.7, "lines"),
        legend.spacing.y = unit(0, "cm"))
pca_plot.ssfull_topSubclades

ggsave(
      filename = "plots/pca_plot.ssfull_topSubclades_r.pdf",
      plot     = pca_plot.ssfull_topSubclades,
      width    = 90,
      height   = 90,
      units    = "mm",
      dpi      = 300,
      device = cairo_pdf
    )

#### a quick PERMANOVA to test the effect of subclade membership on curve parameters

D <- dist(pca_ssfull_topSubclades$x, "euclidean")

perm <- vegan::adonis2(D ~ subclade, data = pca.ssfull_topSubclades.df, permutations = 999)

perm


#summary(pca_schol)

# pick a handful of key traits
traits <- c("rmax", "ctmax", "topt", "breadth")

d_params_ssfull %>%
  # merge with clade info
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(clade != "UND") |>
  filter(!is.na(subclade)) |> 
  # raname coulumn row.names to strain
  #rename(strain = Row.names) |>
  # select columns of interest
  select(strain, clade, subclade, all_of(traits)) %>%
  # pivot longer for plotting
  pivot_longer(cols = traits, names_to = "term", values_to = "value") %>%
  # plot 
  ggplot(aes(subclade, value, fill = subclade)) +
    geom_boxplot() +
    scale_fill_manual(values = subclade_col, name="Subclade") +
    ggh4x::facet_nested_wrap(~ term + clade, scales = "free", ncol = 4) +
    ggpubr::stat_compare_means(aes(label = ..p.signif..), method = "kruskal.test") +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "Subclade", y = "Estimate", title = "Trait differences among Subclades")



dotplot_traits_by_subclade <- function(
  params_df,     # your d_params_ssfull
  clade_info,    # your clade_info (with StrainID → clade, subclade)
  traits,        # character vector of trait names, e.g. c("topt","rmax","ctmax")
  subclades = NULL,      # which subclades to plot; NULL = all
  facet_by_clade = T,    # if FALSE, facet only by term
  filename  = NULL,      # if non-NULL, path to save the PNG
  width_mm  = 180,
  height_mm = 100,
  dpi       = 300
){
  # 1. join metadata and pivot longer
  df_long <- params_df %>%
    left_join(clade_info, by = c("strain" = "StrainID")) %>%
    filter(clade != "UND", !is.na(subclade)) %>%
    { if(!is.null(subclades)) filter(., subclade %in% subclades) else . } %>%
    pivot_longer(
      cols      = all_of(traits),
      names_to  = "term",
      values_to = "value"
    ) |>
    mutate(forcats::fct_rev(forcats::fct_inorder(subclade)))
  
  # 2. build the plot
  plt <- ggplot(df_long,
                aes(x = value,
                    y = subclade,
                    colour = subclade)) +
    geom_jitter(height = 0.15, alpha = 0.35) +
    stat_summary(fun = mean, geom = "point",
                 size = 3, shape = 21, fill = "white") +
    stat_summary(fun.data = mean_cl_boot,
                 geom     = "errorbarh",
                 height   = 0.2) +
    scale_y_discrete(limits = rev) +
    scale_colour_manual(values = subclade_col, guide = FALSE) +
    labs(x = "Estimate", y = "Subclade") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.y      = element_text(face = "bold"),
      strip.text       = element_text(size = 16),
      panel.grid.major = element_line(colour = "grey90"),
      panel.grid.minor = element_blank(),
      legend.position  = "none"
    )
  
  # 3. Add faceting
  if (facet_by_clade) {
      plt <- plt + facet_grid(clade ~ term, scales = "free", space = "free_y") 
  } else {
    plt <- plt + facet_grid(~ term, scales = "free", space = "free_y") 
  }
  
  
  
  # 4. print, and optionally save
  print(plt)
  if(!is.null(filename)){
    ggsave(filename   = filename,
           plot       = plt,
           width      = width_mm,
           height     = height_mm,
           units      = "mm",
           dpi        = dpi)
  }
  
  return(plt)
}




# Plot all subclades:
Subcl_dot_plot <- dotplot_traits_by_subclade(
  params_df = d_params_ssfull,
  clade_info = clade_info,
  traits = traits#,
  #filename = "~/SyncFolder/CA_curto/phys_assays/temp/processed_data/plots/curve_params_dotplot.png"
)

# Plot just two subclades, and save:
topSubcl_dot_plot <- dotplot_traits_by_subclade(
  params_df  = d_params_ssfull,
  clade_info = clade_info,
  traits     = traits,
  subclades  = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),
  facet_by_clade = F,
  filename   = here("output/temp/plots/curve_params_dotplot_abundantSubclades.png")
)

# map subclade -> clade (unique map)
sub2clade <- clade_info %>%
  distinct(subclade, clade) %>%
  filter(!is.na(subclade), clade != "UND")


d_params_schol_long <- d_params_ssfull %>%
  # merge with clade info
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(clade != "UND", !is.na(subclade)) |>
  select(subclade, all_of(traits)) %>%
  pivot_longer(cols = traits, names_to = "term", values_to = "value")

d_params_ssfull <- d_params_ssfull |>
  left_join(clade_info, by = c("strain" = "StrainID"))



posthocs <- d_params_schol_long %>%
  # only keep the subclades you care about
  
  #filter(subclade %in% c("IC","IIG","IIIA","IIIE","IVB","IVE")) %>%
  filter(!subclade %in% c("IIIB", "IVA")) %>%
  group_by(term) %>%
  do({
    dat <- .

    # 1) must have ≥2 subclades and each with ≥2 obs
    sub_counts <- dat %>% count(subclade) 
    if (n_distinct(dat$subclade) < 2 ||
        any(sub_counts$n < 2)) {
      return(tibble()) 
    }

    # 2) normality?
    if (nrow(dat) >= 3) {
      sw_p <- dat |> rstatix::shapiro_test(value) |> pull(p)
    } else {
      sw_p <- NA_real_
    }

    # 3) homogeneity?
    lev_p <- rstatix::levene_test(dat, value ~ subclade)$p

    # 4) choose test
    if (!is.na(sw_p) && sw_p > 0.05 && lev_p > 0.05) {
      # ANOVA + Tukey
      omni <- rstatix::anova_test(dat, value ~ subclade)
      ph   <- rstatix::tukey_hsd(dat, value ~ subclade) %>%
                mutate(method = "Tukey HSD")
      test_type <- "ANOVA"
    } else {
      # Kruskal–Wallis + Dunn
      omni <- rstatix::kruskal_test(dat, value ~ subclade)
      ph   <- rstatix::dunn_test(dat, value ~ subclade, p.adjust.method = "BH") %>%
                mutate(method = "Dunn (BH)")
      test_type <- "KW"
    }

    # 5) build omnibus row
    x_max <- max(dat$value, na.rm = TRUE)
    omnibus_df <- tibble(
      term         = unique(dat$term),
      group1       = NA_character_,
      group2       = NA_character_,
      p.adj        = omni$p,
      p.adj.signif = as.character(
                                  symnum(omni$p,
                                         corr = F,
                                         cutpoints = c(0, 0.001, 0.01, 0.05, 1),
                                         symbols   = c("***","**","*","ns"))
                                  ),
      x.position   = x_max * 1.08,
      method       = test_type
      )
    
    # 6) position pairwise rows just below
    ph_df <- ph %>%
      mutate(
        term         = unique(dat$term),
        p.adj.signif = as.character(
                          symnum(p.adj,
                                 corr = F,
                                 cutpoints = c(0, 0.001, 0.01, 0.05, 1),
                                 symbols   = c("***","**","*","ns"))
                        ),
        x.position = x_max * 1.05)

    bind_rows(omnibus_df, ph_df)
  }) %>%
  ungroup()

# Inspect your posthocs table
posthocs

# 2. Compute max per panel so we can place omnibus labels
panel_ranges <- d_params_schol_long %>%
  filter(term %in% traits) |>
  group_by(term) %>%
  summarise(xmax = max(value, na.rm=TRUE), .groups="drop") %>%
  mutate(xpos = xmax * 1.02)     # 2% beyond the data

# 4. Join in the xpos
posthocs <- posthocs %>%
  left_join(panel_ranges, by = "term")


# 1) Pull out just your omnibus rows
posthocs_omni <- posthocs %>%
  filter(term %in% traits) |>
  filter(is.na(group1)) |>
  mutate(clade = "IV")


# 2) For each term, build a named vector of p’s and feed to multcompLetters
letter_df <- posthocs %>%
  filter(term %in% traits) |>
  filter(!is.na(group1)) %>% 
  select(term, group1, group2, p.adj) |> 
  group_by(term) %>%
  summarise(
    # build a named vector c("A-B"=p1, "A-C"=p2, ...)
    pv = list( setNames(p.adj, paste(group1, group2, sep = "-")) ),
    .groups = "drop"
  ) %>%
  mutate(
    # run multcompLetters on each vector
    letters = map(pv, ~ multcompView::multcompLetters(.x, threshold = 0.05)$Letters)
  ) %>%
  select(-pv) %>%
  unnest_longer(letters, indices_to = "subclade", values_to = "letter") |>
  left_join(panel_ranges, by = "term") |> left_join(sub2clade, by = "subclade")

# 2) Add the text directly
topSubcl_dot_plot +
  geom_text(
    data        = posthocs_omni,
    aes(x        = x.position * 0.68,
        y        = 0.25,
        label    = p.adj.signif),
    inherit.aes = FALSE,
    size = 10,
    hjust = 0      # left‐justify so it sits just outside
  ) +
  geom_text(
    data        = letter_df,
    aes(x     = xpos * 0.1,
        y     = subclade,
        label = letter),
    inherit.aes = FALSE,
    size        = 5,
    hjust       = 0
  )


ggsave(
      filename = here("output/temp/plots/curve_params_dotplot_topSubclades.png"),
      width    = 180,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )



# 2) Add the text directly
Subcl_dot_plot +
  geom_text(
    data        = posthocs_omni,
    aes(x        = x.position * 0.68,
        y        = 1,
        label    = p.adj.signif),
    inherit.aes = FALSE,
    size = 8,
    hjust = 0      # left‐justify so it sits just outside
  ) +
  geom_text(
    data        = letter_df,
    aes(x     = xpos * 0.1,
        y     = subclade,
        label = letter),
    inherit.aes = FALSE,
    size        = 3,
    hjust       = 0
  )


ggsave(
      filename = here("output/temp/plots/curve_params_dotplot_Subclades.png"),
      width    = 180,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )


d_params_ssfull <- d_params_ssfull |>
  select(-c(clade, subclade))


plot_traits_subclade <- function(
  params_df,              # data with columns: strain, <traits...>
  clade_info,             # columns: StrainID, clade, subclade
  traits,                 # e.g., c("topt","rmax","ctmax","breadth","ctmin")
  k_traits     = NULL,    # optionally for traits calculated with LOESS model
  subclades    = NULL,    # keep only these subclades (keeps order if given)
  exclude_subs = NULL,    # drop these subclades
  facet_by_clade = TRUE,  # facet with clade rows; else only by term
  subclade_levels = NULL, # order of subclades
  flip = FALSE,           # horizontal (FALSE) vs vertical (TRUE) layout
  term_labels = NULL,
  show_annotations = c("both","omnibus","letters","none"),
  omnibus_in_clade = NULL, # which clade row should show the omnibus star; NULL -> first
  subclade_col = NULL,    # named vector of subclade colours (names must match levels)
  filename = NULL, width_mm = 180, height_mm = 100, dpi = 300
){
  show_annotations <- match.arg(show_annotations)

  # ---------- 1) Long data (and factor ordering) ----------
  df_long <- params_df %>%
    dplyr::left_join(clade_info, by = c("strain" = "StrainID")) %>%
    dplyr::filter(!is.na(subclade), clade != "UND") %>%
    { if (!is.null(subclades)) dplyr::filter(., subclade %in% subclades) else . } %>%
    { if (!is.null(exclude_subs)) dplyr::filter(., !subclade %in% exclude_subs) else . } %>%
    tidyr::pivot_longer(dplyr::all_of(traits), names_to = "term", values_to = "value") %>%
    dplyr::mutate(
      # order subclades: keep user order if supplied; else by first appearance, then flip so first is top
      subclade = if (!is.null(subclade_levels)) forcats::fct_relevel(subclade, subclade_levels)
                 else if (!is.null(subclades)) forcats::fct_relevel(subclade, rev(subclades))
                 else forcats::fct_rev(forcats::fct_inorder(subclade)),
      clade    = factor(clade)
    )

  y_levels <- levels(df_long$subclade)          # to align annotations
  clade_levels <- levels(df_long$clade)
  if (is.null(omnibus_in_clade)) omnibus_in_clade <- clade_levels[1]
  
  ## define parameter order
  
  if (!is.null(traits)) {
    term_levels <- c("topt","ctmax","rmax", "breadth")
    df_long <- df_long |>
    mutate(term = factor(term, levels = term_levels))
  } else if (!is.null(k_traits)) {
      term_levels <- c("topt","ctmax","rmax", "breadth", k_traits)
  df_long <- df_long |>
    mutate(term = factor(term, levels = term_levels))
  
  }
  
  facet_labeller <- if (!is.null(term_labels)) {
        labeller(term = as_labeller(term_labels))
      } else {
        label_value  # default
      }
  
  
  # ---------- 2) Base plot ----------
  if (!flip) {
    plt <- ggplot2::ggplot(df_long, ggplot2::aes(x = value, y = subclade, colour = subclade)) +
      ggplot2::geom_jitter(height = 0.15, alpha = 0.35) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 3, shape = 21, fill = "white") +
      ggplot2::stat_summary(fun.data = mean_cl_boot, geom = "errorbarh", height = 0.2) +
      ggplot2::labs(x = "Temperature (ºC)", y = "Subclade")
    plt <- plt + if (facet_by_clade)
      ggplot2::facet_grid(clade ~ term, scales = "free", space = "free_y",
                          labeller = facet_labeller)
    else
      ggplot2::facet_grid(~ term, scales = "free", space = "free_y",
                          labeller = facet_labeller)
  } else {
    plt <- ggplot2::ggplot(df_long, ggplot2::aes(y = value, x = subclade, colour = subclade)) +
      ggplot2::geom_jitter(width = 0.15, alpha = 0.35) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 3, shape = 21, fill = "white") +
      ggplot2::stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
      ggplot2::labs(y = "Temperature (ºC)", x = "Subclade") +
      scale_y_continuous(position = "right")
    plt <- plt + if (facet_by_clade)
      ggplot2::facet_grid(term ~ clade, scales = "free", space = "free_x", switch = "y",
                          labeller = facet_labeller)
    else
      ggplot2::facet_grid(term ~ ., scales = "free", space = "free_x", switch = "y",
                          labeller = facet_labeller)
  }

  if (!is.null(subclade_col)) {
    plt <- plt + ggplot2::scale_colour_manual(values = subclade_col, guide = "none")
  } else {
    plt <- plt + ggplot2::guides(colour = "none")
  }

  plt <- plt +
    ggplot2::theme_minimal(base_size = 8) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(face = "bold"),
      strip.text  = ggplot2::element_text(size = 8),
      panel.grid.minor = ggplot2::element_blank()
    )

  # ---------- 3) Stats (omnibus + posthocs) ----------
  # Helper: run per (clade, term) if faceting by clade; else per term (across clades)
  #stats_grouping <- if (facet_by_clade) c("clade","term") else c("term")
  stats_grouping <- "term"

  # map subclade -> clade (ensures clade present if needed)
  sub2clade <- clade_info %>%
    dplyr::distinct(subclade, clade) %>%
    dplyr::filter(!is.na(subclade), clade != "UND")

  stat_input <- df_long %>%
    dplyr::select(clade, subclade, term, value) #%>%
    #{ if (facet_by_clade) dplyr::left_join(., sub2clade, by = "subclade") else . }

  posthocs <- stat_input %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(stats_grouping))) %>%
    dplyr::do({
      dat <- .
      sub_counts <- dplyr::count(dat, subclade)
      if (dplyr::n_distinct(dat$subclade) < 2 || any(sub_counts$n < 2)) return(tibble::tibble())

      sw_p  <- tryCatch(rstatix::shapiro_test(dat, value) %>% dplyr::pull(p), error = function(e) NA_real_)
      lev_p <- tryCatch(rstatix::levene_test(dat, value ~ subclade) %>% dplyr::pull(p), error = function(e) NA_real_)

      if (!is.na(sw_p) && sw_p > 0.05 && !is.na(lev_p) && lev_p > 0.05) {
        omni <- rstatix::anova_test(dat, value ~ subclade)
        ph   <- rstatix::tukey_hsd(dat, value ~ subclade) %>% dplyr::mutate(method = "Tukey HSD")
        test_type <- "ANOVA"
      } else {
        omni <- rstatix::kruskal_test(dat, value ~ subclade)
        ph   <- rstatix::dunn_test(dat, value ~ subclade, p.adjust.method = "BH") %>% dplyr::mutate(method = "Dunn (BH)")
        test_type <- "KW"
      }

      vmax <- max(dat$value, na.rm = TRUE)
      omni_row <- tibble::tibble(
        group1 = NA_character_, group2 = NA_character_,
        p.adj  = as.numeric(omni$p)[1],
        p.adj.signif = as.character(stats::symnum(as.numeric(omni$p)[1],
                                                  corr = FALSE,
                                                  cutpoints = c(0,.001,.01,.05,1),
                                                  symbols   = c("***","**","*","n.s."))),
        xpos = vmax * 1.02, step = "omnibus", method = test_type
      )

      ph_rows <- ph %>%
        dplyr::transmute( 
          group1, group2, p.adj,
          p.adj.signif = as.character(stats::symnum(p.adj, corr = FALSE,
                                                    cutpoints = c(0,.001,.01,.05,1),
                                                    symbols   = c("***","**","*","n.s."))),
          xpos = vmax * 0.08, step = "posthoc", method
        )

      dplyr::bind_rows(omni_row, ph_rows)
    }) %>%
    dplyr::ungroup()

  # Letters per panel
  letter_df <- posthocs %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(stats_grouping))) %>%
    dplyr::summarise(
      pv = list(stats::setNames(p.adj, paste(group1, group2, sep = "-"))),
      .groups = "drop"
    ) %>%
    dplyr::mutate(letters = purrr::map(pv, ~ multcompView::multcompLetters(.x, threshold = 0.05)$Letters)) %>%
    dplyr::select(-pv) %>%
    tidyr::unnest_longer(letters, indices_to = "subclade", values_to = "letter") %>%
    dplyr::mutate(subclade = factor(subclade, levels = y_levels)) |>
    left_join(sub2clade, by = "subclade") |>
    drop_na()

  # One omnibus per term (place in chosen clade if facetted)
  if (facet_by_clade) {
    posthocs_omni_one <- posthocs %>%
      dplyr::filter(step == "omnibus") %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(p.adj.signif = dplyr::first(p.adj.signif),
                       xpos         = dplyr::first(xpos),
                       .groups = "drop") %>%
      dplyr::mutate(clade = factor(omnibus_in_clade, levels = clade_levels))
  } else {
    posthocs_omni_one <- posthocs %>%
      dplyr::filter(step == "omnibus") %>%
      dplyr::group_by(term) %>%
      dplyr::summarise(p.adj.signif = dplyr::first(p.adj.signif),
                       xpos         = dplyr::first(xpos),
                       .groups = "drop")
  }

  
  # compute per-term y positions for annotations (works with facet rows = term)
  rng <- df_long %>%
    dplyr::group_by(term) %>%
    dplyr::summarise(
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      span = max - min,
      # put letters a bit above the top of the data for that term
      lab = max + 0.06 * span,
      # position for the omnibus star
      star = max + 0.12 * span,
      .groups = "drop"
    )
  
  # join positions to your annotation data
  letters_plt <- letter_df %>% dplyr::left_join(rng, by = "term")
  omni_plt    <- posthocs_omni_one %>%
    #dplyr::distinct(term, p.adj.signif) %>%
    dplyr::left_join(rng, by = "term")
  
  # ---------- 4) Add annotations ----------
  if (show_annotations %in% c("both","omnibus")) {
    if (!flip) {
      plt <- plt + ggplot2::geom_text(
        data = omni_plt,
        ggplot2::aes(x = star*0.85, y = -Inf, label = p.adj.signif),
        inherit.aes = FALSE, vjust = 1.2, size = 3
      )
    } else {
      plt <- plt + ggplot2::geom_text(
        data = omni_plt,
        ggplot2::aes(y = star*0.85, x = -Inf, label = p.adj.signif),
        inherit.aes = FALSE, hjust = -0.1, size = 3
      )
    }
  }

  if (show_annotations %in% c("both","letters")) {
    if (!flip) {
      plt <- plt + ggplot2::geom_text(
        data = letters_plt,
        ggplot2::aes(x = lab*0.85, y = subclade, label = letter),
        inherit.aes = FALSE, hjust = 0, size = 2.5
      )
    } else {
      plt <- plt + ggplot2::geom_text(
        data = letters_plt,
        ggplot2::aes(y = lab*0.85, x = subclade, label = letter),
        inherit.aes = FALSE, vjust = 0, hjust = 2.5, size = 2.5
      )
    }
  }

  # ---------- 5) Save & return ---------- 
  print(plt)
  if (!is.null(filename)) {
    ggplot2::ggsave(filename, plot = plt,
                    width = width_mm, height = height_mm,
                    units = "mm", dpi = dpi, bg = "white")
  }

  invisible(list(
    plot = plt,
    data_long = df_long,
    posthocs = posthocs,
    letters  = letter_df,
    omnibus_one = posthocs_omni_one
  ))
}

# helper for a safe default (returns b if a is NULL)
`%||%` <- function(a, b) if (is.null(a)) b else a





out <- plot_traits_subclade(
  params_df = d_params_ssfull,
  clade_info = clade_info,
  traits = c("breadth","ctmax","topt"),
  subclades = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  #subclades = NULL,                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    topt  = "Opt. Temp.",
    ctmax  = "Max. Temp.",
    breadth = "Breadth"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("output/temp/plots/curve_params_dotplot_Subclades_flip.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)

View(out$posthocs)

# per-subclade optimal pH (where avg_k is maximal)
opt_by_sub <- d_params_ssfull |>
  left_join(clade_info, by = c("strain" = "StrainID")) |>
  dplyr::filter(subclade %in% c("IC","IIG","IIIA","IIIE","IVB","IVE")) |>
  dplyr::group_by(subclade) |>
  summarise(n = sum(!is.na(topt)),
            mean_topt = mean(topt),
            med_topt = median(topt),
            topt_se = sd(topt)/sqrt(n),
            topt_sd = sd(topt)
            ) |>
  #dplyr::slice_max(topt, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() #|>
  #dplyr::transmute(subclade, topt = topt)


temp_curve_avg <- temp_curve_avg +
  geom_vline(data = opt_by_sub,
             aes(xintercept = mean_topt),
             linetype = "dashed", linewidth = 0.4, colour = "black",
             inherit.aes = FALSE)

ggpubr::ggarrange(temp_curve_avg, out$plot,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = c("A", "C"),
                  hjust = 0.05, font.label = list(size = 10))

ggsave(
      filename = "plots/temp_curves/combined_avg_params.png"
      ,
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300,
      device = "png",
      bg = "white"
    )


pH_curve_avg_subclades_K <- readRDS(here("output/rds/pH_curve_avg_subclades_K.rds"))
out_k_plot <- readRDS(here("output/rds/out_k_plot.rds"))

ggpubr::ggarrange(temp_curve_avg, pca_plot.ssfull_topSubclades,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = "AUTO",
                  hjust = 0.05, font.label = list(size = 10))

ggsave(
      filename = "plots/temp_curves/combined_avg_params_r_pca.png"
      ,
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )


ggpubr::ggarrange(
  temp_curve_avg, pH_curve_avg_subclades_K,
  out$plot,       out_k_plot,
  ncol = 2, nrow = 2,
  heights = c(1, 1.5),
  labels = c("A", "B", "C", "D"),
  align = "hv",
  hjust = 0.05, font.label = list(size = 10)
)


ggsave(
      filename = here("output/temp/plots/temp_pH_curves_and_dotplot_2.png"),
      width    = 230,
      height   = 120,
      units    = "mm",
      dpi      = 300,
      bg       = "white"
    )




    temp_curve_avg_only_mean <- temp_curve_avg_only +
  # vertical SE band
  geom_rect(
    data = opt_by_sub %>% filter(subclade %in% c("IC","IIG","IIIA","IVB")),
    aes(xmin = mean_topt - topt_se, xmax = mean_topt + topt_se, ymin = -Inf, ymax = Inf, fill = subclade),
    inherit.aes = FALSE,
    alpha = 0.12,
    show.legend = FALSE
  ) +
  geom_vline(data = opt_by_sub |> dplyr::filter(subclade %in% c("IC","IIG","IIIA","IVB")),
             aes(xintercept = mean_topt, color = subclade),
             linetype = "dashed", linewidth = 0.75,
             inherit.aes = FALSE)
print(temp_curve_avg_only_mean)

ggpubr::ggarrange(temp_curve_avg_only_mean, pca_plot.ssfull_topSubclades,
                  ncol = 1,align = "hv",
                  heights = c(1, 1), labels = c("A", "C"),
                  hjust = 0.05, font.label = list(size = 10))

ggsave(
      filename = "plots/temp_curves/combined_avg_curveOnly_r_pca.png"
      ,
      width    = 60,
      height   = 100,
      units    = "mm",
      dpi      = 300,
      device = "png"
    )

```

## TPC with Yeild (K)

```{r fit temperature curves with K, message=FALSE, warning=FALSE, echo=FALSE}
# Filter out strains with too many zero values



d_filtered_k <- d %>%
  group_by(strain, date) %>%
  filter(sum(rate == 0) / n() < 0.5) %>% # Remove strains with 25% or more zero values
  ungroup() |>
  mutate(strain.date = paste(strain, date, sep = ".")) |>
  select(-c(#strain, date,
            r_sd, rate, k_sd)) |>
  rename(rate = k_mean) |>
  relocate(strain.date, .before = 1)

# Filter out strains with too many zero values and keep them for later analysis
d_filtered_out <- d %>%
  group_by(strain, date) %>%
  filter(sum(rate == 0) / n() > 0.5) %>% # Remove strains with 25% or more zero values
  ungroup()

d_filtered_k |>
  filter(temp  %in% c(4,9))


# Set up progress bar
number_of_models <- 5
number_of_curves <- length(unique(d_filtered$strain.date))
pb <- progress::progress_bar$new(total = number_of_curves * number_of_models,
                                 clear = T,
                                 format = "[:bar] :percent :elapsedfull")

set.seed(711)
# Fit models

d_fits_k <- d_filtered_k %>%
  nest(data = c(temp, rate)) %>%
  mutate(
    sharpeschoolhigh = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolhigh_1981')
      
      nls_multstart_progress(
        rate ~ sharpeschoolhigh_1981(temp = temp, r_tref, e, eh, th, tref = 20),
        data = .x,
        iter = c(3, 3, 3, 3),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    sharpeschoolfull = map(data, ~ {
      start_vals <- get_start_vals(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      lower_lims <- get_lower_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      upper_lims <- get_upper_lims(.x$temp, .x$rate, model_name = 'sharpeschoolfull_1981')
      nls_multstart_progress(
        rate ~ sharpeschoolfull_1981(temp = temp, r_tref, e, el, tl, eh, th, tref = 20),
        data = .x,
        iter = rep(3, 6),
        start_lower = pmax(start_vals - 1, lower_lims),
        start_upper = pmin(start_vals + 1, upper_lims),
        lower = lower_lims,
        upper = upper_lims,
        supp_errors = 'Y',
        convergence_count = FALSE
      )
    }),
    pawar = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, model_name = "pawar_2018")
      ll <- get_lower_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      ul <- get_upper_lims(.x$temp, .x$rate, model_name = "pawar_2018")
      nls_multstart_progress(
        rate ~ pawar_2018(temp = temp, r_tref, e, eh, th, tref = 20),
        data        = .x,
        iter        = rep(3, length(sv)),
        start_lower = sv - 5,
        start_upper = sv + 5,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
    johnsonlewin   = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "johnsonlewin_1946")
      ll <- get_lower_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      ul <- get_upper_lims(.x$temp, .x$rate, "johnsonlewin_1946")
      nls_multstart_progress(
        rate ~ johnsonlewin_1946(temp = temp, r0, e ,eh, topt),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 2,
        start_upper = sv + 2,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    }),
      weibull        = map(data, ~ {
      sv <- get_start_vals(.x$temp, .x$rate, "weibull_1995")
      ll <- get_lower_lims(.x$temp, .x$rate, "weibull_1995")
      ul <- get_upper_lims(.x$temp, .x$rate, "weibull_1995")
      nls_multstart_progress(
        rate ~ weibull_1995(temp = temp, a, topt, b, c),
        data        = .x,
        iter        = rep(4, length(sv)),
        start_lower = sv - 1,
        start_upper = sv + 1,
        lower       = ll,
        upper       = ul,
        supp_errors = "Y",
        convergence_count = FALSE
      )
    })
  )



error_rows_k <- d_fits_k %>%
  filter(
    map_lgl(sharpeschoolhigh, is.null) |
    map_lgl(sharpeschoolfull,   is.null) |
    map_lgl(pawar,          is.null) |
    map_lgl(johnsonlewin,          is.null) |
    map_lgl(weibull,          is.null) #|
    #map_lgl(briere2,          is.null)
    )

successful_fits_k <- d_fits_k %>%
  filter(
    !map_lgl(sharpeschoolhigh, is.null) |
    !map_lgl(sharpeschoolfull,   is.null) |
    !map_lgl(pawar,          is.null) |
    !map_lgl(johnsonlewin,          is.null) |
    !map_lgl(weibull,          is.null)
    )

# Print summary of errors
print("Rows with fitting errors:")
print(error_rows)
```

```{r fitted temperature curves analysis k, message=FALSE, warning=FALSE, echo=FALSE}
# 1. Pivot to long: one row per strain × model
d_long_k <- successful_fits_k %>%
  pivot_longer(
    cols      = sharpeschoolhigh:weibull,     # or your five model columns
    names_to  = "model",
    values_to = "fit"
  ) %>%
  filter(!map_lgl(fit, is.null))             # drop failed fits

# 2. Compute AIC, AICc, BIC, sigma, df.resid
d_stats_k <- d_long_k %>%
  mutate(
    # glance() gives AIC, BIC, sigma, df.resid, etc.
    glance = map(fit, broom::glance),
    AICc   = map_dbl(fit, MuMIn::AICc)
  ) %>%
  unnest(glance) %>%
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  select(strain, strain.date, model, AIC, AICc, BIC, sigma, df.residual)

# 3. Within each strain, compute ΔAICc and Akaike weight
d_compare_k <- d_stats_k %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()

# 4a. “Best” model per strain
best_by_strain_k <- d_compare_k %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, AICc, weight)

overall_k <- d_compare_k %>%
  group_by(model) %>%
  summarise(
    prop_best   = mean(delta == 0),
    mean_weight = mean(weight)
  ) %>%
  arrange(desc(prop_best), desc(mean_weight))

print(overall_k)

# 4b. (Optional) Full model‐averaged prediction or parameters
#    You can join `weight` back onto your predictions or 
#    onto rTPC::calc_params() output and take weighted averages.

# === Examples of exploring the results ===

# How many strains picked each model?
best_by_strain_k %>%
  count(best_model) |>
  mutate(percent = (n/sum(n))*100)


# 1. See exactly which rows are infinite
d_compare_k %>% 
  filter(!is.finite(AICc)) %>% 
  count(model) %>% 
  arrange(desc(n)) 

# Peek at means
d_compare_k %>% 
  filter(is.finite(AICc)) |>
  group_by(model) |>
  summarise(mean = mean(AICc, na.rm = T),
            sd   = sd(AICc, na.rm = T),
            n    = n())

hist(d_compare_k$delta[d_compare_k$model == "sharpeschoolfull"], breaks = 100)

# Visualize ΔAICc distributions
ggplot(d_compare_k, aes(model, delta)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = expression(Delta~AIC[c]), x = NULL)

hist(d_compare_k$AICc[d_compare_k$model == "sharpeschoolfull"])

# Visualize AICc distributions
ggplot(d_compare_k, aes(model, AICc)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "AICc", x = NULL)


# -----------------------------------------------------------------------------
# 2. Compute fit statistics & Akaike weights
# -----------------------------------------------------------------------------
d_stats_k <- d_stats_k %>%
  group_by(strain.date) %>%
  mutate(
    delta  = AICc - min(AICc),
    weight = exp(-0.5 * delta) / sum(exp(-0.5 * delta))
  ) %>%
  arrange(strain, delta) %>%
  ungroup()


best_by_strain_k <- d_stats_k %>%
  filter(delta == 0) %>%
  select(strain, strain.date, best_model = model, best_AICc = AICc)

# -----------------------------------------------------------------------------
# 3. Extract TPC trait parameters and do model-averaging vs best-model
# -----------------------------------------------------------------------------

params_long_k <- d_long_k %>%
  mutate(params = map(fit, calc_params, .progress = T)) %>%
  select(strain.date, model, params) %>%
  unnest(params) |>
  pivot_longer(
    cols      = c(rmax, topt, ctmin, ctmax, e, eh, q10,
                  thermal_safety_margin, thermal_tolerance,
                  breadth, skewness),
    names_to  = "term",
    values_to = "estimate"
  )
# → columns: strain, model, term, estimate


```

```{r model-avg-params-and-preds k, message=FALSE, warning=FALSE, echo=FALSE}
# 4b) augment each fit
d_preds_k <- d_long_k %>%
  left_join(pred_grids, by = "strain.date") %>%
  mutate(preds = map2(fit, newdata, ~ augment(.x, newdata = .y), .progress = T)) %>%
  select(strain.date, model, preds) %>%
  unnest(preds)

# 4c) join weights and best‐model info
d_preds_k <- d_preds_k %>%
  left_join(d_stats_k %>% select(strain, strain.date, model, weight), by = c("strain.date","model")) %>%
  left_join(best_by_strain_k, by = "strain.date")

# 4e) extract best‐model prediction
best_preds_k <- d_preds_k %>%
  filter(model == "sharpeschoolfull") %>%
  select(strain.date, temp, best_fitted = .fitted) |>
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = FALSE       # keep the original strain.date if you still need it
  ) %>%
  # (optionally) turn your date string into a real Date
  mutate(
    date = as.Date(date)
  )

# -----------------------------------------------------------------------------
# Now you have:
# - compare_params: per‐strain, per‐trait avg vs best
# - avg_preds & best_preds: per‐strain fitted curves
# -----------------------------------------------------------------------------
```

### Plot predictions and observations

```{r preds k , message=FALSE, warning=FALSE, echo=FALSE}

preds_k <- merge(best_preds_k, clade_info,
               by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(!is.na(subclade))

d_filtered_wmd_k <- merge(d_filtered_k |> separate(
                                col    = strain.date,
                                into   = c("strain","date"),
                                sep    = "\\.",      # split on dot
                                extra  = "merge",    # if there are more dots, merge them into the last column
                                remove = FALSE),     # keep the original strain.date if you still need it
                        clade_info,
                        by.x = "strain", by.y = "StrainID", all.x = T
                        ) |>
  left_join(d |> select(strain,
                        date,
                        temp,
                        k_sd),
            by = c("strain", "date", "temp")) |>
  filter(!is.na(subclade))

```

We can then plot the predictions of each curve using **ggplot2**.

```{r plot fitted thermal curves k, fig.height = 12, fig.width=8, message=FALSE, warning=FALSE, echo=FALSE}

 ## plot top subclades
plot_temp_curves(
  preds          = preds_k,
  observed       = d_filtered_wmd_k,
  all_data       = all.data,
  strain_cols    = strain_subclade_col,
  subclades      = c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE"),
  facet_by_clade = FALSE,
  filename       = "plots/temp_curves_topSubclades_k.png",
  nrow           = 1,
  width_mm       = 300,
  height_mm      = 75
)

 ## plot all subclades
plot_temp_curves(
  preds          = preds_k,
  observed       = d_filtered_wmd_k,
  all_data       = all.data,
  strain_cols    = strain_subclade_col,
  subclades      = NULL,
  facet_by_clade = T,
  filename       = "plots/temp_curves.png",
  nrow           = 1
)


  correct_order <- c(
  "IA", "IB", "IC",
  "IIA", "IIB", "IIC", "IID", "IIE", "IIF", "IIG", "IIH", "IIJ", "IIK",
  "IIIA", "IIIB", "IIIC", "IIID", "IIIE", "IIIF",
  "IVA", "IVB", "IVC", "IVD", "IVE",
  "VA"
)

preds_k <- preds_k |> mutate(subclade = factor(subclade, levels = correct_order))


for (subcld in c("IA", "IB", "IC",
                 "IIE", "IIG", "IIJ",
  "IIIA", "IIIB", "IIID", "IIIE",
  "IIIF",
  "IVA", "IVB", "IVC", "IVE", NA)) {
  
   # get all strains in this subclade
  all_strains <- preds_k %>%
    filter(subclade == subcld) %>%
    pull(strain) %>%
    unique() %>%
    sort()
  
   # split into chunks of up to 20
  strain_chunks <- split(all_strains, 
                         ceiling(seq_along(all_strains) / 20))
  
  # loop over each chunk (page)
  for (i in seq_along(strain_chunks)) {
    this_strains <- strain_chunks[[i]]
    
    temp_curve <- preds_k %>%
      filter(subclade == subcld, strain %in% this_strains) |> 
  ggplot(aes(x = temp,
                    color = strain,
                    group = strain.date,
                    y = best_fitted)) +
  scale_color_manual(values = strain_subclade_col, guide = "none") +
  geom_line(alpha = 1, linewidth = 2) +
  geom_point(data = d_filtered_wmd_k %>% filter(subclade == subcld,
                                              strain %in% this_strains),
             aes(temp, rate, color = strain, shape = date), alpha = 0.75, size = 5) +
  scale_shape_discrete(name = "Date") +
  #geom_smooth(aes(fill = strain)) +
  #scale_fill_manual(values = strain_clade_col) +
  geom_errorbar(data = d_filtered_wmd_k %>% filter(subclade == subcld,
                                              strain %in% this_strains),
                  aes(x = temp, ymin = rate-k_sd, ymax = rate+k_sd, color = strain),
                  width = .4, inherit.aes = F, alpha = 0.65) +
  scale_x_continuous(breaks = sort(unique(all.data$Temp))) +
  ggh4x::facet_nested_wrap(.~strain, ncol = 4) +
  ylim(0, 0.65) +
  labs(title = paste("Subclade", subcld, "(page", i, "of", length(strain_chunks), ")")
       #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))
       ) +
  labs(y = expression("Yield (OD" [600] * ")")) +
  xlab("Temperature (ºC)") +
  apatheme +
  theme(
        panel.grid.major.x = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.major.y = element_line(color = "grey85", linetype = "dashed"),
        panel.grid.minor   = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 10),
        legend.position    = "bottom",
        legend.box         = "horizontal",
        axis.text.y = element_text(size = 20),
        strip.background = element_blank(),
        text = element_text(size = 35),
        #strip.text.x = element_blank(),
        plot.title = element_text(color = "black", size = 24 ),
        plot.subtitle = element_text(color = "black", size = 16))
    
  print(temp_curve)
  
  # now save it
    fn <- sprintf("plots/temp_curves/temp_curves_k_%s_page%02d.png", subcld, i)
    ggsave(
      filename = fn,
      plot     = temp_curve,
      width    = 600,
      height   = 350,
      units    = "mm",
      dpi      = 300
    )
  }
}



    
    
    
preds_avg_k <- preds_k |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND"))) |>
  group_by(subclade, temp) |>
  summarise(avg_k = mean(best_fitted))





### first recode subclade IB
temp_curve_avg_k <- preds_k %>%
    filter(!(
      (strain == "32G3B09" & date == "2024-03-14") |
      (strain == "39G1A09" & date == "2024-03-14") |
      (strain == "39G1A10" & date == "2024-03-14") |
      (strain == "30G3A10" & date == "2024-04-15") |
      (strain == "33G2A09" & date == "2024-04-15") |
      (strain == "35G4A05" & date == "2024-04-15") |
      (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
      (strain == "28G1A03" & date == "2024-03-14") |
      (strain == "28G1B05" & date == "2024-03-14") |
      (strain == "28G4A04" & date == "2024-03-14") |
      (strain == "35G1A07" & date == "2024-03-14") |
      (strain == "39G1A08" & date == "2024-03-14") |
      (strain == "30D10" & date == "2024-03-14") |
      (strain == "23G1B05" & date == "2024-03-14") |
      (subclade == "IIIE" & date == "2024-04-15") |
      (strain == "30G1B03" & date == "2024-03-14") |
      (strain == "23G1B05" & date == "2024-03-14") |
      (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
      (strain == "36G2A07" & date == "2024-03-14") |
      (strain == "36G3B02" & date == "2024-03-14") |
      (strain == "36G3B05" & date == "2024-03-14") |
      (strain == "37G4A03" & date == "2024-03-14") |
      (clade == "UND"))) |>
      filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) |>
      ggplot(aes(x = temp,
                        color = strain,
                        group = strain.date,
                        y = best_fitted)) +
      geom_line(alpha = 0.25, linewidth = 0.1) +
      scale_color_manual(values = strain_clade_col, guide = "none") +
      # 2) tell ggplot we want a *new* colour scale
      ggnewscale::new_scale_color() +
      geom_smooth(data = preds_avg_k |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")),
                 aes(temp, avg_k, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = T) +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_col) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_col) +
      scale_x_continuous(breaks = seq(4,46, 6)) +
      ylim(0, 0.5) +
      facet_wrap(.~ subclade, nrow = 1) +
      #labs(title = "Temperature performance",
           #subtitle = substitute(paste("mean growth rate (", italic("r"), ") by Temperature"))) +
      #ylab(substitute(paste("Growth rate (", italic("r"),")"))) +
      ylab(expression("Yield (OD" [600] * ")")) +
      xlab("Temperature (ºC)") +
      apatheme +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 4.5),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            #strip.text.x = element_blank(),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
            #legend.position = c(0.95,0.85),
            #legend.justification = c(1, 1),         # anchor legend’s top-right corner
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(temp_curve_avg_k)


ggsave(
      filename = "plots/temp_curves/temp_curves_avg_topSubclades_k.png",
      width    = 140,
      height   = 80,
      units    = "mm",
      dpi      = 300
    )
  

```

#### Thermal performance parameters

The traits of each thermal performance curve can also easily be
calculated.

```{r params k, message=FALSE, warning=FALSE, echo=FALSE}
# stack models and calculate extra params
d_params_k <- pivot_longer(d_fits_k, names_to = 'model_name', values_to = 'fit',
                         c(sharpeschoolhigh,sharpeschoolfull,pawar)) %>%
  mutate(params = map(fit, calc_params, .progress = T)) %>%
  select(strain.date, model_name, params) %>%
  unnest(params) %>%
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = F       # keep the original strain.date if you still need it
  ) %>%
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

glimpse(d_params_k)

d_params_k <- d_params_k |>
   filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND")))

d_params_k <- d_params_k %>% filter(!is.na(clade))


d_params_ssfull_k <- d_params_k %>%
  filter(model_name == "sharpeschoolfull") %>%
  select(strain, date, strain.date, rmax, ctmin, ctmax, topt, breadth) %>%
  filter(date != "2024-01-14") |>
  select(-c(date)) %>%
  tibble::column_to_rownames("strain.date")

d_params_ssfull_k[sapply(d_params_ssfull_k, is.infinite)] <- NA

pca_k <-prcomp(na.omit(d_params_ssfull_k) %>% filter(!strain %in% c("09D09", "23G4B04")) |> select(-strain)
             ,
             scale. = T, center = T)


pca.df_k <- data.frame(pca_k$x) %>%
  select(PC1, PC2) %>%
  tibble::rownames_to_column("strain.date") |> 
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # keep the original strain.date if you still need it
  ) %>%
    merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

PCAloads_k<-data.frame(Variables = rownames(pca.df_k$rotation), pca.df_k$rotation)

pca_plot_k<-ggplot(pca.df_k #%>% filter(Row.names != "23G4B04")
                 , aes(PC1, PC2)) +
  geom_point(data = pca.df, aes(color = subclade),  size = 5, alpha = 0.5) +
  #geom_text(data = pca.df, aes(x=PC1, y=PC2-0.15), label=pca.df$Row.names, size=2.75) +
  scale_color_manual(values = subclade_col, name="Clade") +
  geom_segment(data = PCAloads,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               arrow = arrow(length = unit(0.3, "cm")),
               colour="grey", inherit.aes = FALSE) +
 geom_text(data=PCAloads,
           aes(x=PC1*2.5, y=PC2*2.5, label=Variables),
           size=6,
           colour = "darkblue") +
  xlab(paste0("PC1 (",round(summary(pca_k)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca_k)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 18))
pca_plot_k

summary(pca_k)


d_params_ssfull_topSubclades_k <- d_params_k %>%
  filter(model_name == "sharpeschoolfull") %>%
  filter(!strain %in% c("09D09", "23G4B04")) |>
  filter(subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVC")) |>
  select(date, strain.date, ctmin, ctmax, topt, breadth) %>%
  filter(date != "2024-01-14") |>
  select(-c(date)) %>%
  tibble::column_to_rownames("strain.date")

pca_k_topSubclades <-prcomp(na.omit(d_params_ssfull_topSubclades_k), scale. = T, center = T)

pca_k_topSubclades_df <- data.frame(pca_k_topSubclades$x) %>%
  select(PC1, PC2) %>%
  tibble::rownames_to_column("strain.date") |> 
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # keep the original strain.date if you still need it
  ) %>%
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

PCAloads_k_topSubclades <- data.frame(Variables = rownames(pca_k_topSubclades$rotation),
                                      pca_k_topSubclades$rotation) |>
                                        mutate(Variables = recode(Variables,
                                                        "ctmin" = "Min. Temp",
                                                        "topt" = "Opt. Temp",
                                                        "ctmax" = "Max. Temp",
                                                        "breadth" = "Breadth"))


pca_plot.k_topSubcladesl<-ggplot(pca_k_topSubclades_df
                 , aes(PC1, PC2)) +
  geom_point(data = pca_k_topSubclades_df, aes(color = subclade),  size = 2, alpha = 0.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads_k_topSubclades,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               colour="grey", inherit.aes = FALSE) +
 geom_label(data=PCAloads_k_topSubclades,
           aes(x=PC1*2.5, y=PC2*2.5, label=Variables),
           size=2,
           colour = "darkblue") +
  lims(x = c(-4,4),
       y = c(-3,3)) +
  xlab(paste0("PC1 (",round(summary(pca_k_topSubclades)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca_k_topSubclades)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 10),
        legend.position = "inside",
        legend.position.inside = c(0.85, 0.85),
        legend.background = element_rect(fill = alpha("white", 0), color = NA),
        legend.key.size = unit(0.5, "lines"),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 7),
        legend.key.height = unit(0.5, "lines"),
        legend.key.width = unit(0.8, "lines"),
        legend.spacing.y = unit(0.1, "cm"))
pca_plot.k_topSubcladesl

ggsave(
      filename = "plots/pca_plot.ssfull_topSubclades_k.pdf",
      plot     = pca_plot.k_topSubcladesl,
      width    = 90,
      height   = 90,
      units    = "mm",
      dpi      = 300,
      device = cairo_pdf
    )

# pick a handful of key traits
traits <- c("rmax", "ctmax", "topt", "breadth")

d_params_ssfull_k %>%
  # merge with clade info
  merge(., clade_info, by.x = "strain", by.y = "StrainID", all.x = T) |>
  filter(clade != "UND") |>
  filter(!is.na(subclade)) |> 
  # raname coulumn row.names to strain
  #rename(strain = Row.names) |>
  # select columns of interest
  select(strain, clade, subclade, all_of(traits)) %>%
  # pivot longer for plotting
  pivot_longer(cols = traits, names_to = "term", values_to = "value") %>%
  # plot 
  ggplot(aes(subclade, value, fill = subclade)) +
    geom_boxplot() +
    scale_fill_manual(values = subclade_col, name="Subclade") +
    ggh4x::facet_nested_wrap(~ term + clade, scales = "free", ncol = 4) +
    ggpubr::stat_compare_means(aes(label = ..p.signif..), method = "kruskal.test") +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(x = "Subclade", y = "Estimate", title = "Trait differences among Subclades")



### dot plots ####

out_k <- plot_traits_subclade(
  params_df = d_params_ssfull_k,
  clade_info = clade_info,
  traits = c("breadth","ctmax","topt"),
  subclades = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  #subclades = NULL,                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    topt  = "Opt. Temp.",
    ctmax  = "Max. Temp.",
    breadth = "Breadth (Temp)"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("output/temp/plots/curve_params_dotplot_Subclades_flip_k.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)




ggpubr::ggarrange(temp_curve_avg_k, out$plot,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = "AUTO",
                  hjust = 0.05, font.label = list(size = 10))

ggsave(
      filename = "plots/temp_curves/combined_avg_params_k.png"  #"~/SyncFolder/CA_curto/phys_assays/temp/processed_data/plots/curve_params_dotplot_Subclades.png"
      ,
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )


```

#### Fitting curves for Yeild (K) with LOESS

##### Temperature curves parameters calculation

Now we can calculate the pH curves parameters for each strain:

1.  Area under the curve
2.  Range
3.  Optimal pH

```{r calculate curve paramters, message=TRUE, warning=TRUE}
library(pracma)

head(d)

#write.table(random_data[, c(1:3)], file = "rTPC_test_data.tsv", row.names = FALSE, quote = F, sep = "\t")

# Function to calculate area under the curve for each strain
calculate_auc <- function(data) {
  if (nrow(data) == 0) {
    return(data.frame(strain = NA, area_under_curve = NA))
  }
  
  loess_model_k <- loess(k_mean ~ temp, data = data)
  predicted_values_k <- predict(loess_model_k, data$temp)
  area_under_curve_k <- trapz(data$temp, predicted_values_k)
  return(data.frame(strain = unique(data$strain),
                    date = unique(data$date),
                    area_under_curve_k = area_under_curve_k))
}

# Group by strain and calculate area under the curve for each strain
result_auc <- d %>%
  group_by(strain, date) %>%
  group_split() %>%
  lapply(calculate_auc) %>%
  bind_rows()

# Print the result data frame
print(result_auc)


# Function to calculate max r_mean and its corresponding pH for each strain
get_max_k_mean <- function(data) {
  # Fit the loess model
  loess_model_k <- loess(k_mean ~ temp, data = data)
  
  # Predict values for a range of x-values
  predicted_values_k <- predict(loess_model_k, data$temp)
  
  # Find the index of the maximum r_mean value
  max_index_k <- which.max(predicted_values_k)
  
  # Get the corresponding pH value at the maximum r_mean
  max_temp_k <- data$temp[max_index_k]
  
  # Get the maximum r_mean value
  max_k_mean <- predicted_values_k[max_index_k]
  
  # Return a data frame with the results
  data.frame(strain = unique(data$strain),
             date = unique(data$date),
             opt_temp_k = max_temp_k,
             max_k_mean = max_k_mean)
}

# Apply the function for each strain in the result_df
max_k_mean_results <- d %>%
  group_by(strain, date) %>%
  do(get_max_k_mean(.))

# Print the results
print(max_k_mean_results)



result_range_k <- d %>%
  filter(k_mean != 0) %>%
  group_by(strain, date) %>%
  summarise(min_temp_k = min(temp),
            max_temp_k = max(temp),
            breadth_k = max(temp) - min(temp))

print(result_range_k)



curve_params <- merge(result_auc, max_k_mean_results, by = c("strain", "date")) %>%
  merge(., result_range_k, by = c("strain", "date")) |>
  left_join(clade_info, by = c("strain" = "StrainID")) |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND") | is.na(clade)),
    date != "2024-01-14")

#write.table(as.data.frame(curve_params$strain), file = "~/SyncFolder/CA_curto/isolates/fastANI_out_WI/working_isolates.tsv", row.names = FALSE, quote = F, sep = "\t")

#write.table(as.data.frame(curve_params |>
#                            select(strain) |>
#                            rename(StrainID = strain) |>
#                            mutate(pH = "yes")),
#            file = "~/SyncFolder/CA_curto/phys_assays/pH/pH_strain_table.txt", row.names = FALSE, quote = #F, sep = "\t")


#write.table(curve_params, file = "curve_params.tsv", row.names = FALSE, quote = F, sep = "\t")


### PCA of curve parameters
curve_params <- curve_params %>%
  mutate(strain.date = paste(strain, date, sep = ".")) |> tibble::column_to_rownames("strain.date")


k_traits <- c(colnames(curve_params |> select(ends_with("_k") & !starts_with("area"))), "max_k_mean")

pca <-prcomp(curve_params |>
               filter(subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVC")) |>
               select(all_of(k_traits))
             , scale. = T, center = T)


pca.df <- data.frame(pca$x) %>%
  select(PC1, PC2) %>%
  tibble::rownames_to_column("strain.date") |> 
  separate(
    col    = strain.date,
    into   = c("strain","date"),
    sep    = "\\.",      # split on dot
    extra  = "merge",    # if there are more dots, merge them into the last column
    remove = T       # keep the original strain.date if you still need it
  ) |> left_join(clade_info, by = c("strain" = "StrainID"))

pca.df$clade_color <- clade_colors$color[match(pca.df$clade, clade_colors$variable)]
pca.df$subclade_color <- clade_colors$color[match(pca.df$subclade, clade_colors$variable)]
PCAloads <- data.frame(Variables = rownames(pca$rotation), pca$rotation) |>
  mutate(Variables = recode(Variables,
                            "min_temp_k" = "Min. temp.",
                            "opt_temp_k" = "Opt. temp.",
                            "max_temp_k" = "Max. temp.",
                            "max_k_mean" = "Yeild",
                            "breadth_k" = "Breadth"))

pca_plot<-ggplot(pca.df
                 , aes(PC1, PC2)) +
  geom_point(data = pca.df, aes(color = subclade),  size = 5, alpha = 0.75) +
  #geom_text(data = pca.df, aes(x=PC1, y=PC2-0.15), label=pca.df$Row.names, size=2.75) +
  scale_color_manual(values = subclade_col, name="Subclade") +
  geom_segment(data = PCAloads,
               aes(x=0,xend=PC1*2.5,y=0,yend=PC2*2.5),
               arrow = arrow(length = unit(0.3, "cm")),
               colour="grey", inherit.aes = FALSE) +
 ggrepel::geom_label_repel(data=PCAloads,
           aes(x=PC1*2.5, y=PC2*2.5, label=Variables),
           size=4,
           colour = "darkblue") +
 #annotate("text", x = -7.5 , y = -6, label = "Stress = 0.1", size = 3.5) +
 #annotate("text", x = -7.5, y = -6.5, label = "[PERMANOVA ~ Site: F = 3.28, P < 0.001]", size = 3.5) +
  xlab(paste0("PC1 (",round(summary(pca)$importance[2,1] * 100, 2),"%)")) +
  ylab(paste0("PC2 (",round(summary(pca)$importance[2,2] * 100, 2),"%)")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 18))
pca_plot

summary(pca)


### dot plots ####

out_k_loess <- plot_traits_subclade(
  params_df = curve_params |> select(strain, all_of(k_traits)),
  clade_info = clade_info,
  traits = k_traits,
  subclades = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  #subclades = NULL,                 # or c("IC","IIG","IIIA","IIIE","IVB","IVE")
  #exclude_subs = c("IIIB","IVA"),
  subclade_levels = c("IC","IIG","IIIA","IIIE","IVB","IVE"),
  facet_by_clade = F,
  flip = T,
  term_labels = c(
    opt_temp_k  = "Opt. Temp.",
    max_temp_k  = "Max. Temp.",
    min_temp_k = "Min. Temp.",
    breadth_k = "Breadth",
    max_k_mean = "Yeild"
  ),
  show_annotations = "both",
  omnibus_in_clade = "I",           # show omnibus star in clade I row
  subclade_col = subclade_col,      # named vector of colours
  filename = here("output/temp/plots/curve_params_dotplot_Subclades_flip_k_loess.png"),
  width_mm = 180, height_mm = 120, dpi = 300
)




avg_k <- d_filtered_wmd_k |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND") | is.na(clade)),
    date != "2024-01-14") |>
  group_by(subclade, temp) |>
  summarise(avg_k = mean(rate))


temp_curve_avg_k_subclades_loess <- d_filtered_wmd_k |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND") | is.na(clade)),
    date != "2024-01-14") |>
                        filter(clade != "UND" & !is.na(subclade) &
                                 subclade %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")) |>
   ggplot(aes(x = temp,
                    color = strain,
                    group = strain,
                    y = rate)) +
  scale_color_manual(values = strain_subclade_col) +
  geom_smooth(aes(fill = strain),
              se = F, linewidth = 0.1, alpha = 0.25) +
  scale_fill_manual(values = strain_subclade_col) +
  scale_x_continuous(breaks = seq(4,46, 6)) +
      # 2) tell ggplot we want a *new* colour scale
      ggnewscale::new_scale_color() +
      geom_smooth(data = avg_k |> filter(subclade  %in% c("IC", "IIG", "IIIA", "IIIE", "IVB", "IVE")),
                 aes(temp, avg_k, color = subclade, fill = subclade), alpha = 0.35,
                 method = "gam",
                 inherit.aes = F,
                linewidth = 0.75,
                se = T) +
      scale_colour_manual(name   = "Subclade",
                          values = subclade_col) +
      scale_fill_manual(name   = "Subclade",
                              values = subclade_col) +
      ylim(0, 0.4) +
      facet_wrap(.~ subclade, nrow = 1) +
      labs(y = expression("Yield (OD" [600] * ")")) +
      xlab("Temperature (Cº)") +
      apatheme +
      theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 4.5),
            axis.text.y = element_text(size = 6),
            strip.background = element_blank(),
            text = element_text(size = 8),
            plot.title = element_text(color = "black", size = 18),
            plot.subtitle = element_text(color = "black", size = 16),
            legend.position = "none",
    legend.background    = element_rect(fill = scales::alpha("white", 0.6),
                                        colour = NA),
    legend.title         = element_text(size = 5),
    legend.text          = element_text(size = 4),
    legend.margin        = margin(rep(1,4)))
print(temp_curve_avg_k_subclades_loess)

ggpubr::ggarrange(temp_curve_avg_k_subclades_loess, out_k_loess$plot,
                  ncol = 1,align = "hv",
                  heights = c(1, 1.5), labels = "AUTO",
                  hjust = 0.05, font.label = list(size = 10))

ggsave(
      filename = "plots/temp_curves/combined_avg_params_k_loess.png"  #"~/SyncFolder/CA_curto/phys_assays/temp/processed_data/plots/curve_params_dotplot_Subclades.png"
      ,
      width    = 100,
      height   = 100,
      units    = "mm",
      dpi      = 300
    )




```

######### Count of strains with acceptable data

```{r count of strains, echo=FALSE, warning=FALSE, message=FALSE}

### let's get a list of all strains that have been tested for temperature performance

strain_table <- d_params |>
  select(strain) |>
  distinct() |>
  rename(StrainID = strain) |>
  mutate(temp = "yes")

write.table(strain_table,
            file = here("output/temp/temp_strain_table.txt"),
            row.names = F,
            col.names = T,
            quote = F)



### lets also export the curve parameters data for correlations


# Rebuild the per-curve sharpeschoolfull table from the stable `d_params`.
# NOTE: the `d_params_ssfull` object is mutated further above (a left_join
# drops its strain.date rownames), so using it here in a clean top-to-bottom
# render produces NA dates and an empty table. `d_params` already carries
# clade/subclade and the per-run strain/date exclusions applied above.
d_params_ssfull_export <- d_params |>
  filter(model_name == "sharpeschoolfull", date != "2024-01-14") |>
  select(strain, date, rmax, ctmin, ctmax, topt, breadth, clade, subclade)


dups <- data.frame(StrainID = d_params_ssfull_export$strain[duplicated(d_params_ssfull_export$strain)])

dups <- merge(dups, clade_info, by = "StrainID", all.x = T)

d_params_ssfull_export <- d_params_ssfull_export |>
  filter(!(
    (strain == "32G3B09" & date == "2024-03-14") |
    (strain == "39G1A09" & date == "2024-03-14") |
    (strain == "39G1A10" & date == "2024-03-14") |
    (strain == "30G3A10" & date == "2024-04-15") |
    (strain == "33G2A09" & date == "2024-04-15") |
    (strain == "35G4A05" & date == "2024-04-15") |
    (strain == "23G1B05" & date == "2024-03-14") | ## might have to repeat this one for a few temps
    (strain == "28G1A03" & date == "2024-03-14") |
    (strain == "28G1B05" & date == "2024-03-14") |
    (strain == "28G4A04" & date == "2024-03-14") |
    (strain == "35G1A07" & date == "2024-03-14") |
    (strain == "39G1A08" & date == "2024-03-14") |
    (strain == "30D10" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (subclade == "IIIE" & date == "2024-04-15") |
    (strain == "30G1B03" & date == "2024-03-14") |
    (strain == "23G1B05" & date == "2024-03-14") |
    (strain  %in%  c("26G3A01", "26G3A06") & date == "2024-04-15") |
    (strain == "36G2A07" & date == "2024-03-14") |
    (strain == "36G3B02" & date == "2024-03-14") |
    (strain == "36G3B05" & date == "2024-03-14") |
    (strain == "37G4A03" & date == "2024-03-14") |
    (clade == "UND" & !is.na(clade)))) |>
  select(-c(clade, subclade))


write.table(d_params_ssfull_export,
            file = here("output/temp/temp_curve_params_table.txt"),
            row.names = F,
            col.names = T,
            quote = F)



```

#### Correlation distance matrix correlation

For the next steep we have calculated pairwise ANI distance for all
strains. We will calculate distance matrices for all parameters all
together and also individually and assess the correlation at different
ANI distances

##### Distance matrix of curve parameters

```{r distance matrix of curve parameters and mantel tests,  message=FALSE, warning=FALSE, echo=FALSE, fig.height = 18, fig.width=24, eval=FALSE}
# NOTE: ANI-vs-thermal Mantel analysis disabled for now (deferred).
# It needs NA-robust handling of curve parameters before it will run.
library(progress)


## Read distance data in table format
dat <- read.table(here("data/shared/fastANI.out"), header = F, sep = "\t")
#read in distance matrix
dm <- as.matrix(read.table(here("data/shared/fastANI.out.matrix"), header = T, row.names = 1, sep = "\t"))
## Let's change the NCBI identifiers for more informative identifiers. For that we need to read in the master metadata table.
row.names(dm) <- master_md$Name[match(row.names(dm), master_md$FILE_NAME)]

#Create a function to get the upper triangle of the matrix

lower_to_symmetric <- function(lower_df) {
  n <- nrow(lower_df)
  matrix_data <- matrix(NA, nrow = n, ncol = n)
  
  for (i in 1:n) {
    for (j in 1:i) {
      matrix_data[i, j] <- lower_df[i, j]
      matrix_data[j, i] <- lower_df[i, j]
    }
  }
  
  return(matrix_data)
}

# apply the created function to the distance matrix to get a square symmetrical matrix
symmetric_matrix <- lower_to_symmetric(dm)

# Print the resulting matrix
#print(symmetric_matrix)

#set row and column names
row.names(symmetric_matrix) <- row.names(dm)
colnames(symmetric_matrix) <- row.names(dm)

# check that all row names in the similarity matrix are found in the master metadata
# result should be character(0)
setdiff(row.names(symmetric_matrix), master_md$Name)


### now let's just select rows and columns of the matrix according to the list we just loaded
symmetric_matrix <- symmetric_matrix[rownames(symmetric_matrix) %in% unique(d_params$strain), colnames(symmetric_matrix) %in% unique(d_params$strain)]
setdiff(unique(d_params$strain), rownames(symmetric_matrix))

# let's leave out from the analysis for now the following strains
symmetric_matrix <- symmetric_matrix[!rownames(symmetric_matrix) %in% c("26G3B03", "28D03", "35G3B04"), !colnames(symmetric_matrix) %in% c("26G3B03", "28D03", "35G3B04")]
## let's remove them also from the parameters table
d_params <- d_params %>% filter(!strain %in% c("26G3B03", "28D03", "35G3B04"))

## A few strains were measured on two dates; keep one row per strain within
## each model (latest date) so `strain` can serve as unique row names below.
d_params <- d_params |>
  arrange(desc(date)) |>
  distinct(model_name, strain, .keep_all = TRUE)

#now let's calculate genetic distance as  distance = 100 - ANI
gdm <- as.dist(100 - symmetric_matrix, diag = T, upper = T)


## now let's also create the phenotypic distance matrix with the curve parameters
## let's create a loop to calculate distance matrices for both models

# Initialize lists to store distance matrices and Mantel test results
distance_matrices_all <- list()
mantel_results_all <- list()


# Loop to calculate distance matrices for both models
for (model in unique(d_params$model_name)) {
  
  # Filter for each model and select parameters of interest
  z <- as.data.frame(scale(d_params %>%
                             filter(model_name == model) %>% 
                             select(strain, rmax, ctmax, topt) %>% 
                             tibble::column_to_rownames("strain")))
  
  # Calculate distance matrix for all parameters all together
  euc_z <- dist(z, method = "euclidean", diag = TRUE, upper = TRUE)
  
  distance_matrices <- list()
  mantel_results <- list()
  
  for (var in colnames(d_params %>%
                        filter(model_name == model) %>% 
                        select(strain, rmax, ctmax, topt) %>% 
                        tibble::column_to_rownames("strain"))) {
    ## check for infinie values in values
    inf_names <- d_params %>%
      filter(model_name == model) %>%
      tibble::column_to_rownames("strain") %>%
      select(all_of(var)) %>%
      filter(is.infinite(!!sym(var))) %>%
      row.names(.)
    
    if (length(inf_names) == 0) {
      
      distance_matrices[[var]] <- dist(d_params %>%
                                         filter(model_name == model) %>% 
                                         tibble::column_to_rownames("strain") %>% 
                                         select(all_of(var)),
                                       method = "euclidean", diag = TRUE, upper = TRUE)
      
      # Perform Mantel test without subsetting if no Inf values are found
      mo_var <- vegan::mantel(gdm, distance_matrices[[var]], method = "pearson", permutations = 9999)
    }
    
    else {
      distance_matrices[[var]] <- dist(d_params %>%
                                         filter(model_name == model & !strain %in% inf_names) %>% 
                                         tibble::column_to_rownames("strain") %>% 
                                         select(all_of(var)),
                                       method = "euclidean", diag = TRUE, upper = TRUE)
      
      # Remove the same rows and columns from gdm matrix
      sub_gdm <- as.matrix(gdm)[!(rownames(as.matrix(gdm)) %in% inf_names), !(rownames(as.matrix(gdm)) %in% inf_names)]
      sub_gdm <- as.dist(sub_gdm, diag = TRUE, upper = TRUE)
      
      # Perform Mantel test without subsetting if no Inf values are found
      mo_var <- vegan::mantel(sub_gdm, distance_matrices[[var]], method = "pearson", permutations = 9999)
    }
    
    # Store Mantel test results in the overall list
    mantel_results[[var]] <- list(
      statistic = mo_var$statistic,
      significance = mo_var$signif
    )
  }
  
  # Add the distance matrix that summarizes all traits to the distance_matrices list
  distance_matrices[["all_parameters"]] <- euc_z
  
  # Store the distance matrices in the overall list
  distance_matrices_all[[model]] <- distance_matrices
  
  # Perform overall Mantel test
  mo <- vegan::mantel(gdm, euc_z, method = "pearson", permutations = 9999)
  
  # Store Mantel test results
  mantel_results_all[[model]] <- mantel_results
  mantel_results_all[[model]][["all_parameters"]] <- list(
    statistic = mo$statistic,
    significance = mo$signif
  )
}

# Print Mantel test results
#print(mantel_results_all)


# load table to find groups of different degrees of similarity
ANI_df <- dat
rm(dat)
ANI_df$V1 <- master_md$Name[match(ANI_df$V1, master_md$FILE_NAME)]
ANI_df$V2 <- master_md$Name[match(ANI_df$V2, master_md$FILE_NAME)]


## now let's just keep the observations only for the working
ANI_df <- ANI_df %>%
  rename(query = V1,
         ref = V2,
         ANI = V3,
         fragment_mappings = V4,
         total_query_fragments = V5) %>%
  filter(query %in% unique(d_params$strain) &
           ref %in% unique(d_params$strain) &
           !query %in% c("26G3B03", "28D03", "35G3B04") &
           !ref %in% c("26G3B03", "28D03", "35G3B04")) %>%
  mutate(dist = 100 - ANI) %>% # first transform ANI to distance and add to tabl
  filter(dist != 0) #filter out identical pairs

ANI_df <- ANI_df %>%
  mutate(query_ordered = pmin(query, ref),   # Get the "minimum" of the two strings
         ref_ordered = pmax(query, ref)) %>% # Get the "maximum" of the two strings
  select(-query, -ref) %>%                   # Remove the original columns
  rename(query = query_ordered, ref = ref_ordered) # Rename the ordered columns back

# Remove duplicates based on the ordered pairs
ANI_df <- ANI_df %>%
  distinct(query, ref, .keep_all = TRUE)

# Subset distances within the given range
  #subset_df <- ANI_df %>% filter(dist > 0 & dist < 2.5)
  


# Function to perform Mantel test and return a data frame with results
perform_mantel_test <- function(dist_range, gdm, euc_z, permutations = 9999) {
  # Subset distances within the given range
  subset_df <- ANI_df %>% filter(dist > dist_range[1] & dist < dist_range[2])
  
  # Get unique row and column names from the subset
  subset_ids <- unique(c(subset_df$ref, subset_df$query))
  
   # Filter subset_ids to include only those present in both gdm and euc_z
  valid_ids <- intersect(intersect(subset_ids, rownames(as.matrix(gdm))), rownames(as.matrix(euc_z)))
  
  # Create submatrices using the valid IDs
  sub_gdm <- as.matrix(gdm)[valid_ids, valid_ids]
  sub_euc_z <- as.matrix(euc_z)[valid_ids, valid_ids]
  
  # Ensure sub_gdm and sub_euc_z have the same dimensions and row/column names
  common_ids <- intersect(rownames(sub_gdm), rownames(sub_euc_z))
  sub_gdm <- sub_gdm[common_ids, common_ids]
  sub_euc_z <- sub_euc_z[common_ids, common_ids]
  
  # Perform Mantel test and extract the correlation coefficient and p-value
  mantel_result <- vegan::mantel(sub_gdm, sub_euc_z, method = "pearson", permutations = permutations)
  
   # Count the number of comparisons
  num_comparisons <- length(subset_ids) * (length(subset_ids) - 1) / 2
  # Count the number of unique strains
  num_strains_in_range <- length(subset_ids)
  
  # Check if mantel_result$signif is not NULL and not NA
  if (!is.null(mantel_result$signif) && !is.na(mantel_result$signif)) {
    # If there are significant results, create a data frame with results
    result_df <- data.frame(Range = paste(dist_range, collapse = "-"),
                            Correlation_Coefficient = mantel_result$statistic,
                            P_Value = mantel_result$signif,
                            Num_Comparisons = num_comparisons,  # Add number of comparisons
                            Num_strains = num_strains_in_range,
                            stringsAsFactors = FALSE)
  } else {
    # If there are no significant results, create a data frame with NA values
    result_df <- data.frame(Range = paste(dist_range, collapse = "-"),
                            Correlation_Coefficient = NA,
                            P_Value = NA,
                            Num_Comparisons = num_comparisons,  # Add number of comparisons
                            Num_strains = num_strains_in_range,
                            stringsAsFactors = FALSE)
  }
  
  return(result_df)
}


generate_non_overlapping_ranges <- function(start, end, step, increment = 0.0001) {
  # Calculate the number of bins
  num_bins <- floor((end - start) / step) + 1
  
  # Initialize a list to store the ranges
  ranges <- vector("list", num_bins)
  
  for (i in 0:(num_bins - 1)) {
    if (i == 0) {
      bin_start <- start
      bin_end <- start + step
    } else {
      bin_start <- ranges[[i]][2] + increment
      bin_end <- bin_start + step - increment
    }
    # Round the start and end to three decimal places for clarity
    ranges[[i + 1]] <- c(round(bin_start, 4), round(bin_end, 4))
  }
  
  return(ranges)
}

# Example usage:
distance_ranges <- generate_non_overlapping_ranges(0, 16, 1, increment = 0.0001)
print(distance_ranges)


# Initialize list to store Mantel test results
mantel_gds_results <- list()

# Create a progress bar
total_tasks <- length(names(distance_matrices_all)) * length(distance_ranges) * length(names(distance_matrices_all[[1]]))
pb <- progress_bar$new(total = total_tasks, format = "[:bar] :current/:total (:percent) :elapsedfull")

# Loop over each model and perform Mantel test for each parameter distance matrix
for (model in names(distance_matrices_all)) {
  distance_matrices <- distance_matrices_all[[model]]

  # Loop over each distance range and perform the Mantel test using lapply
  mantel_gds_results[[model]] <- lapply(names(distance_matrices), function(mat_name) {
    lapply(seq_along(distance_ranges), function(j) {
      pb$tick()
      perform_mantel_test(
        distance_ranges[[j]],
        gdm = gdm,  # Assuming gdm is defined elsewhere in your code
        euc_z = distance_matrices[[mat_name]],  # Access the corresponding euc_z matrix
        permutations = 9999
      ) %>%
        mutate(Parameter = mat_name, Model = model)  # Add matrix name and model name as new columns
    })
    }) %>%
  unlist(recursive = FALSE) %>%
  bind_rows()  # Combine all data frames into a single data frame
}

# Combine all Mantel test results into a single data frame
mantel_gds_results_df <- bind_rows(mantel_gds_results)

# Print the resulting data frame with Mantel test results and matrix names
head(mantel_gds_results_df,n = 20)


## plot ANI distance vs corr coefficient and significance

# Custom label function to add model names inside the plot
model_label <- function(variable, value) {
  return(paste(value))
}

unique(mantel_gds_results_df$Range)

# Convert 'Range' to a factor with the desired order
mantel_gds_results_df$Range <- factor(mantel_gds_results_df$Range, levels = unique(mantel_gds_results_df$Range))

# Create the plot using ggplot2
mantel_cor_plot <- ggplot(mantel_gds_results_df %>% filter(Model != "gaussian"),
                          aes(x = Range, y = Correlation_Coefficient, group = Parameter)) +
  geom_line() +
  #geom_line(color = "black", group = 1) +
  #geom_point(size = 3, aes(shape = factor(ifelse(P_Value < 0.05, 16, 1)))) +
  geom_point(size = 6, aes(shape = Parameter, fill = factor(ifelse(P_Value < 0.05, "significant", "non-significant")))) +
  #scale_shape_manual(values = c("significant" = 21, "non-significant" = 21), name = "P value", labels = c("< 0.05", "> 0.05")) +
  scale_fill_manual(values = c("significant" = "black", "non-significant" = "white"), name = "P value", labels =  c("< 0.05", "> 0.05")#, guide = guide_legend(override.aes = list(shape = 21))
                    ) +
  #facet_wrap(~Model, ncol = 1, scales = "free_y") +
  scale_shape_manual(values = c("all_parameters" = 21,  # Circle
                                "ctmax" = 22,  # Square
                                "rmax" = 23,  # Triangle
                                "topt" = 24),  # Diamond
                     limits = c("all_parameters", "ctmax", "rmax", "topt"),
                     labels = c("All", "Max. T", "rmax", "Opt. T")) +
  #scale_shape_manual(values = c("1" = 1, "16" = 16),
  #                   limits = c("16", "1"),
  #                   labels = c("< 0.05", "> 0.05")) +
  #scale_shape_manual(values = c("1" = 1, "19" = 19),
  #                   limits = c("19", "1"),
  #                   labels = c("< 0.05", "> 0.05")) +
  #geom_point(color = "black", size = 3, shape = 19) +
  #scale_color_manual(values = RColorBrewer::brewer.pal(n = 6, name = "Dark2"), labels = c("All",
  #                                                                                        #"Breadth",
  #                                                                                        "ctmax",
  #                                                                                        #"ctmin",
  #                                                                                        "rmax",
  #                                                                                        "Opt. T")
  #                   ) +
  #scale_color_viridis_d(labels = c("All", "AUC", "Breadth", "Max. pH", "Max. gorwth rate", "Min. pH", "Optimal pH")) +
  #geom_text(aes(label = ifelse(P_Value < 0.05, "*", "")), color = "black", vjust = 0, size = 8) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  ylim(0, 1) +
  scale_x_discrete(breaks = seq(0, 17, by = 1), expand = c(0.025, 0.025)) +
  #ggtitle(label = "Correlation between ANI distance and phenotypic distance",
  #        subtitle = "Trait: Thermal performance") +
  labs(x = "",
       y = "Mantel Correlation Coefficient", shape = " Parameter", fill = "P value") +
  #annotate("text", x = 7.5, y = 0.5, label = substitute(paste("* ", italic("p"), " < 0.05")), size = 4) +
  apatheme +
  theme(
   text = element_text(size = 24),
   legend.position = c(0.95, 0.8), # Adjust position to your preference
   legend.background = element_rect(fill = alpha("white", 0.25)),
   legend.box.background = element_rect(color = NA),
   legend.text = element_text(size = 15), # Reduce legend text size
   legend.title = element_text(size = 16), # Reduce legend title size
   legend.key.size = unit(0.25, "lines"),
   legend.spacing.y = unit(1, "pt"),
   #axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), # Rotate x-axis labels for better readability
   axis.text.x = element_blank(),
   axis.ticks.x = element_blank(),
   axis.line.x = element_blank(),
   strip.background = element_blank(), # Remove facet strips
   strip.text = element_blank() # Remove facet strip text
 ) +
  #geom_text(data = mantel_gds_results_df %>%
  #            group_by(Model) %>%
  #            summarize(y = max(Correlation_Coefficient, na.rm = TRUE) * 0.9), 
  #          aes(x = 1, y = y, label = Model),
  #          inherit.aes = FALSE, hjust = 0, vjust = 1, size = 4, fontface = "bold") +
  guides(
    fill = guide_legend(override.aes = list(shape = c(19,1), size = 5)),
    shape = guide_legend(order = 1, override.aes = list(shape = c(19, 15, 18, 17), size = 5)) # Ensure that the shape legend appears first
  )

# Display the plot
print(mantel_cor_plot)


# some quick calculations

avg_cl <- 7.37
sd_cl <- 3.11

ul_cl <- avg_cl + sd_cl
ll_cl <- avg_cl - sd_cl

rng_cl <- ul_cl - ll_cl
rng_cl

avg_sb <- 4.19
sd_sb <- 2.73

ul_sb <- avg_sb + sd_sb
ll_sb <- avg_sb - sd_sb

rng_sb <- ul_sb - ll_sb
rng_sb



hist_plot <- ggplot(ANI_df, aes(x = dist)) +
  geom_histogram(binwidth = 1, fill = "black", color = "black", alpha = 0.7, boundary = 0) +
  scale_x_continuous(breaks = seq(0, 17, by = 1), expand = c(0,0)) +
  geom_vline(xintercept = avg_sb, color = "black", linetype = "dashed", linewidth = 1.5) +
  #geom_vline(xintercept = ll_sb, color = "red", linetype = "dashed") +
  #geom_vline(xintercept = ul_sb, color = "red", linetype = "dashed") +
  geom_vline(xintercept = avg_cl, color = "black", linetype = "dashed", linewidth = 1.5) +
  #geom_vline(xintercept = ul_cl, color = "blue", linetype = "dashed") +
  #geom_vline(xintercept = ll_cl, color = "blue", linetype = "dashed") +
  # Add labels for avg_cl and avg_sb
  annotate("text", x = avg_sb, y = 3200, label = paste("Avg. subclades:", round(avg_sb, 2), sep = " "),
           color = "black", hjust = -0.05, size = 8, angle = 0) +
  annotate("text", x = avg_cl, y = 3200, label = paste("Avg. clades:", round(avg_cl, 2), sep = " "),
           color = "black", hjust = -0.05, size = 8, angle = 0) +
  labs(x = "ANI distance (%)", y = "Count") +
  labs(x = "ANI distance (%)", y = "Count") +
  #ggtitle("Histogram of Genetic Distances") +
  apatheme +
  theme(text = element_text(size = 24))
print(hist_plot)



# Combine the plots
combined_plot <- ggpubr::ggarrange(
  mantel_cor_plot,
  hist_plot,
  ncol = 1,
  nrow = 2,
  heights = c(3, 1),
  align = "v"
)

print(combined_plot)

ggsave("plots/ANIvsTherm_distance.png",
       width = 50,
       height = 16,
       units = "cm",
       dpi = 300)

```

##### Heat map

```{r heatmap, eval=FALSE}
# NOTE: disabled for now — this chunk references `d_params_schol`, which is
# never defined in the document (separate pre-existing bug to resolve later).

head(d_params)

curve_params <- d_params_schol %>%
  select(rmax, topt, ctmax)



library(ggh4x)
# Add a new column for rownames (Strain IDs)
curve_params$strain <- rownames(curve_params)

curve_params <- merge(curve_params, clade_info, by.x = "strain", by.y = "StrainID", all.x = T)

# Arrange the data by the clade variable
curve_params <- curve_params %>%
  arrange(clade, subclade, strain)



# Assuming your data is stored in a dataframe called curve_params
# Convert the data to long format
curve_params_long <- pivot_longer(curve_params,
                                  cols = where(is.numeric),
                                  names_to = "Parameter",
                                  values_to = "Value")

# Scale the data to z-scores by grouping by 'Parameter'
curve_params_long <- curve_params_long %>%
  group_by(Parameter) %>%
  mutate(Value_z = scale(Value)) %>%
  ungroup()


clade_color_vector <- setNames(clade_colors$color, clade_colors$variable)


# Create the heatmap
ggplot(curve_params_long %>% filter(clade != "UND")
       ,
       aes(x = Parameter,
                              y = factor(strain, levels = unique(strain)),
                              fill = Value_z)) +
  geom_tile() +
  scale_fill_gradient2(#colors = c("blue", "white", "red"),
    low = "blue", mid = "white", high = "red",
                       #values = scales::rescale(c(-2, 0, 2)),  # Adjust these values to control the sharpness
                       #breaks = c(-2, -1, 0, 1, 2),  # Adjust breaks for sharper transitions
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +  # Adjust colors as necessary
  labs(x = "Parameter", y = "Strain", fill = "Z-score") +
  ggh4x::facet_grid2(clade ~ ., scales = "free_y", space = "fixed", switch = "y",
              strip = ggh4x::strip_themed(
                background_y = elem_list_rect(
                  fill = clade_color_vector[unique(curve_params_long$clade)]
                )
              )) +
  scale_x_discrete(position = "top", labels = c("Max T.", "rmax", "Opt. T.")
                   ) +
  apatheme +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 16, face = "bold"),  # Align text horizontally and center
    axis.text.y = element_blank(),  # Remove y-axis text
    axis.ticks.y = element_blank(),  # Remove y-axis ticks
    strip.placement = "inside",  # Place facet strips outside the plot area
    strip.text = element_text(face = "bold"),  # Bold text for facet labels
    panel.spacing = unit(0, "lines"),  # Adjust spacing between panels
    strip.background = element_blank(),  # Remove the default strip background
    strip.text.y.left = element_text(angle = 0, vjust = 1, hjust = 1),  # Align facet labels on the left
    legend.title=element_text(size=10),
    legend.title.align = 0.5,
    legend.text=element_text(size=9),
    legend.position = "bottom",
    legend.justification = "center",
    legend.box = "horizontal",
    legend.box.spacing = unit(0, "lines"),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    text = element_text(size = 22),
    panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank()# Ensure the y-axis title is removed
  ) +
   scale_y_discrete(position = "right") 
  # Align facet labels on the left

ggsave("plots/heatmap_curve_params.png",
       width = 6,
       height = 14,
       units = "cm",
       dpi = 600)


# 1. Make named vectors of colours for each level
clade_cols <- c(
  I   = "#7171FF",
  II  = "#EEC900",
  III = "#e31a1c",
  IV  = "#DC89FF"
)

subclade_cols <- c(
  IA = "#7ECFE8", IB = "#009ACD", IC = "#0000E8",
  IIE = "#EEDC82", IIG = "#CD9B1D", IIJ = "#EEAD0E",
  IIIA = "#FF8D8D", IIIB = "#B03060", IIID = "#8B0000", IIIF = "#8B475D",
  IVA = "#E6C4FF", IVB = "#9300FF", IVC = "#4B0082", IVE = "#8968CD"
)

# Create the horizontal heatmap
ggplot(curve_params_long %>% filter(clade != "UND",
                                    !is.na(subclade)),
       aes(x = factor(strain, levels = unique(strain)),
           y = Parameter,
           fill = Value_z)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +  # Adjust colors as necessary
  labs(x = "Strain", y = "Parameter", fill = "Z-score") +
  ggh4x::facet_nested(. ~ clade + subclade, scales = "free_x", space = "fixed"
                      ,
                     strip = ggh4x::strip_nested(
      background_x  = elem_list_rect(fill = c(clade_cols, subclade_cols)
      )
    )
  ) +
  scale_y_discrete(labels = c("Max T.", "rmax", "Opt. T.")) +  # Adjust the y-axis labels
  theme_minimal() +
  theme(
    text = element_text(size = 22),
    axis.text.y = element_text(angle = 0, hjust = 0.5, size = 14, face = "bold"),  # Align text horizontally and center
    axis.text.x = element_blank(),  # Remove x-axis text (Strain)
    axis.ticks.x = element_blank(),  # Remove x-axis ticks (Strain)
    strip.placement = "inside",  # Place facet strips inside the plot area
    strip.text = element_text(face = "bold"),  # Bold text for facet labels
    panel.spacing = unit(0, "lines"),  # Adjust spacing between panels
    strip.background = element_blank(),  # Remove the default strip background
    strip.text.x.top = element_text(angle = 0, vjust = 1, hjust = 0.5),  # Align facet labels on the top
    legend.title = element_text(size = 11, margin = margin(b = 0)), 
    legend.text = element_text(size = 10),
    legend.position = "right",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.box.spacing = unit(0, "lines"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),  # Ensure the y-axis title is removed
    #plot.margin = margin(10, 10, 10, 10)  # Adjust plot margins if necessary
  )

ggsave("plots/heatmap_curve_params.png",
       width = 27,
       height = 4,
       units = "cm",
       dpi = 600)



```
