---
title: "Finalized_pH_Curve_Template_2Col"
author: "BLu12"
date: "2022-09-21"
output: html_document
---

README: This markdown file is meant to do the bulk of the work for the assessment of optimal pH for growth of the strain. 
```{r}
library(growthcurver) #Make sure to note this was just added for the knitted file
library(MASS) 
library(reshape2) 
library(reshape) 

#TODO: Primary information to be entered by the user. Time increment will be in minutes.
time_increment <- 
total_hours <- 

S1ID <- ""
S2ID <- ""
S3ID <- ""
S4ID <- ""
S5ID <- ""


#Read in data, and replace the time column with a new time column
data <- read.csv(file='[FILE NAME]', header = TRUE)

#Initial Stuff ----

data <- data[c(-2)]
tminutes <- (total_hours*60/time_increment) + 1
data$Time <- (1:tminutes)*time_increment

#Getting the data from the plate and employing growthcurver for analysis
summary_plate <- SummarizeGrowthByPlate(plate = data)

#Creating the list for the pHs to match that of the wells
pH_list <- list()
for(i in 0:7){
  pH_list[3*i + 1] <- 3.2 + 0.3*i
  pH_list[3*i + 2] <- 5.6 + 0.3*i
}

#Initializing variables that will be used to create the lists of growth rates for each strain ----
s1count <- 1
s2count <- 1
s3count <- 1
s4count <- 1
s5count <- 1
strain1_g_list <- list()
strain2_g_list <- list()
strain3_g_list <- list()
strain4_g_list <- list()
strain5_g_list <- list()

#Modifies the growth rate lists so that they have the growth rates in them ----
for(i in 1:96){
  
  if(substring(summary_plate$sample[i], 2, nchar(summary_plate$sample[i])) %in% c(1,2)){
    strain1_g_list[s1count] <- summary_plate$r[i]
    s1count <- s1count + 1 
  }
  
  else if(substring(summary_plate$sample[i], 2, nchar(summary_plate$sample[i])) %in% c(3,4)){
    strain2_g_list[s2count] <- summary_plate$r[i]
    s2count <- s2count + 1 
  }
  
  else if(substring(summary_plate$sample[i], 2, nchar(summary_plate$sample[i])) %in% c(7,8)){
    strain3_g_list[s3count] <- summary_plate$r[i]
    s3count <- s3count + 1 
  }
  
  else if(substring(summary_plate$sample[i], 2, nchar(summary_plate$sample[i])) %in% c(9,10)){
    strain4_g_list[s4count] <- summary_plate$r[i]
    s4count <- s4count + 1 
  }
  
  else if(substring(summary_plate$sample[i], 2, nchar(summary_plate$sample[i])) %in% c(11,12)){
    strain5_g_list[s5count] <- summary_plate$r[i]
    s5count <- s5count + 1 
  }
  
}

#Plots the growth rates in accordance with the respective pH. ----
plot(pH_list, strain1_g_list, ylab = "Growth Rate", xlab = "pH", main = S1ID)
plot(pH_list, strain2_g_list, ylab = "Growth Rate", xlab = "pH", main = S2ID)
plot(pH_list, strain3_g_list, ylab = "Growth Rate", xlab = "pH", main = S3ID)
plot(pH_list, strain4_g_list, ylab = "Growth Rate", xlab = "pH", main = S4ID)
plot(pH_list, strain5_g_list, ylab = "Growth Rate", xlab = "pH", main = S5ID)

#Section on Creating the New Tables ----

#Creation of the new table that will just have Strain ID, pH, and Growth Rate
colnames <- seq(1, 16, 1)
rownames <- list("Strain", "pH", "Growth_Rate")

#Creation of the Growth Rate Table for Strain 1 ----
s1g_table <- matrix(c(rep(S1ID, 16), pH_list, strain1_g_list), byrow = TRUE,nrow = 3,ncol = 16 ,dimnames = list(rownames, colnames))

#Continuous transformations of the table to turn them into data frames so that they can easily be manipulated
fin1 <- as.data.frame(t(as.data.frame(s1g_table)))
newfin1 <- transform(fin1, pH = as.numeric(pH), Strain = as.character(Strain), Growth_Rate = as.numeric(Growth_Rate))
melt_data1 <- melt(newfin1, id=c("Strain", "pH"))
table_sorted1 <- melt_data1[order(melt_data1$Strain, melt_data1$pH),]
#table_sorted1 is the final table for Strain 3, to view it, undo the comment out of the following line
#table_sorted1

#Creation of the Growth Rate Table for Strain 2 ----
s2g_table <- matrix(c(rep(S2ID, 16), pH_list, strain2_g_list), byrow = TRUE,nrow = 3,ncol = 16 ,dimnames = list(rownames, colnames))

#Continuous transformations of the table to turn them into data frames so that they can easily be manipulated
fin2 <- as.data.frame(t(as.data.frame(s2g_table)))
newfin2 <- transform(fin2, pH = as.numeric(pH), Strain = as.character(Strain), Growth_Rate = as.numeric(Growth_Rate))
melt_data2 <- melt(newfin2, id=c("Strain", "pH"))
table_sorted2 <- melt_data2[order(melt_data2$Strain, melt_data2$pH),]
#table_sorted2 is the final table for Strain 3, to view it, undo the comment out of the following line
#table_sorted2

#Creation of the Growth Rate Table for Strain 3 ----
s3g_table <- matrix(c(rep(S3ID, 16), pH_list, strain3_g_list), byrow = TRUE,nrow = 3,ncol = 16 ,dimnames = list(rownames, colnames))

#Continuous transformations of the table to turn them into data frames so that they can easily be manipulated
fin3 <- as.data.frame(t(as.data.frame(s3g_table)))
newfin3 <- transform(fin3, pH = as.numeric(pH), Strain = as.character(Strain), Growth_Rate = as.numeric(Growth_Rate))
melt_data3 <- melt(newfin3, id=c("Strain", "pH"))
table_sorted3 <- melt_data3[order(melt_data3$Strain, melt_data3$pH),]
#table_sorted3 is the final table for Strain 3, to view it, undo the comment out of the following line
#table_sorted3

#Creation of the Growth Rate Table for Strain 4 ----
s4g_table <- matrix(c(rep(S4ID, 16), pH_list, strain4_g_list), byrow = TRUE,nrow = 3,ncol = 16 ,dimnames = list(rownames, colnames))

#Continuous transformations of the table to turn them into data frames so that they can easily be manipulated
fin4 <- as.data.frame(t(as.data.frame(s4g_table)))
newfin4 <- transform(fin4, pH = as.numeric(pH), Strain = as.character(Strain), Growth_Rate = as.numeric(Growth_Rate))
melt_data4 <- melt(newfin4, id=c("Strain", "pH"))
table_sorted4 <- melt_data4[order(melt_data4$Strain, melt_data4$pH),]
#table_sorted4 is the final table for Strain 4, to view it, undo the comment out of the following line
#table_sorted4

#Creation of the Growth Rate Table for Strain 5 ----
s5g_table <- matrix(c(rep(S5ID, 16), pH_list, strain5_g_list), byrow = TRUE,nrow = 3,ncol = 16 ,dimnames = list(rownames, colnames))

#Continuous transformations of the table to turn them into data frames so that they can easily be manipulated
fin5 <- as.data.frame(t(as.data.frame(s5g_table)))
newfin5 <- transform(fin5, pH = as.numeric(pH), Strain = as.character(Strain), Growth_Rate = as.numeric(Growth_Rate))
melt_data5 <- melt(newfin5, id=c("Strain", "pH"))
table_sorted5 <- melt_data5[order(melt_data5$Strain, melt_data5$pH),]
#table_sorted5 is the final table for Strain 5 to view it, undo the comment out of the following line
#table_sorted5

#CSV File Export **User Input Needed** ----

#Bringing all the dataframes together and taking out the variable column and putting it as the title of the value column
total <- rbind(table_sorted1, table_sorted2, table_sorted3, table_sorted4, table_sorted5)
names(total)[names(total)=="value"] <- "Growth_Rate"
total <- total[c(1,2,4)]
#To look at the dataframe, undo the comment out of the following line
#total
```

The function of this block of code is to create a csv file with just the data from the plate that was done. It also adds that data into a larger csv file that intends to have all the data aggregated into one big file.
```{r}
#Outputting into its own CSV File
write.csv(total,"[SINGLE CSV TARGET]", row.names = FALSE)

#Outputting into the main CSV File
data2 <- read.csv(file='[MAIN CSV TARGET]', header = TRUE)
total2 <- rbind(data2, total)
write.csv(total2,"[MAIN CSV TARGET]", row.names = FALSE)
```

The function of this block of code is to take in the data from the big csv file and graph the pH vs growth rate curves. This one will aim to have the average growth rate on the y-axis, and also have the error bars for them. 
```{r}

data3 <- read.csv(file='[MAIN CSV FILE]', header = TRUE)

#Sorts the data with Strain and pH as the main things that are accounted for
melt_data <- melt(data3, id = c("Strain", "pH"))

#Creates an Strain-pH column, which is meant to help with uniquely identifying each set of strain-pH combinations
for(i in 1:nrow(melt_data)){
  melt_data$SPH[i] <- paste(melt_data$Strain[i], "pH", toString(melt_data$pH[i]), sep = "_")
}

#Creates a new table that has the average growth rate as the value
average_table <- cast(melt_data, SPH~variable, mean)

#New list for the pH for the average_table that will make it possible to be graphed
pH_list2 <- seq(3.2, 7.7, 0.3)

#This counts how many unique strains are listed in the average_table variable
numstrain <- nrow(average_table)/16

#This plots the graph of each average growth rate vs pH for the strain
# for(i in 0:(numstrain-1)){
#   plot(pH_list2, average_table[(1 + 24*i):(24 + 24*i),2], ylab = "Growth Rate Mean", xlab = "pH", main = melt_data$Strain[1 + 24*i])
# }

#This adds a pH column to the average_table for reasons I do not yet know
average_table$pH <- as.numeric(pH_list2) 

#This does statistics for me!
ell <- melt_data[order(melt_data$Strain, melt_data$pH),]
se <- function(x) sd(x)/sqrt(length(x))
ksenia <- cast(melt_data, SPH~variable, se)
sen <-unique(unlist(ell[,1], use.names = FALSE))
#This plots again, with error bars, each error bar's length is that of a single standard deviation from the mean
for(i in 0:(numstrain-1)){
  plot(pH_list2, average_table[(1+16*i):(16+16*i),2], ylab = "Approximated Growth Rate Mean", xlab = "pH", main = sen[i+1])  
  arrows(x0=as.numeric(pH_list2), y0=average_table[(1+16*i):(16+16*i),2] - ksenia[(1+16*i):(16+16*i),2], x1=as.numeric(pH_list2), y1=average_table[(1+16*i):(16+16*i),2]+ ksenia[(1+16*i):(16+16*i),2], code=3, col="blue", lwd=2, angle=90, length = 0.05)
}


```