---
title: "GC_initial_analysis_384_growthrates_pkg"
author: "Alberto Barron"
date: "`r Sys.Date()`"
output: html_document
---

### Description

This script is meant to analyse the growth data for a single pH assay on a 384 well plate. Each plate has 24 pH levels in which we grow fifteen (15) different isolates plus the blanks. In order to do this this script:

1 - Takes the output raw data from the plate reader and do background correction, plot the data after background correction, plot the blanks to be sure that everything looks okay.

2 - It uses the package 'growthcurver' to create a model for the growth curve of each strain on each pH level and plots the predicted values from the model along with the observed values from the plate reader.

3 - Get the growth curve parameters for every combination of pH and strain and plots the estimated growth rate (r) against the pH level to obtain a pH response curve and exports the growth curve parameters to text file to consolidate all data for further analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, cache = F)
options(scipen=10000, digits = 12)
set.seed(seed = 711)
```

###### General settings

In this chunk of code we load the required packages, set the general aesthetics for the plots and set the working directory.

```{r general settings, message=FALSE, warning=FALSE, results='hold'}

#library(growthcurver)
library(lubridate)
library(ggplot2)
library(dplyr)
library(reshape2)
library(purrr)

apatheme=theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family='Arial'))

taxa_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD", 
  "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861", "red",
  "blue", "#483D8B", "gray", "darkcyan", "darkgoldenrod4", "green", "#8B0000",
  "darkgreen", "yellow", "lightblue", "forestgreen", "darkorange4", "purple", "salmon", "steelblue", "pink")

#set working directory
# working directory handled by here::here()
```

#### Loading the data

First, we have to load the raw data that the plate reader gives us. To do this let's set the file name, date and isolate IDs.

**ATTENTION: This is the only part of the script that should be modified to change the input file name, strain names and date**

```{r set up file name strains and date}
# Set the name of the output file from the plate reader
raw_plate_reads_file<-'pHassay_033123_merged.csv'

# Set the names of the strains in this plate
S1<-"25G2A02" # Strain in Row Row A
S2<-"25G2A02" # Strain in Row B
S3<-"25G2A02" # Strain in Row C
S4<-"35G2B04" # Strain in Row D
S5<-"35G2B04" # Strain in Row E
S6<-"35G2B04" # Strain in Row F 
S7<-"35G3B04" # Strain in Row G
S8<-"35G3B04" # Strain in Row H
S9<-"35G3B04" # Strain in Row I
S10<-"25D05" # Strain in Row J
S11<-"25D05" # Strain in Row K
S12<-"25D05" # Strain in Row L
S13<-"28D03" # Strain in Row M
S14<-"28D03" # Strain in Row N
S15<-"28D03" # Strain in Row O

# Set Value with the date of the pH assay on format yyyy-mm-dd
date<-"2023-03-31"
```

Now the data is ready to be loaded into R

```{r load data}
# Read in the Data
data <- read.csv(file=raw_plate_reads_file, header = TRUE, stringsAsFactors = F)
# Remove the temperature column
data <- data[,-2]

# Change time from HH:MM:SS to hours and fractions in decimals
#data$Time<-period_to_seconds(hms(data$Time))/3600
```

#### Make a map of the plate

In this step we create a plate map that matches the strain ID and pH to their corresponding well.

```{r}
## Create a map of the plate layout that matches strain name, pH and well
# A list of the pH levels in each row
phlist<-as.factor(c(rep(seq(3.2, 10.2 , 0.3), 16))) # pH levels for Each row

SVEC<-c(S1,S2,S3,S4,S5,S6,S7,S8,S9,S10,S11,S12,S13,S14,S15)

# A list of the strains in each row
strainlist<-c()

for (i in 1:length(SVEC)) {
  tmpvec<-rep(SVEC[i],24)
  strainlist<-c(strainlist, tmpvec)
  rm(tmpvec)
}
# Add the negative wells
strainlist<-c(strainlist, rep("NEG", 24))

# in case of replicates inside the same plate run the following lines (uncomented)
replicate<-as.factor(c(rep(c(
  rep(1, 24),
  rep(2, 24),
  rep(3, 24)), 5),
  rep(NA, 24)))


# Put together both vectors (lists) into a data frame
#platemap<-cbind.data.frame(strainlist, phlist)
platemap<-cbind.data.frame(strainlist, phlist, replicate)
rm(phlist, strainlist)

# Get the names of the names columns from the data frame (A1, A2,..., H12) to a single column named "well" in our plate map data frame
platemap$well<-colnames(data[,-1])
```

### Plotting the raw data

In this step we will take a first look to the data and do the background correction and plot the data to see if everything looks as expected.

```{r preliminary results, fig.width=12, fig.height = 8}
# Reshape the table with many columns, one for each well to one whith just time, well name, and OD
mdata<-reshape2::melt(data, id = "Time", variable.name = "well")

# Now let's map the strain name and pH level to the long data frame 
mdata<-merge(mdata, platemap, by = "well", all.y = T)
colnames(mdata)[4]<-"strain"
colnames(mdata)[5]<-"pH"


# let's subset the blank wells so we can substract them from the OD measurments
mdata_neg<-mdata %>% dplyr::filter(strain == "NEG")


#temp_dataset <- mdata %>%
#      filter(strain == unique(SVEC)[1] & replicate == 1) %>%
#      merge(., mdata_neg[,-c(1,4,6)], by = c("Time", "pH")) %>%
#        mutate(OD = value.x - value.y) %>%
#        select(-c(value.y, value.x)) %>%
#        arrange(well, Time)



## Method 1: match blank readings by time and pH and then subtract the blanks from the OD values
for (i in 1:length(unique(SVEC))) {
  # inner loop to filter by replicate and substract the mean negative OD values
    for (k in c(1:3)) {
  if (!exists("mdata_temp")){
    mdata_temp <- data.frame()
  }
    
  if (exists("mdata_temp")) {
    
    temp_dataset <- mdata %>%
      filter(strain == unique(SVEC)[i] & replicate == k) %>%
      merge(., mdata_neg[,-c(1,4,6)], by = c("Time", "pH")) %>%
      mutate(OD = value.x - value.y) %>%
      select(-c(value.y, value.x)) %>%
      arrange(well, Time)

    mdata_temp<-unique(rbind(mdata_temp, temp_dataset))
    rm(temp_dataset)
    }
  }
}

### Method 2: average blank values and substract from the OD values
#negs<-as.character(unique(mdata_neg$well))
#dnegm<-data.frame(Time = data$Time,
#                  negemeanOD = apply(data[, negs], 1, mean))

#for (i in 1:length(SVEC)) {
#  # inner loop to filter by pH and substract the mean negative OD values
#  for (j in 1:length(unique(platemap$phlist))) {
#    for (k in c(1:3)) {
#      
#   
#    
#  # if the merged dataset doesn't exist, create it
#  if (!exists("mdata_temp")){
#    mdata_temp <- data.frame()
#  }
#    
#  if (exists("mdata_temp")) {
#    
#    temp_dataset <- mdata %>%
#      filter(strain == SVEC[i] & pH == unique(platemap$phlist)[j] & replicate == k) %>%
#        mutate(OD = value - dnegm$negemeanOD) %>%
#        select(-value) %>%
#        arrange(well, Time)
#
#    mdata_temp<-unique(rbind(mdata_temp, temp_dataset))
#    rm(temp_dataset)
#  }
#    }
#  }
#}


mdata <- mdata_temp
rm(mdata_temp)


## now we have our data ready to plot the preliminary growth curves

phlabs<-data.frame(label= sort(unique(mdata$pH)),  # Create data for text
                        pH = sort(unique(mdata$pH)),
                        x = rep(10, 24),
                        y = rep(0.35, 24))
for (i in 1:(length(unique(SVEC)))){
plot<-mdata %>%
  filter(strain == unique(SVEC)[i]) %>% 
  ggplot(aes(x = Time, y = OD)) +
    geom_text(phlabs, mapping = aes(x = x, y = y, label = paste("pH",label))) +
    geom_point(aes(colour = replicate), alpha=0.3, size = .7) +
    ylim(0, 0.4) +
    xlab("Time (hrs)") +
    facet_wrap(~pH, ncol = 6) +
    labs(title = paste("pH assay for isolate", unique(SVEC)[i]),
         subtitle = date) +
    apatheme +
    theme(strip.background = element_blank(),
          strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 12),
          plot.subtitle = element_text(color = "black", size = 10))
  
  print(plot)
  }

## let's plot the negatives just to be sure that there's no growth in those

ggplot(mdata_neg, aes(x = Time, y = value)) +
    geom_point(alpha=0.7) +
    ylim(0, 0.4) +
    xlab("Time (hrs)") +
    geom_text(phlabs, mapping = aes(x = x, y = y, label =  paste("pH",label))) +
    facet_wrap(~pH, ncol = 6) +
    labs(title = "pH assay NEG",
            subtitle = date) +
    apatheme +
    theme(strip.background = element_blank(),
          strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 12),
          plot.subtitle = element_text(color = "black", size = 10))


```

Now we use 'gorwthrates'

```{r}
library(growthrates)

#transform time to minutes

mdata$Time<-mdata$Time*60*60

mdata$OD[mdata$OD < 0]<-0

  
  
many_spline_fits <- all_easylinear(OD ~ Time | strain + pH + replicate,
                                data = mdata, spar = 0.5)


splitted.data <- multisplit(mdata, c("strain", "pH", "replicate"))
dat <- splitted.data[[2]]

fit <- fit_easylinear(dat$Time, dat$OD)

summary(fit)

par(mfrow = c(1, 2))
plot(fit, log = "y")
plot(fit)



ggplot(mdata %>% mutate(x = rep('x', nrow(mdata))),
       aes(x = pH, y = OD, fill = replicate)) +
  geom_boxplot() +
  facet_grid(strain~replicate) +
  geom_point(alpha=0.2, size = 0.2,
    position = position_jitterdodge())

```