---
title: "Growth Curves analysis for a 96 Well Plate pH assay"
author: "Alberto Barron"
date: "`r date`"
output: html_document
params:
  date: !r date
  strains: !r SVEC
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, cache = F)
knitr::opts_knit$set(root.dir = here::here())
options(scipen=10000, digits = 12)
set.seed(seed = 711)
```

## Strains

`r paste0("- ",SVEC, collapse = "\n")`

### Description

This script is meant to analyse the growth data for a single pH assay on a 96 well plate. Each plate has 24 pH levels in which we grow three (3) different isolates plus the blanks. In order to do this this script:

1 - Takes the output raw data from the plate reader and do background correction, plot the data after background correction, plot the blanks to be sure that everything looks okay.

2 - It uses the package 'growthcurver' to create a model for the growth curve of each strain on each pH level and plots the predicted values from the model along with the observed values from the plate reader.

3 - Get the growth curve parameters for every combination of pH and strain and plots the estimated growth rate (r) against the pH level to obtain a pH response curve and exports the growth curve parameters to text file to consolidate all data for further analysis.

###### General settings
In this chunk of code we load the required packages, set the general aesthetics for the plots and set the working directory.
```{r general settings, message=FALSE, warning=FALSE, results='hold'}

library(growthcurver)
library(lubridate)
library(ggplot2)
library(dplyr)
library(reshape2)
library(purrr)

apatheme=theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family='Futura Medium'))

#set working directory
## working directory handled by here::here()
getwd()
```

#### Loading the data

First, we have to load the raw data that the plate reader gives us. To do this let's set the file name, date and isolate IDs.

**ATTENTION: This is the only part of the script that should be modified to change the input file name, strain names and date**

```{r set up file name strains and date}
# Set the name of the output file from the plate reader
#raw_plate_reads_file<-'pHassay_020923.csv'
#
## Set the names of the strains in this plate
#S1<-"35G2A07" # Strain in columns 1-3
#S2<-"25D05" # Strain in columns 4-6
#S3<-"35G3B04" # Strain in columns 10-12
#
## Set Value with the date of the pH assay on format yyyy-mm-dd
date#<-"2023-02-09"
```

Now the data is ready to be loaded into R

```{r load data}
# Read in the Data
#data <- read.csv(file=raw_plate_reads_file, header = TRUE, stringsAsFactors = F)
# Remove the temperature column
data <- data[,-2]

# Change time from HH:MM:SS to hours and fractions in decimals
data$Time<-period_to_seconds(hms(data$Time))/3600
```

#### Make a map of the plate

In this step we create a plate map that matches the strain ID and pH to their corresponding well.

```{r}
## Create a map of the plate layout that matches strain name, pH and well
# A list of the pH levels in each row
phlist<-c(rep(seq(3.2, 8.0 ,2.4), 4), # pH leves for row A 
          rep(seq(3.5, 8.3, 2.4), 4), # pH leves for row B
          rep(seq(3.8, 8.6, 2.4), 4), # pH leves for row C
          rep(seq(4.1, 8.9, 2.4), 4), # pH leves for row D
          rep(seq(4.4, 9.2, 2.4), 4), # pH leves for row E
          rep(seq(4.7, 9.5, 2.4), 4), # pH leves for row F
          rep(seq(5.0, 9.8, 2.4), 4), # pH leves for row G
          rep(seq(5.3, 10.1, 2.4), 4))# pH leves for row H

# A list of the strains in each row
strainlist<-c(rep(c(rep(S1, 3),    # Wells in columns 1-3
                    rep(S2, 3),    # Wells in columns 4-6
                    rep("NEG", 3), # Wells in columns 7-9
                    rep(S3, 3)),   # Wells in columns 1-12
                  8))              # Repeat 8 times for each row

# Put together both vectors (lists) into a data frame
platemap<-cbind.data.frame(strainlist, phlist)

# Get the names of the names columns from the data frame (A1, A2,..., H12) to a single column named "well" in our plate map data frame
platemap$well<-colnames(data[,-1])
```

### Plotting the raw data

In this step we will take a first look to the data and do the background correction and plot the data to see if everything looks as expected.

```{r preliminary results, fig.width=12, fig.height = 8}
# Reshape the table with many columns, one for each well to one whith just time, well name, and OD
mdata<-reshape2::melt(data, id = "Time", variable.name = "well")

# Now let's map the strain name and pH level to the long data frame 
mdata<-merge(mdata, platemap, by = "well", all.y = T)
colnames(mdata)[4]<-"strain"
colnames(mdata)[5]<-"pH"

# let's subset the blank wells so we can substract them from the OD measurments
mdata_neg<-mdata %>% dplyr::filter(strain == "NEG")

## Now let's substract the blank reading for each reading at each time point and pH level

#SVEC<-c(S1,S2,S3)

for (i in 1:length(SVEC)) {

  if (!exists("mdata_temp")){
    mdata_temp <- data.frame()
  }
    
  if (exists("mdata_temp")) {
    
    temp_dataset <- mdata %>%
      filter(strain == (SVEC)[i]) %>%
      merge(., mdata_neg[,-c(1,4,6)], by = c("Time", "pH")) %>%
      mutate(OD = value.x - value.y) %>%
      select(-c(value.y, value.x)) %>%
      arrange(well, Time)

    mdata_temp<-unique(rbind(mdata_temp, temp_dataset))
    rm(temp_dataset)
    }
  }

## set al negative values to zero "0"
mdata_temp$OD[mdata_temp$OD < 0] <- 0
mdata <- mdata_temp
rm(mdata_temp)

## now we have our data ready to plot the preliminary growth curves

phlabs<-data.frame(label= sort(unique(mdata$pH)),  # Create data for text
                        pH = sort(unique(mdata$pH)),
                        x = rep(10, 24),
                        y = rep(0.7, 24))
for (S in c(S1, S2, S3)){
plot<-mdata %>%
  filter(strain == S) %>% 
  ggplot(aes(x = Time, y = OD)) +
    geom_point(alpha=0.5) +
    ylim(0, 1) +
    xlab("Time (hrs)") +
    geom_text(phlabs, mapping = aes(x = x, y = y, label = paste("pH",label)), family = "mono") +
    facet_wrap(~pH, ncol = 6) +
    labs(title = paste("pH assay for isolate",S),
         subtitle = date) +
    apatheme +
    theme(strip.background = element_blank(),
          strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 12),
          plot.subtitle = element_text(color = "black", size = 10))
  
  print(plot)
  }

## let's plot the negatives just to be sure that there's no growth in those

ggplot(mdata_neg, aes(x = Time, y = value)) +
    geom_point(alpha=0.5) +
    ylim(0, 0.1) +
    xlab("Time (hrs)") +
    geom_text(phlabs, mapping = aes(x = x, y = y-0.6, label =  paste("pH",label)), family = "mono") +
    facet_wrap(~pH, ncol = 6) +
    labs(title = "pH assay NEG",
            subtitle = date) +
    apatheme +
    theme(strip.background = element_blank(),
          strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 12),
          plot.subtitle = element_text(color = "black", size = 10))


```

### Analyze the growth curves

In this step we will use the 'growthcurver' package to calculate growth curve parameters and plot the adjusted model to the plots.

```{r ploting predicted values, fig.width=12, fig.height = 8}
## We have removed the blank values from our data frame, next we can transform it again to it's original form to run 'growthcurver' on that data set
#mdata<-mdata[,-c(3:5)]
dmdata<-reshape2::dcast(mdata[-c(2,4)], Time ~ well, value.var = "OD")

## Create a function to run growthcurver::SummarizeGrowth on all columns of the data
summG <- function(x) {SummarizeGrowth(dmdata$Time, x, t_trim = 0, bg_correct = "none")}
## Apply that function over the columns of the data
models.all<-lapply(dmdata[2:ncol(dmdata)], summG)

n<-length(names(models.all))
# Let's check which curves had a poor fit or could not be fitted
plate_summary<-data.frame(sample = character(n),
                          k = numeric(n),
                          n0  = numeric(n),
                          r = numeric(n),
                          t_mid = numeric(n),
                          t_gen = numeric(n),
                          auc_l = numeric(n),
                          auc_e = numeric(n),
                          sigma = numeric(n),
                          note = character(n),
                          stringsAsFactors = FALSE)

n<-1
for (col_name in names(models.all)) {
  
      plate_summary$sample[n] <- col_name
      plate_summary$k[n] <-models.all[[col_name]]$vals$k
      plate_summary$n0[n]  <-models.all[[col_name]]$vals$n0
      plate_summary$r[n] <-models.all[[col_name]]$vals$r
      plate_summary$t_mid[n] <-models.all[[col_name]]$vals$t_mid
      plate_summary$t_gen[n] <-models.all[[col_name]]$vals$t_gen
      plate_summary$auc_l[n] <-models.all[[col_name]]$vals$auc_l
      plate_summary$auc_e[n] <-models.all[[col_name]]$vals$auc_e
      plate_summary$sigma[n] <-models.all[[col_name]]$vals$sigma
      plate_summary$note[n]  <-models.all[[col_name]]$vals$note

      n<-n+1
}

plate_summary<-merge(plate_summary, platemap, by.x = "sample", by.y = "well", all.y = T)
plate_summary<-plate_summary %>% filter(strainlist != "NEG")
plate_summary %>% filter(note != "") %>% select(sample, note, strainlist, phlist)

max.OD.df <- as.data.frame(apply(dmdata[,-1], 2, max))
max.OD.df$sample <- row.names(max.OD.df)
colnames(max.OD.df)[1] <- "max.OD"
plate_summary <- merge(plate_summary, max.OD.df, by = "sample")

plate_summary <- plate_summary %>%
  #mutate(note = case_when(k < 0.1 ~ "k < 0.1",
  #                        k > 1 ~ "k > 1",
  #                        TRUE ~ note)) %>%
  mutate(r = case_when(k < 0.086 ~ 0,
                          k > 1 ~ r,
                          TRUE ~ r)) %>%
  mutate(k = case_when(k < 0.086 ~ 0,
                          k > 1 ~ max.OD,
                          TRUE ~ k))

#Let's get a list of the well ID of those that could not be fitted
not_fit<-c(plate_summary %>% filter(note == "cannot fit data") %>% select(sample))$sample


df.predicted.plate <- data.frame(Time = dmdata$Time)
for (i in names(dmdata[2:ncol(dmdata)])[!names(dmdata[2:ncol(dmdata)]) %in% not_fit]){
  df.predicted.plate[[i]] <- predict(models.all[[i]]$model)
  }

for (i in not_fit){
  df.predicted.plate[[i]] <- rep(0.085, nrow(df.predicted.plate))
}

#df.predicted.plate <- data.frame(Time = dmdata$Time)
#for (i in names(dmdata[2:ncol(dmdata)])){
#  df.predicted.plate[[i]] <- predict(models.all[[i]]$model)
#  }

melt1 <- melt(dmdata, id.vars = "Time", variable.name = "well", value.name = "OD")
melt2 <- melt(df.predicted.plate, id.vars = "Time", variable.name = "well", value.name = "pred.OD")
#df.final <- cbind(melt1, pred.OD=melt2[,3])
df.final <- merge(melt1, melt2, by = c("Time", "well"))

#df.final<-merge(df.final, platemap, by = "well", all.y = T)
df.final<- df.final %>%
  merge(., platemap, by = "well", all.y = T) %>%
  merge(., plate_summary[,c(1,2,4,9,10)], by.x = "well", by.y = "sample")
colnames(df.final)[5]<-"strain"
colnames(df.final)[6]<-"pH"

df.final<-df.final %>% filter(strain != "NEG")

for (S in c(S1, S2, S3)){

tmp<- df.final %>%
  filter(strain == S) %>%
  select(well, pH, note, r, k, sigma) %>%
  distinct(pH, .keep_all = TRUE) %>%
  merge(., phlabs, by = "pH")
  
  
  
plot<-df.final %>%
  filter(strain == S) %>% 
  ggplot(aes(x = Time, y = OD)) +
    geom_point(alpha=0.3) +
    geom_line(aes(y=pred.OD), color="red") +
    ylim(0, 0.8) +
    xlab("Time (hrs)") +
    geom_text(tmp, mapping = aes(x = x, y = y, label =  paste("pH",label)), family = "mono") +
    geom_text(tmp, mapping = aes(x = x+10, y = y-0.05, label = note), family = "mono") +
    geom_text(tmp, mapping = aes(x = x, y = y+0.05, label = well), family = "mono") +
    geom_text(tmp, mapping = aes(x = x, y = y-0.1, label = paste("r:",round(r, digits = 3))), family = "mono") +
    geom_text(tmp, mapping = aes(x = x, y = y-0.15, label = paste("K:",round(k, digits = 3))), family = "mono") +
    geom_text(tmp, mapping = aes(x = x+5, y = y-0.20, label = paste("sigma:",round(sigma, digits = 2))), family = "mono") +
    facet_wrap(~pH, ncol = 6) +
    labs(title = paste("pH assay for isolate",S),
         subtitle = date) +
    apatheme +
    theme(strip.background = element_blank(),
          strip.text.x = element_blank(),
          plot.title = element_text(color = "black", size = 12),
          plot.subtitle = element_text(color = "black", size = 10))
  
  print(plot)
  rm(tmp)
  }

#df.final %>% filter(well == "D12") %>%
#  ggplot(aes(x = Time, y = OD)) + geom_point() +
#  geom_line(aes(y=pred.OD), color="red")

```

### pH curve

Now let's take a look at the pH curves, that is  the estimated growth rate (r) as a function of the pH level. For this we are going to plot the growth rate against the pH levels.

```{r pH curve, fig.width=12, fig.height = 8}
## First let's get the parametters that growthcurver::SummarizeGrowthByPlate
growth.values.plate <- plate_summary %>% arrange(sample) #%>% filter(note == "")

## Now let's add the strain and pH level information, as well as the date to keep track of when the assay was performed
#growth.values.plate<-merge(x = growth.values.plate, y = platemap, by.x = "sample", by.y = "well", all.y = T)
colnames(growth.values.plate)[11]<-"strain"
colnames(growth.values.plate)[12]<-"pH"
#growth.values.plate<-growth.values.plate %>% filter(strain != "NEG")
growth.values.plate$date<-rep(date, length(row.names(growth.values.plate)))

pH_curve<-ggplot(growth.values.plate #%>% filter(note == "")
                 ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = r)) +
  #geom_line() +
  scale_color_discrete(
    #values = taxa_colors,
    name = "Isolate") +
  geom_smooth(aes(fill = strain)) +
  scale_fill_discrete(#values = taxa_colors,
                      name = "Isolate") +
  geom_point(size = 3) +
  facet_wrap(~strain, ncol = 2) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.3)) +
  ylim(0, 1.5) +
  labs(title = "pH curve",
       subtitle = date) +
  ylab("Growth rate (r)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        legend.position = "none")

print(pH_curve)


pH_curve_K<-ggplot(growth.values.plate #%>% filter(note == "")
                   ,
                 aes(x = pH,
                    color = strain,
                    group = strain,
                    y = k)) +
  #geom_line() +
  scale_color_discrete(
    #values = taxa_colors,
    name = "Isolate") +
  geom_smooth(aes(fill = strain)) +
  scale_fill_discrete(#values = taxa_colors,
                      name = "Isolate") +
  geom_point(size = 3) +
  facet_wrap(~strain, ncol = 2) +
  scale_x_continuous(breaks = seq(3.2, 10.1, 0.3)) +
  ylim(0, 0.8) +
  labs(title = "pH curve K",
       subtitle = date) +
  ylab("Carrying capacity (K)") +
  apatheme +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
        legend.position = "none")

print(pH_curve_K)
```

##### Exporting the processed data

Finally, let's. export the processed data, the table with the estimated growth curve parameters so we can consolidate it all into a large data file that we will use to store all the replicates for each strain that we use for the pH assays.

```{r exporting processed data}
write.table(growth.values.plate, file = paste(here::here("data/pH/GC_values_pHassay"),date,"txt", sep = "."), sep = "\t", row.names = FALSE, quote = FALSE)
```