---
title: "Growth curve analysis for a 384-well plate pH assay"
author: "Alberto Barron"
date: "`r Sys.Date()`"
output: html_document
params:
  raw_file: "data-raw/pH/pHassay_example.csv"
  date: "2025-01-01"
  strains: !r NULL
  replicates: "auto"
  blank_method: "match"
  drop_neg_ph: !r NULL
  offset_wells: !r NULL
  problem_wells: !r NULL
  high_k_action: "zero"
  blank_baseline: 0.085
  output_file: !r NULL
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(here)
source(here("R/pH/provenance/gc_helpers.R"))

knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here())
options(scipen = 10000, digits = 12)
set.seed(seed = 711)
```

### Description

This template analyzes raw plate-reader data for a single pH assay on a 384-well
plate. Each plate holds 24 pH levels (columns) in which up to 15 isolates are
grown across the strain rows, plus a row of blanks (`NEG`). It:

1. Loads the raw reads, subtracts the blanks (background correction), and plots
   the corrected curves and the blanks for QC.
2. Fits a logistic growth model per well with **growthcurver** and overlays the
   predicted vs. observed curves.
3. Extracts growth parameters for every strain × pH combination, plots the pH
   response curves (growth rate `r` and carrying capacity `k`), and exports the
   parameter table for downstream consolidation.

All heavy lifting lives in `R/pH/provenance/gc_helpers.R`; the chunks below just
wire the helpers together.

### Configuration

The template runs either from `params` (via `run_gc_template.R`) or standalone by
editing the fallback values in the chunk below. **This is the only chunk you need
to edit for a manual run.**

```{r config}
# Strain in each row A..O (repeat a strain to encode in-plate replicates).
cfg_strains <- params$strains %||% c(
  "Wood-1",  "Wood-2",  "Wood-11", "Wood-31", "Wood-46",
  "Wood-47", "Wood-52", "Pine-20", "Desert-30", "Mmlr14-006",
  "Mcba15-009", "Mcba15-003", "23G3A05", "23G1B05", "Wood-1"
)

cfg_raw_file      <- params$raw_file
cfg_date          <- params$date
cfg_replicates    <- params$replicates      # "auto", "none", or integer vector
cfg_blank_method  <- params$blank_method    # "match" or "mean"
cfg_drop_neg_ph   <- params$drop_neg_ph     # e.g. c(6.2, 6.5) or NULL
cfg_offset_wells  <- params$offset_wells    # character vector or NULL
cfg_problem_wells <- params$problem_wells   # character vector or NULL
cfg_high_k_action <- params$high_k_action   # "zero" or "rescue"
cfg_baseline      <- params$blank_baseline

# Default output path mirrors the committed GC_values naming convention.
cfg_output_file <- params$output_file %||%
  here(paste("data/pH/GC_values_pHassay", cfg_date, "txt", sep = "."))
```

## Strains

`r paste0("- ", cfg_strains, collapse = "\n")`

### Load the data and build the plate map

```{r load-and-map}
data <- gc_load_plate_reads(here(cfg_raw_file))

platemap <- gc_build_platemap(
  well_names = colnames(data)[-1],
  strains    = cfg_strains,
  replicates = cfg_replicates
)

mdata_long <- gc_to_long(data, platemap)
```

### Background correction and raw curves

```{r raw-plots, fig.width=12, fig.height=8}
mdata_corr <- mdata_long |>
  gc_subtract_blanks(
    method      = cfg_blank_method,
    drop_neg_ph = cfg_drop_neg_ph
  ) |>
  gc_apply_offset(offset_wells = cfg_offset_wells)

phlabs <- gc_phlabs(mdata_corr)

for (s in unique(cfg_strains)) {
  print(gc_plot_raw(mdata_corr, s, cfg_date, phlabs))
}

# QC: the blank wells should show no growth.
mdata_neg <- mdata_long |> dplyr::filter(strain == "NEG")
gc_plot_neg(mdata_neg, cfg_date, phlabs)
```

### Fit growth models

```{r fit, fig.width=12, fig.height=8}
fit <- gc_fit_models(mdata_corr)

# Report any wells with a fit note (poor fit / could not fit).
fit$summary |>
  merge(platemap, by.x = "sample", by.y = "well") |>
  dplyr::filter(note != "", strain != "NEG") |>
  dplyr::select(sample, note, strain, pH)

predicted <- gc_predicted(fit)

params_corr <- gc_correct_params(
  summary        = fit$summary,
  platemap       = platemap,
  wide           = fit$wide,
  blank_baseline = cfg_baseline,
  high_k_action  = cfg_high_k_action,
  problem_wells  = cfg_problem_wells
)

df_final <- gc_join_final(mdata_corr, predicted, platemap, params_corr)

for (sr in unique(df_final$strain.rep)) {
  print(gc_plot_fitted(df_final, sr, cfg_date, phlabs))
}
```

### pH response curves

```{r ph-curves, fig.width=12, fig.height=8}
growth_values <- params_corr |>
  dplyr::arrange(sample)
growth_values$date <- cfg_date

print(gc_plot_ph_curve(growth_values, y = "r", date = cfg_date))
print(gc_plot_ph_curve(growth_values, y = "k", date = cfg_date))
```

### Export the processed data

```{r export}
gc_export_values(growth_values, cfg_output_file)
```
